---
title: "1.4 Rational Consumer Choice"
output: 
  learnr::tutorial: 
    progressive: true
runtime: shiny_prerendered
author: 
  name: "Created by [Jim Bang](http://www.github.com/bangecon)"
  email: BangJamesT@sau.edu
  affiliation: "[St. Ambrose University](http://www.sau.edu)"
date: "`r format(Sys.Date(), '%d %b %Y')`"
---

<style type="text/css">
#doc-metadata {
  margin-bottom: 10px;
  font-size: 0.8em;
}
</style>

```{r setup, include = FALSE}
library(learnr)
library(gradethis)
library(Ryacas)
library(Deriv)
library(Rsolnp)
gradethis_setup()
tutorial_options(exercise.eval = TRUE, exercise.reveal_solution = FALSE)
knitr::opts_chunk$set(echo = FALSE, fig.align = "center", fig.height = 6, fig.width = 6)
knitr::include_graphics
```

## Constant Elasticity of Substitution

Suppose that we can describe a consumer's preferences between two goods using a constant elasticity of substitution utility function with $r = 1/2$: 

$$U(x) = (x_1^{1/2} + x_2^{1/2})^2, $$ 

so that the elasticity of substitution is $\sigma = \frac{1}{1-r} = 2$. 

Suppose as well that the consumer has income of \$100 and that the prices of good one and good two are \$2 and \$1, respectively. 

1. Computational solution. Find the optimal bundle using `solnp`

```{r computeSolve, exercise = TRUE, exercise.reveal_solution = TRUE}

```

```{r computeSolve-solution}
r = 0.5
p1 = 2
p2 = 1
m = 100
ux <- function(x) {
  - (x[1]^r + x[2] ^ r) ^ (1/r)
}
bc <- function(x) {
  p1 * x[1] + p2 * x[2]
}
equilibrium <- solnp(
  c(1, 1),
  ux,
  ineqfun = bc,
  ineqLB = 0,
  ineqUB = m,
  LB = c(0, 0),
  UB = c(Inf, Inf)
)
equilibrium$pars
-equilibrium$lagrange
-tail(equilibrium$values, 1)
```

```{r computeSolve-check}
grade_this({if (!inherits(.result, c("numeric"))) {
    fail("Your class of your answer should be a numeric scalar")}
  if (abs(tail(.result$values, 1)) != 1250) { 
    fail("The maximum of the utility function should be 1,250") }
  pass()
})
```

2. Graphical solution. Graph the budget constraint with the optimal bundle and its indifference curve. 

```{r graph, exercise = TRUE, exercise.reveal_solution = TRUE, fig.height=6.5, fig.width=4.5, fig.align='center'}

```

```{r graph-solution}
curve(
  ((-tail(equilibrium$values, 1))^r - x^r) ^ (1/r),
  0,
  m / p1,
  xlab = expression(x[1]),
  ylab = expression(x[2]),
  xlim = c(0, m / p1),
  ylim = c(0, m / p2),
  col = 'darkgreen'
)
abline(m / p2, -p1 / p2,
       col = 'red')
points(equilibrium$pars[1],
       equilibrium$pars[2],
       col = 'blue',
       pch = 20)
```

```{r graph-check}

```
