---
title: "1.4 Rational Consumer Choice Application: Insurance - Practice"
output: 
  learnr::tutorial: 
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
author: 
  name: "Created by [Jim Bang](http://www.github.com/bangecon)"
  email: BangJamesT@sau.edu
  affiliation: "[St. Ambrose University](http://www.sau.edu)"
date: "`r format(Sys.Date(), '%d %b %Y')`"
---

<style type="text/css">
#doc-metadata {
  margin-bottom: 10px;
  font-size: 0.8em;
}
</style>

```{r setup, include = FALSE}
library(learnr)
library(gradethis)
library(ggcheck)
gradethis_setup()
tutorial_options(exercise.reveal_solution = FALSE)
knitr::opts_chunk$set(echo = FALSE, fig.align = "center", fig.height = 6, fig.width = 6)
knitr::include_graphics
```

## Insurance and Risk Premia

### Overview

Choice under uncertainty involves maximizing utility over different *states* or outcomes of some random event. A simple way of formulating this problem is to consider two possible states: a bad outcome in which the consumer suffers some loss of assets and occurs with probability $\pi$; and a good outcome in which the consumer keeps their entire asset endowment and occurs with probablity $1-\pi$. 

The consumer starts with an initial endowment of assets, $E$. They may choose to insure as much of that endowment against a fixed loss ($L < E$) as they want at price, or *premium*, of $p$. Without insurance, the consumer only enjoys $E - L$ (with probability $\pi$) if the bad outcome occurs; they enjoy $E$ fully (with probability $1-\pi$) if there is no loss. 

The consumer may "buy" additional units of bad-state consumption by insuring the losses at price, $p$ for each dollar of bad-state consumption they choose to insure. The consumer chooses the amount of insurance, $I$ before observing the uncertain state outcome. 

To make the problem more concrete, use the following parameter values: 
1. $E = \$100,000$
2. $L = \$50,000$
3. $\pi = 0.2$
4. $p = 0.2$

### Budget Constraint

The budget constraint for this consumer works out to be: 

$$pc_b + (1-p)c_g \le E(m)$$
where $E(m)$ is the expected income of the consumer, i.e. $E - pL$.

Plot the entire budget constraint for the consumer. Add points for where the consumer chooses not to insure (fully consuming $E$ in the good state and taking the full loss and consuming only $E-L$ in the bad state) and where they choose to fully insures (consuming $E$ minus the cost of fully insuring $L$ in both states). Include x and y limits to make sure that the plot includes the full budget line from intercept to intercept.

Store your plot with the name `g`. 

```{r insuranceBC, exercise = TRUE}

```

```{r insuranceBC-solution, messages = FALSE, warnings = FALSE}
library(ggplot2) 
E  = 100000
L  =  50000
pi =  0.2
p  =  0.2
g <- ggplot() + 
  geom_function(fun = function(x) (E - pi * L) / (1 - p) - (p / (1 - p)) * x,
                color = 'red') + 
  geom_point(aes(x = E - L, y = E), color = 'blue') + 
  geom_point(aes(x = E - p*L, y = E - p*L), color = 'blue') + 
  lims(x = c(0, (E - pi*L)/p), y = c(0, (E - pi*L)/(1-p))) + 
  labs(x = "Consumption in the 'Bad' State", y = "Consumption in the 'Good' State")
g
```

```{r insuranceBC-check}
grade_this({
  if (
    ith_geom_is(.result, "function", 1) &
    ith_geom_is(.result, "point", 2) &
    ith_geom_is(.result, "point", 3)) {
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

Add a `lims()` layer to plot the *relevant* budget constraint for the consumer between the choice not to insure and fully insuring. 

```{r insuranceBC2, exercise = TRUE}

```

```{r insuranceBC2-setup, messages = FALSE, warnings = FALSE}
library(ggplot2) 
E  = 100000
L  =  50000
pi =  0.2
p  =  0.2
g <- ggplot() + 
  geom_function(fun = function(x) (E - pi * L) / (1 - p) - (p / (1 - p)) * x,
                color = 'red') + 
  geom_point(aes(x = E - L, y = E), color = 'blue') + 
  geom_point(aes(x = E - p*L, y = E - p*L), color = 'blue') + 
  lims(x = c(0, (E - pi*L)/p), y = c(0, (E - pi*L)/(1-p))) + 
  labs(x = "Consumption in the 'Bad' State", y = "Consumption in the 'Good' State")
```

```{r insuranceBC2-solution, messages = FALSE, warnings = FALSE}
g <- g +
  lims(x = c(E - L, E - p*L), y = c(E - p*L, E)) 
g
```

```{r insuranceBC2-check}
grade_this({
  pass_if(
    layer_scales(.result)$x$range$range[1] == 50000 & 
    layer_scales(.result)$x$range$range[2] == 90000 & 
    layer_scales(.result)$y$range$range[1] == 90000 & 
    layer_scales(.result)$y$range$range[2] == 100000)
  fail(random_encouragement())
})
```

### Preferences

Suppose that the consumer has preferences that we can express as the square root of good and bad state consumption levels, and that the overall level of utility is based on the *expected value* of whichever state we observe.

$$E[u(c_b, c_g)] = \pi\sqrt{c_b} + (1-\pi)\sqrt{c_g}$$

Add layers to the previous plot for the indifference curves for the consumer when they do not insure, and when they fully insure. 

Set x and y limits that correspond to the *relevant* range of the budget constraint from the previous part of the problem. 

Note: to do this you will need to solve for consumption in the good state (the y-variable) as a function of the level of utility (e.g. $u(E-L, E)$) and different levels of consumption in the bad state (the x-variable).  

Does it look like this consumer will fully insure, not insure, or partially insure? 

```{r insuranceUtility, exercise = TRUE}

```

```{r insuranceUtility-hint}
# I stored the plot up through the previous step of the exercise as `g`.
# To get the function for the indifference curve at a specific level of utility, evaluate `ux` at that consumption bundle and invert the utility function for x2 as a function of x1. 
```

```{r insuranceUtility-setup, messages = FALSE, warnings = FALSE}
library(ggplot2) 
E  = 100000
L  =  50000
pi =  0.2
p  =  0.2
ux = function(x) {
  pi*x[1]^0.5 + (1-pi)*x[2]^0.5
}
g <- ggplot() + 
  geom_function(fun = function(x) (E - pi * L) / (1 - p) - (p / (1 - p)) * x,
                color = 'red') + 
  geom_point(aes(x = E - L, y = E), color = 'blue') + 
  geom_point(aes(x = E - p*L, y = E - p*L), color = 'blue') +
  lims(x = c(E - L, E - p*L), y = c(E - p*L, E)) + 
  labs(x = "Consumption in the 'Bad' State", y = "Consumption in the 'Good' State")
```

```{r insuranceUtility-solution, messages = FALSE, warnings = FALSE}
g <- g + 
  geom_function(
    fun = function(x) ((ux(c(E-L, E)) - pi*x^0.5)/(1-pi))^2, 
    color = 'darkgreen') +
  geom_function(
    fun = function(x) ((ux(c(E-p*L, E-p*L)) - pi*x^0.5)/(1-pi))^2, 
    color = 'darkgreen') 
g
```

```{r insuranceUtility-check}
grade_code()
```

### Equilibrium

Solve for the equilibrium using `MicroWithR::solnp2()`. 

```{r insuranceEquilibrium, exercise = TRUE}

```

```{r insuranceEquilibrium-setup, messages = FALSE, warnings = FALSE}
library(ggplot2) 
E  = 100000
L  =  50000
pi =  0.2
p  =  0.2
ux = function(x) {
  pi*x[1]^0.5 + (1-pi)*x[2]^0.5
}
```

```{r insuranceEquilibrium-hint}
# 
```

```{r insuranceEquilibrium-solution, messages = FALSE, warnings = FALSE}
library(MicroWithR) 
bc <- function(x) {
  p*x[1] + (1-p)*x[2] 
}
eq <- solnp2(
  pars = c(1,1), 
  fun = ux, 
  ineqfun = bc,
  ineqLB = -Inf, 
  ineqUB = E - pi*L, 
  LB = c(0,0),
  control = list(trace = 0),
  max = TRUE
)
eq
```

```{r insuranceEquilibrium-check}
grade_this({if (!inherits(.result, c("list"))) {
    fail("Your class of your answer should be a list")}
  if (round(ux(.result$pars), 1) != 300) { 
    fail("The maximum of the utility function should be about 300") }
  pass()
})
```

