---
title: "2.4 Competitive Firms' Output Choices in the Short Run"
output: 
  learnr::tutorial: 
    progressive: true
runtime: shiny_prerendered
author: 
  name: "Created by [Jim Bang](http://www.github.com/bangecon)"
  email: BangJamesT@sau.edu
  affiliation: "[St. Ambrose University](http://www.sau.edu)"
date: "`r format(Sys.Date(), '%d %b %Y')`"
---

<style type="text/css">
h1{font-size: 24pt}
h2{font-size: 20pt}
h3{font-size: 18pt}
h4,h5,h6{font-size: 16pt}
body{font-size: 16pt}
#doc-metadata {
  margin-bottom: 10px;
  font-size: 0.66em;
}
</style>

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(Ryacas)
library(Deriv)
gradethis_setup(exercise.reveal_solution = TRUE, exercise.eval = TRUE)
knitr::opts_chunk$set(echo = FALSE, fig.height = 6, fig.width = 6)
```

<style type="text/css">
h1{font-size: 24pt}
h2{font-size: 20pt}
h3{font-size: 18pt}
h4,h5,h6{font-size: 16pt}
body{font-size: 16pt}
</style>

## The Firm's Three Problems

1. Input cost minimization
2. Output profit maximization
3. Input profit maximization

### Competitive Input Markets

$$\matrix{ \matrix{\text{1. Many, many buyers} \\ \text{2. Homogeneous inputs} \\ \text{3. Perfect information}} \Biggr\} \\ \text{4. Free entry/exit} \\ \text{  }} \matrix{\text{ } \\ \implies \\ \text{ } \\ \implies \\ \text{ } }  \matrix{\Biggl\{ \matrix{\text{Firms are} \\ \text{price} \\ \text{takers}} \\ \text{Profits/losses} \\ \text{are temporary}}$$

### The Producer's Input-Hiring Problem

1. Objective: 
2. Endogenous (choice) variable(s): 
3. Exogenous variable(s): 

### Short Run

For simplicity we will solve the model using a Cobb-Douglas production function with input shares 1/3 and 2/3, and TFP parameter 100: 

$$f(L,K) = 100L^{1/3}K^{2/3}$$.

The firm's goal is to choose inputs optimally to maximize profits. 

$$\max_{L,K} \{ \pi(L,K) = P \cdot f(L,K) - (wL + rK) \} $$

From the firm's perspective, markets determine input and output prices *exogenously*. Set $P = 10$, $w = 1$, $r = 2$

In the short run, firms' previous choice of capital firms constrain their choices. Set $K = \bar{K} = 25$

Inserting all of these substitutions and simplifying, we have:

$$\max_{L} \{ 5000 \cdot (5L)^{1/3} - (L + 100) \}$$.

## Analytical Approach

```{r}

```

```{r}
w = 1
r = 2
pi  <- ysym('10*100*L^(1/3)*K^(2/3) - (w*L + r*K)')
dpi <- deriv(pi, c('L', 'K')) 
MPL <- solve(dpi[1], 'w') |>
  y_rmvars()
MPK <- solve(dpi[2], 'r') |>
  y_rmvars()
MRTS <- MPL/MPK 
MRTS <- yac_str(MRTS) |>
  Simplify() |> 
  ysym()
Kstar <- solve(MRTS, w/r, 'K') |>
  y_rmvars()
MPL <- with_value(MPL, 'K', Kstar) 
solution <- solve(MPL, w, 'L')
```

```{r}

```

## Graphical Approach



```{r}

```

```{r}

```

```{r}

```

## Numerical Approach



```{r}

```

```{r}

```

```{r}

```
