---
title: "3.1 Market Demand in Competitive Markets"
output: 
  learnr::tutorial: 
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
author: 
  name: "Created by [Jim Bang](http://www.github.com/bangecon)"
  email: BangJamesT@sau.edu
  affiliation: "[St. Ambrose University](http://www.sau.edu)"
date: "`r format(Sys.Date(), '%d %b %Y')`"
---

<style type="text/css">
h1{font-size: 24pt}
h2{font-size: 20pt}
h3{font-size: 18pt}
h4,h5,h6{font-size: 16pt}
body{font-size: 16pt}
#doc-metadata {
  margin-bottom: 10px;
  font-size: 0.66em;
}
</style>

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(ggcheck)
gradethis_setup(exercise.reveal_solution = TRUE, exercise.eval = TRUE)
knitr::opts_chunk$set(echo = FALSE, fig.height = 6, fig.width = 6)
```

## Deriving Market Demand

Market demand is the "horizontal sum" (adding up the quantities along the x-axis) of the individual demand curves. 

### Review: The Consumer's Problem

The consumer problem maximizes utility subject to the budget constraint. For example, using a Cobb-Douglas utility function with budget shares 0.25 and 0.75 on goods 1 and 2, respectively; prices 0.75 and 1.0; and income equal to 100. 

$$\max_x \bigg\{ U(x_1, x_2) = x_1^{0.25}x_2^{0.75} \bigg\}$$

$$\text{s.t.: }0.75x_1+x_2 \le 100$$
The "Lagrangian" function for our example is: 

$$\mathcal{L} = x_1^{0.25}x_2^{0.75} + \lambda (100 - 1.5x_1 - x_2)$$

Solve this problem for the given prices as a warm-up. Print the optimal consumption (`pars`), Lagrange multiplier (`lagrange`), and value of the utility function at the optimal consumption. 

```{r singleConsumer, exercise = TRUE, exercise.reveal_solution = TRUE}

```

```{r singleConsumer-solution}
library(MicroWithR)
a = 0.25
p1 = 1.5
p2 = 1
m = 30    
ux <-
  function(x)
    x[1] ^ a * x[2] ^ (1 - a) 
bc <-
  function(x)
    p1 * x[1] + p2 * x[2]    
eq <- solnp2(
  c(1, 1),
  ux,
  ineqfun = bc,
  ineqLB = 0,
  ineqUB = m,
  LB = c(0, 0),
  UB = c(Inf, Inf), 
  max = TRUE
)
eq
```

```{r singleConsumer-check}
grade_this({if (round(.result$pars, 1) != c(5.0 ,22.5)) { 
    fail("The utility-maximizing bundle is 5 units of good 1 and 22.5 units of good 2.") }
  pass()
})
```

### Individual Consumer Demand

In module 1, we derived the individual consumer demand in two ways: (1) analytically, solving $x_1$ as a function of income, preferences, and prices using `Ryacas`; (2) computationally, solving for many numerical values of $x_1$ changing only $p_1$, while holding all other parameters constant. 

Since this involves a `for()` loop, I've coded the computational method for you here: 

```{r individualDemandSetup}
library(MicroWithR)
a = 0.25
p1 = 1.5
p2 = 1
m = 30    
ux <-
  function(x)
    x[1] ^ a * x[2] ^ (1 - a) 
bc <-
  function(x)
    p1 * x[1] + p2 * x[2]    
```

```{r individualDemand, echo = TRUE}
# p1 is a vector to loop through the different prices.
p1 <- seq(0.25, 2.5, 0.05)
bc <- function(x) p1[i]*x[1] + p2*x[2] 
c1 <- NULL
for (i in 1:length(p1)) {
  if(i == 1) {
    pars <- c(1,1)
  } else {
    pars <- c
  }
  c <- solnp2(
    pars = pars,
    ux,
    ineqfun = bc,
    ineqLB = 0,
    ineqUB = m,
    LB = c(0, 0),
    UB = c(Inf, Inf), 
    control = list(trace = 0), 
    max = TRUE
  )$pars
  c1[i] <- c[1]
}
head(c1)
```

Plot the graph of the consumer's demand function. 

```{r individualDemandCurve, exercise = TRUE}

```

```{r individualDemandCurve-setup}
library(MicroWithR)
a = 0.25
p1 = 1.5
p2 = 1
m = 30    
ux <-
  function(x)
    x[1] ^ a * x[2] ^ (1 - a) 
bc <-
  function(x)
    p1 * x[1] + p2 * x[2]    
# p1 is a vector to loop through the different prices.
p1 <- seq(0.25, 2.5, 0.05)
bc <- function(x) p1[i]*x[1] + p2*x[2] 
c1 <- NULL
for (i in 1:length(p1)) {
  if(i == 1) {
    pars <- c(1,1)
  } else {
    pars <- c
  }
  c <- solnp(
    pars = pars,
    ux,
    ineqfun = bc,
    ineqLB = 0,
    ineqUB = m,
    LB = c(0, 0),
    UB = c(Inf, Inf), 
    control = list(trace = 0)
  )$pars
  c1[i] <- c[1]
}
```

```{r individualDemandCurve-hint}
# In the computational solution we defined numerical vectors containing the values for the prices of good 1 ranging from 0.25 to 2.5 (`p1`) and the corresponding consumption levels of good 1 (`c1`). All you need to do is plot the connection of these points using `geom_line`. 
```

```{r individualDemandCurve-solution}
library(ggplot2)
ggplot() + 
  geom_line(aes(x = c1, y = p1, color = "D")) + 
  lims(x = c(0, 25.1),y = c(0, 2.5)) + 
  labs(x = "Quantity", y = "Price", color = "") + 
  scale_color_manual(breaks = c("D"), values = c("blue"))
```

```{r individualDemandCurve-check}
grade_this({
  if (uses_geoms(.result, "line")) {
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

The analytical solution for the demand curve was:

<details><summary>
Click here to see the code for deriving the demand function using `Ryacas` functions.
</summary>
```{r demand, echo = TRUE, message = FALSE, warning = FALSE}
library(Ryacas)
library(Deriv)
u <- ysym('x1^(alpha) * x2^(1-alpha)')
L <- u + ysym('lambda*(m - p1*x1 - p2*x2)')
dL <- deriv(L, c('x1', 'x2', 'lambda'))
MU <- deriv(u, c('x1', 'x2'))
MRS <- yac_str(MU[1]/MU[2]) |> 
  Simplify() |> 
  ysym()
D <- solve(c(MRS, dL[[3]]), c(ysym('p1/p2'), 0), c('x1', 'x2')) |> 
  y_rmvars() |>
  simplify()
D[1] <- with_value(D[1], 'x2', D[2]) |> 
  simplify()
```
</details>

$$x_1(p_1, \alpha, m) = `r tex(D[1])`$$

### Identical Consumers

A simple (and in real life wrong) thing we could do is assume all consumers have identical incomes and preferences. If this is the case then demand for good on is simply:  
$$D_1(p_1, \alpha, m) = N `r tex(D[1])`,$$
where $N$ is the number of consumers that comprise the market.  

Plot the market demand curve for good 1 for a market with 200,000 consumers who are all identical to the one for whom we solved the equilibrium in the previous exercises. 

```{r marketDemand, exercise = TRUE}
```

```{r marketDemand-setup}
library(ggplot2)
```

```{r marketDemand-hint}
# First you will need to add up the individual demands, but then you will need to invert the demand curve to get price as the "dependent" (y) variable.
# Never mind the fact that price isn't really the dependent variable that the consumer chooses. Blame Marshall!
```

```{r marketDemand-solution}
ggplot() + 
  geom_function(fun = function(x) 1500000/x, aes(color = "D")) + 
  lims(x = c(0,2500000), y = c(0, 2.5)) +
  labs(x = "Quantity", y = "Price", color = "") + 
  scale_color_manual(breaks = c("D"), values = c("blue"))
```

```{r marketDemand-check}
grade_this({
  if (uses_geoms(.result, "function") | uses_geoms(.result, "line")) {
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

## Appendix A: Heterogeneous Consumers

### Setup

Heterogeneous consumers is trickier, because we must solve each consumer's choice problem separately. But, if we can at least assume consumers have at least the same *type* of preferences (e.g. Cobb-Douglas or constant elasticity of substitution), we can at least use a loop to solve the problem. What's more, we can assign the relevant household parameters (income, preferences) randomly. This isn't *too* hard in R since it just requires a second "for" loop (this time over consumers) within the for loop over different prices. 

<details style="line-height:105%"><summary>Click here to view the code for the graph.</summary>
1. `runif()` to randomly assign 200 different values of the good-1 budget share parameter, $\alpha$, according to a uniform continuous distribution between 0.1 and 0.9.
2. `rlnorm()` to randomly assign 200 incomes, $m$, according to a lognormal distribution with a mean of about 100. 
3. Initialize a matrix to hold the consumption values for each consumer and price with rows equal to the number of different values of price and columns equal to the number of consumers.
</details>
  
<br>  
  
The code below sets up the parameter vectors for a loop over prices between 0.25 and 2.5 for 200 heterogeneous consumers and prints the basic summary statistics for the budget-share parameters ($\alpha$) and incomes ($m$). 

```{r differentConsumersSetup, echo = TRUE}
library(MicroWithR)
set.seed(8675309) # Set the random "seed" so we all get the exact same results
p1 <- seq(0.25, 2.5, 0.05)
p2 <- 1
a <- runif(200, min = 0.01, max = 0.49) # Sequence of random budget shares
m <- rlnorm(200, meanlog = log(30), sdlog = 0.25) # Sequence of random incomes
c1 <- matrix(nrow = length(p1), ncol = length(a)) # Matrix to store quantitites
ux <- function(x)
    - x[1] ^ a[j] * x[2] ^ (1 - a[j]) 
bc <- function(x)
  p1[i] * x[1] + p2 * x[2]
summary(cbind(a, m))
```

### Individual Demand Schedules

The following code to builds a matrix of individual demand schedules for good 1 (`c1`). The output prints (1) the sum of the consumers' individual quantities along the rows at each price; (2) the summary statistics for the market demand variable. 

```{r differentConsumersEquilibrium, echo = TRUE}
for (i in 1:length(p1)) {
  if (i == 1) {
    pars <- c(1, 1)
  } else {
    pars <- c$pars
  }
  for (j in 1:length(m)) {
    c <-
      solnp(
        pars = pars,
        ux,
        ineqfun = bc,
        ineqLB = 0,
        ineqUB = m[j],
        LB = c(0, 0),
        UB = c(Inf, Inf),
        control = list(trace = 0)
      )
    c1[i, j] <- c$pars[1]
  }
}
marketDemand <- rowSums(c1)
summary(marketDemand)
```

### Heterogeneous Consumers - Individual Demand Curves

The following code plots the individual demand curves for the first 10 consumers (column) of the individuals' demand schedules. Rainbow the tastes!

```{r differentConsumersPlots, excercise = TRUE}
ggplot() + 
  geom_line(aes(x = c1[, 1], y = p1)) + 
  geom_line(aes(x = c1[, 2], y = p1), color = "green3") + 
  geom_line(aes(x = c1[, 3], y = p1), color = "darkblue") + 
  geom_line(aes(x = c1[, 4], y = p1), color = "darkorange2") + 
  geom_line(aes(x = c1[, 5], y = p1), color = "red3") + 
  geom_line(aes(x = c1[, 6], y = p1), color = "blue") + 
  geom_line(aes(x = c1[, 7], y = p1), color = "purple3") + 
  geom_line(aes(x = c1[, 8], y = p1), color = "brown4") + 
  geom_line(aes(x = c1[, 9], y = p1), color = "gray") + 
  geom_line(aes(x = c1[, 10], y = p1), color = "gold2") + 
  labs(x = "Quantity", y = "Price") + 
  lims(x = c(0,30), y = c(0, 2.5))
```

### Heterogeneous Consumers - Market Demand Curve

The next chunk plots the market demand for the 200 consumers. 

```{r marketEquilibrium-solution}
ggplot() + 
  geom_line(aes(x = marketDemand, y = p1), color = "blue") + 
  labs(x = "Quantity", y = "Price") + 
  lims(x = c(0, 5000), y = c(0 ,2.5))
```

This is a case where computational methods are a clear win! deriving the sum of 200 different and nonlinear demand functions would be much more tedious, if even possible at all! 
