---
title: "1.5 Comparative Statics"
editor_options: 
  markdown: 
    wrap: 72
output: 
  learnr::tutorial: 
    progressive: true
runtime: shiny_prerendered
author: 
  name: "Created by [Jim Bang](http://www.github.com/bangecon)"
  email: BangJamesT@sau.edu
  affiliation: "[St. Ambrose University](http://www.sau.edu)"
date: "`r format(Sys.Date(), '%d %b %Y')`"
---

<style type="text/css">
h1{font-size: 24pt}
h2{font-size: 20pt}
h3{font-size: 18pt}
h4,h5,h6{font-size: 16pt}
body{font-size: 16pt}
#doc-metadata {
  margin-bottom: 10px;
  font-size: 0.66em;
}
</style>

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(Deriv)
library(Ryacas)
library(Rsolnp)
gradethis_setup()
tutorial_options(exercise.eval = TRUE, exercise.reveal_solution = TRUE)
knitr::opts_chunk$set(echo = FALSE, fig.align = "center", fig.height = 6, fig.width = 6)
knitr::include_graphics
par(mar = c(4, 4, 1, 1))
u <- ysym('x1^(alpha) * x2^(1-alpha)')
L <- u + ysym('lambda*(m - p1*x1 - p2*x2)')
dL <- deriv(L, c('x1', 'x2', 'lambda'))
MU <- deriv(u, c('x1', 'x2'))
MRS <- yac_str(MU[1]/MU[2]) |> 
  Simplify() |> 
  ysym()
D <- solve(c(MRS, dL[[3]]), c(ysym('p1/p2'), 0), c('x1', 'x2')) |> 
  y_rmvars() |>
  simplify()
D[1] <- with_value(D[1], 'x2', D[2]) |> 
  simplify()
ux <- function(x) -x[1] ^ (a) * x[2] ^ (1 - a)
bc <- function(x) p1 * x[1] + p2 * x[2]
bcLowP  <- function(x) (p1 - dp1) * x[1] + p2 * x[2]
bcHighP <- function(x) (p1 + dp1) * x[1] + p2 * x[2]
a = 0.25; p1 = 1.5; p2 = 1; m = 100; dm = 5; dp1 = 0.5
initialEquilibrium <- solnp(c(1, 1), ux, ineqfun = bc, ineqLB = 0, ineqUB = m, 
                            LB = c(0, 0), UB = c(Inf, Inf))
dmEquilibrium <- solnp(c(1, 1), ux, ineqfun = bc, ineqLB = 0, ineqUB = m + dm, 
                            LB = c(0, 0), UB = c(Inf, Inf))
dmEquilibrium2 <- solnp(c(1, 1), ux, ineqfun = bc, ineqLB = 0, ineqUB = m - dm, 
                            LB = c(0, 0), UB = c(Inf, Inf))
lowPEquilibrium <- solnp(c(1, 1), ux, ineqfun = bcLowP, ineqLB = 0, ineqUB = m, 
                            LB = c(0, 0), UB = c(Inf, Inf))
highPEquilibrium <- solnp(c(1, 1), ux, ineqfun = bcHighP, ineqLB = 0, ineqUB = m, 
                            LB = c(0, 0), UB = c(Inf, Inf))
u0 <- -tail(initialEquilibrium$values, 1)
x0 <- c(initialEquilibrium$pars)
xTilde <- solnp(pars = x0, fun = bcLowP, ineqfun = ux, ineqLB = -Inf, 
                ineqUB = -u0, LB = c(0, 0), UB = c(Inf, Inf))
```

## Overview

### Basic Idea

Make a *hypothesis* of what would happen if *one* of the exogenous
determinants changes.

1.  Ceteris paribus
2.  Small changes

### Determinants

In the Cobb-Douglas example, the demand function for good $i$ depends on: 

1. Preferences (the budget-share parameters, $\alpha$ and $1 - \alpha$);
2. Income ($m$); 
3. The price of good $i$ ($p_i$);
4. The price(s) of the other good(s) ($p_j$).

We will focus mainly on income and price. 

### Types of Hypotheses

1.  Qualitative: Which direction?
2.  Quantitative: How much?

### Approaches

1.  Graphical (as long as precise values aren't important)
2.  Computational
3.  Anayltical

## Computational Approach

### Initial Solution

$$x_1^* = 50/3$$ 

$$x_2^* = 75$$

### Income Change

To measure the effect computationally, we have to specify a numerical
amount for the income change, say \$5.

1.  For a consumer with the same preferences (Cobb-Douglass with
    $\alpha = 0.25$), and facing the same prices ($p_1 = 1.5$;
    $p_2 = 5$), compute the new equilibrium for an increase in income
    from \$100 to \$105 (`dm = 5`). Print the (already-stored) initial
    and new optimal bundles and the initial and new values of the
    utility function.

```{r cdIncomeChange, exercise = TRUE, exercise.reveal_solution = TRUE}

```

```{r cdIncomeChange-solution}
dm = 5
dmEquilibrium <-
  solnp(
    c(1, 1),
    ux,
    ineqfun = bc,
    ineqLB = 0,
    ineqUB = m + dm,
    LB = c(0, 0),
    UB = c(Inf, Inf)
  )
initialEquilibrium[c(1, 3)]
dmEquilibrium[c(1, 3)]
```

```{r cdIncomeChange-check}
grade_code()
```

$$x_1^{**} = `r round(dmEquilibrium$pars[1], 3)`$$ 

$$x_2^{**} = `r round(dmEquilibrium$pars[2], 3)`$$

$$\frac{\Delta x_1}{\Delta m} \Bigg|_{x_1 = x_1^*} \approx `r round((dmEquilibrium$pars[1] - initialEquilibrium$pars[1])/dm, 3)`$$

### Price Change 

Let's analyze the effect of a \$0.50 increase in price, and a \$0.50
decrease in price for good 1.

1.  For a consumer with the same preferences (Cobb-Douglass with
    $\alpha = 0.25$), and facing the price of good 2 and income
    ($p_2 = 1$), compute the new equilibria for an increase in the price
    of good 1 from \$1.50 to \$2.00 (`dp1 = 0.5`); repeat for an equal
    decrease in $p_1$ to \$1.00.
2.  Print the (already-stored) initial and new optimal bundles.

```{r priceChange, exercise = TRUE, exercise.reveal_solution = TRUE}


```

```{r priceChange-solution}
dp1 = 0.5
lowPEquilibrium <-
  solnp(
    c(1, 1),
    ux,
    ineqfun = bcLowP,
    ineqLB = 0,
    ineqUB = m,
    LB = c(0, 0),
    UB = c(Inf, Inf)
  )
highPEquilibrium <-
  solnp(
    c(1, 1),
    ux,
    ineqfun = bcHighP,
    ineqLB = 0,
    ineqUB = m,
    LB = c(0, 0),
    UB = c(Inf, Inf)
  )
initialEquilibrium[1]
lowPEquilibrium[1]
highPEquilibrium[1]
```

```{r priceChange-check}
grade_code()
```

Consumption of good 1 after the price decrease is: 

$$x_1^{\dagger} = `r round(lowPEquilibrium$pars[1], 3)`$$  

Consumption of good 1 after the price increase is: 

$$x_1^{\dagger\dagger} = `r round(highPEquilibrium$pars[1], 3)`$$

$$\frac{\Delta x_1}{\Delta p_1} \Bigg|_{x_1 = x_1^*} \approx `r round((highPEquilibrium$pars[1] - lowPEquilibrium$pars[1])/(dp1*2), 3)`$$

## Graphical Approach

### Income Change

Run the following code to see the graph of the effect of an increase and
decrease in income equal to 5.

```{r incomeChangeGraph, exercise = TRUE}
curve(
  (-tail(initialEquilibrium$values, 1)) ^ (1 / (1 - a)) / (x ^ (a / (1 - a))),
  from = 0,
  to = m / p1,
  xlim = c(0, (m + dm) / p1),
  ylim = c(0, (m + dm) / p2),
  xlab = expression(x[1]),
  ylab = expression(x[2]),
  col = 'dark green'
)
abline(m / p2,-p1 / p2, col = 'red')
points(initialEquilibrium$pars[1],
       initialEquilibrium$pars[2],
       col = 'blue',
       pch = 20)
curve(
  (-tail(dmEquilibrium$values, 1)) ^ (1 / (1 - a)) / (x ^ (a / (1 - a))),
  add = TRUE,
  col = 'dark green',
  lty = 'dashed'
)
abline((m + dm) / p2,-p1 / p2, col = 'red', lty = 'dashed')
points(dmEquilibrium$pars[1],
       dmEquilibrium$pars[2],
       col = 'blue',
       pch = 20)
curve(
  (-tail(dmEquilibrium2$values, 1)) ^ (1 / (1 - a)) / (x ^ (a / (1 - a))),
  add = TRUE,
  col = 'dark green',
  lty = 'dashed'
)
abline((m - dm) / p2,-p1 / p2, col = 'red', lty = 'dashed')
points(dmEquilibrium2$pars[1],
       dmEquilibrium2$pars[2],
       col = 'blue',
       pch = 20)
arrows(
  x0 = 0,
  y0 = 0,
  x1 = dmEquilibrium$pars[1],
  y1 = dmEquilibrium$pars[2]
)
```

```{r incomeChangeGraph-solution}
curve(
  (-tail(initialEquilibrium$values, 1)) ^ (1 / (1 - a)) / (x ^ (a / (1 - a))),
  from = 0,
  to = m / p1,
  xlim = c(0, (m + dm) / p1),
  ylim = c(0, (m + dm) / p2),
  xlab = expression(x[1]),
  ylab = expression(x[2]),
  col = 'dark green'
)
abline(m / p2,-p1 / p2, col = 'red')
points(initialEquilibrium$pars[1],
       initialEquilibrium$pars[2],
       col = 'blue',
       pch = 20)
curve(
  (-tail(dmEquilibrium$values, 1)) ^ (1 / (1 - a)) / (x ^ (a / (1 - a))),
  add = TRUE,
  col = 'dark green',
  lty = 'dashed'
)
abline((m + dm) / p2,-p1 / p2, col = 'red', lty = 'dashed')
points(dmEquilibrium$pars[1],
       dmEquilibrium$pars[2],
       col = 'blue',
       pch = 20)
curve(
  (-tail(dmEquilibrium2$values, 1)) ^ (1 / (1 - a)) / (x ^ (a / (1 - a))),
  add = TRUE,
  col = 'dark green',
  lty = 'dashed'
)
abline((m - dm) / p2,-p1 / p2, col = 'red', lty = 'dashed')
points(dmEquilibrium2$pars[1],
       dmEquilibrium2$pars[2],
       col = 'blue',
       pch = 20)
arrows(
  x0 = 0,
  y0 = 0,
  x1 = dmEquilibrium$pars[1],
  y1 = dmEquilibrium$pars[2]
)
```

```{r incomeChangeGraph-check}
grade_code()
```

In the Cobb-Douglas example, consumption of each good increases by equal
proportions to the change in income (partly because of the fact that
consumers spend a constant share of their budget on each good). This
means that the *income expansion path* is a straight line from the
origin. <br> (In fact, a linear income expansion path is a feature of
many utility functions that have the property of being
[*homothetic*](https://en.wikipedia.org/wiki/Homothetic_preferences),
which also includes the CES utility function.)

### Price Change

Graph the indifference curves and budget lines for each equilibrium:

1.  $p_1 = \$1.50$ (Plot from $x_1 = 0$ to $x_1 = m/p_1$ and from
    $x_2 = 0$ to $x_2 = m/p_2$; indifference curve `dark green` and
    budget line `red`)
2.  $p_1 = \$2.00$ (Indifference curve )
3.  $p_1 = \$1.00$

```{r priceChangeGraph, exercise = TRUE, exercise.reveal_solution = TRUE}
# Solve for consumption (and utility at that consumption) for the original price, the price after an increase, and after a decrease. 
dp1 = 0.5
bcLowP  <- function(x) (p1 - dp1) * x[1] + p2 * x[2]
bcHighP <- function(x) (p1 + dp1) * x[1] + p2 * x[2]
lowPEquilibrium <- solnp(c(1, 1), ux, ineqfun = bcLowP, ineqLB = 0, ineqUB = m,
    LB = c(0, 0), UB = c(Inf, Inf))
highPEquilibrium <- solnp(c(1, 1), ux, ineqfun = bcHighP, ineqLB = 0, ineqUB = m,
    LB = c(0, 0), UB = c(Inf, Inf))
# Plot the indifference curves and budget lines below.


```

```{r priceChangeGraph-solution}
dp1 = 0.5
bcLowP  <- function(x) (p1 - dp1) * x[1] + p2 * x[2]
bcHighP <- function(x) (p1 + dp1) * x[1] + p2 * x[2]
lowPEquilibrium <-
  solnp(
    c(1, 1),
    ux,
    ineqfun = bcLowP,
    ineqLB = 0,
    ineqUB = m,
    LB = c(0, 0),
    UB = c(Inf, Inf)
  )
highPEquilibrium <-
  solnp(
    c(1, 1),
    ux,
    ineqfun = bcHighP,
    ineqLB = 0,
    ineqUB = m,
    LB = c(0, 0),
    UB = c(Inf, Inf)
  )
curve(
  (-tail(initialEquilibrium$values, 1)) ^ (1 / (1 - a)) / (x ^ (a / (1 - a))),
  from = 0,
  to = m / p1,
  xlim = c(0, m / (p1 - dp1)),
  ylim = c(0, m / p2),
  xlab = expression(x[1]),
  ylab = expression(x[2]),
  col = 'dark green'
)
abline(m / p2,-p1 / p2, col = 'red')
curve(
  (-tail(lowPEquilibrium$values, 1)) ^ (1 / (1 - a)) / (x ^ (a / (1 - a))),
  add = TRUE,
  col = 'dark green',
  lty = 'dashed'
)
abline(m / p2,-(p1 - dp1) / p2, col = 'red', lty = 'dashed')
curve(
  (-tail(highPEquilibrium$values, 1)) ^ (1 / (1 - a)) / (x ^ (a / (1 - a))),
  add = TRUE,
  col = 'dark green',
  lty = 'dashed'
)
abline(m / p2,-(p1 + dp1) / p2, col = 'red', lty = 'dashed')
```

```{r priceChangeGraph-check}
grade_code()
```

## Analytical Approach

Basic approach: 

1. Find the demand functions in general form, without specific parameter values for $m$, $\alpha$, $p_1$, and $p_2$. 
2. Take the derivative of the demand functions with respect to each variable. 

R can actually do this super-fast (by vectorizing the derivatives)!

Solving the first order conditions for $x_1$ and $x_2$ in terms of the parameters
(the general-form solution from last time), we get the demand functions. Run the following code to get this result.

```{r demand, exercise = TRUE}
u <- ysym('x1^(alpha) * x2^(1-alpha)')
L <- u + ysym('lambda*(m - p1*x1 - p2*x2)')
dL <- deriv(L, c('x1', 'x2', 'lambda'))
MU <- deriv(u, c('x1', 'x2'))
MRS <- yac_str(MU[1]/MU[2]) |> 
  Simplify() |> 
  ysym()
D <- solve(c(MRS, dL[[3]]), c(ysym('p1/p2'), 0), c('x1', 'x2')) |> 
  y_rmvars() |>
  simplify()
D[1] <- with_value(D[1], 'x2', D[2]) |> 
  simplify()
```

```{r demand-solution}
u <- ysym('x1^(alpha) * x2^(1-alpha)')
L <- u + ysym('lambda*(m - p1*x1 - p2*x2)')
dL <- deriv(L, c('x1', 'x2', 'lambda'))
MU <- deriv(u, c('x1', 'x2'))
MRS <- yac_str(MU[1]/MU[2]) |> 
  Simplify() |> 
  ysym()
D <- solve(c(MRS, dL[[3]]), c(ysym('p1/p2'), 0), c('x1', 'x2')) |> 
  y_rmvars() |>
  simplify()
D[1] <- with_value(D[1], 'x2', D[2]) |> 
  simplify()
```

```{r demand-check}
grade_code()
```

$$x_1(p_1, m) = `r tex(D[1])`$$

$$x_2(p_2, m) = `r tex(D[2])`$$

Notice that for these preferences the increase in consumption for $x_1$ is linear in the amount of income (it only depends on the exogenous parameters $\alpha$ and $p_1$).

Run the following code to take the derivative of demand for $x_1$ and $x_2$ with respect to income and prices. 

```{r analyticalStatics, exercise = TRUE}
statics <- deriv(D, c('m', 'p1', 'p2'))
```

```{r analyticalStatics-solution}
statics <- deriv(D, c('m', 'p1', 'p2'))
```

```{r analyticalStatics-check}
grade_code()
```

substituting our parameter values, we have: 

$$\nabla x_1(m, p_1, p_2) = \begin{pmatrix} \frac{d}{dm}x_1(m, p_1) \\ \frac{d}{dp_1}x_1(m, p_1) \\ \frac{d}{dp_2}x_1(m, p_1) \end{pmatrix} 
 = \begin{pmatrix} `r tex(statics[1,1])` \\  `r tex(statics[2,1])` \\ `r tex(statics[3,1])` \end {pmatrix} 
 = \begin{pmatrix} \frac{0.25}{1.50} \\ \frac{-25}{1.50^2} \\ 0 \end{pmatrix}
 \approx \begin{pmatrix} 0.16667 \\ -11.11 \\ 0 \end{pmatrix}$$

and 

$$\nabla x_2(m, p_1, p_2) = \begin{pmatrix} \frac{d}{dm}x_2(m, p_1) \\ \frac{d}{dp_1}x_2(m, p_1) \\ \frac{d}{dp_2}x_2(m, p_1) \end{pmatrix} 
 = \begin{pmatrix} `r tex(statics[1,2])` \\  `r tex(statics[2,2])` \\ `r tex(statics[3,2])` \end {pmatrix} 
 = \begin{pmatrix} \frac{0.75}{1} \\ 0 \\ \frac{-75}{1^2} \end{pmatrix}
 \approx \begin{pmatrix} 0.75 \\ 0 \\ -75 \end{pmatrix}$$

## Engel Curves

The *Engel Curve* plots the demand for a good against the level of
income. In our Cobb-Douglas example, this "curve" is a straight line.
Plot the Engel curve for the example.

```{r cdEngel, exercise = TRUE, exercise.reveal_solution = TRUE}

```

```{r cdEngel-solution}
curve(
  (a / p1) * x,
  from = 0,
  to = 150,
  xlab = expression("Income"),
  ylab = expression(x[1])
)
```

```{r cdEngel-check}
grade_code()
```

A "simple" example of a nonlinear income expansion path comes from the
special case of perfect complements, where:

$$u(x) = min\{x_1^2, x_2\}.$$

In this example, for any nonzero set of prices, the consumer would
choose to set $x_2 = x_1^2$ (to put consumption in the corner of the
"L". Run the following code to see the analytical solution and its graph
for different levels of income.

```{r nonlinearIEP, exercise = TRUE}
m4 = seq(60, 120, 20)
complements <- function(x) {
  - min(x[1] ^ 2, 25 * x[2])
}
nonlinearIEPEquilibrium060m <-
  solnp(
    c(1, 1),
    complements,
    ineqfun = bc,
    ineqLB = 0, 
    ineqUB = m4[1],
    LB = c(0,0), 
    UB = c(Inf, Inf)
  )
nonlinearIEPEquilibrium080m <-
  solnp(
    c(1, 1),
    complements,
    ineqfun = bc,
    ineqLB = 0, 
    ineqUB = m4[2],
    LB = c(0,0), 
    UB = c(Inf, Inf)
  )
nonlinearIEPEquilibrium100m <-
  solnp(
    c(1, 1),
    complements,
    ineqfun = bc,
    ineqLB = 0, 
    ineqUB = m4[3],
    LB = c(0,0), 
    UB = c(Inf, Inf)
  )
nonlinearIEPEquilibrium120m <-
  solnp(
    c(1, 1),
    complements,
    ineqfun = bc,
    ineqLB = 0, 
    ineqUB = m4[4],
    LB = c(0,0), 
    UB = c(Inf, Inf)
  )
par(mfrow = c(1, 3))
plot(
  x = rep(nonlinearIEPEquilibrium060m$pars[1], 10),
  y = seq(nonlinearIEPEquilibrium060m$pars[2], 120, length.out = 10),
  type = 'l',
  col = 'dark green',
  xlim = c(0, m4[4] / p1),
  ylim = c(0, m4[4] / p2),
  xlab = expression(x[1]),
  ylab = expression(x[2])
)
segments(
  x0 = nonlinearIEPEquilibrium060m$pars[1],
  x1 = 120,
  y0 = nonlinearIEPEquilibrium060m$pars[2],
  y1 = nonlinearIEPEquilibrium060m$pars[2],
  col = 'dark green'
)
segments(
  x0 = nonlinearIEPEquilibrium080m$pars[1],
  x1 = nonlinearIEPEquilibrium080m$pars[1],
  y0 = nonlinearIEPEquilibrium080m$pars[2],
  y1 = 120,
  col = 'dark green'
)
segments(
  x0 = nonlinearIEPEquilibrium080m$pars[1],
  x1 = 120,
  y0 = nonlinearIEPEquilibrium080m$pars[2],
  y1 = nonlinearIEPEquilibrium080m$pars[2],
  col = 'dark green'
)
segments(
  x0 = nonlinearIEPEquilibrium100m$pars[1],
  x1 = nonlinearIEPEquilibrium100m$pars[1],
  y0 = nonlinearIEPEquilibrium100m$pars[2],
  y1 = 120,
  col = 'dark green'
)
segments(
  x0 = nonlinearIEPEquilibrium100m$pars[1],
  x1 = 120,
  y0 = nonlinearIEPEquilibrium100m$pars[2],
  y1 = nonlinearIEPEquilibrium100m$pars[2],
  col = 'dark green'
)
segments(
  x0 = nonlinearIEPEquilibrium120m$pars[1],
  x1 = nonlinearIEPEquilibrium120m$pars[1],
  y0 = nonlinearIEPEquilibrium120m$pars[2],
  y1 = 120,
  col = 'dark green'
)
segments(
  x0 = nonlinearIEPEquilibrium120m$pars[1],
  x1 = 120,
  y0 = nonlinearIEPEquilibrium120m$pars[2],
  y1 = nonlinearIEPEquilibrium120m$pars[2],
  col = 'dark green'
)
abline(100 / p2,-p1 / p2, col = 'red')
curve(x ^ 2 / 25, add = TRUE)
curve(
  p2 * x ^ 2 / 25 + p1 * x,
  xlim = c(0, nonlinearIEPEquilibrium120m$pars[1]),
  ylim = c(0, m4[4]),
  xlab = expression(x[1]),
  ylab = 'Income'
)
curve(
  p2 * x + 5 * p1 * sqrt(x),
  xlim = c(0, nonlinearIEPEquilibrium120m$pars[2]),
  ylim = c(0, m4[4]),
  xlab = expression(x[2]),
  ylab = 'Income'
)
par(mfrow = c(1,1))
```

```{r nonlinearIEP-solution}
m4 = seq(60, 120, 20)
complements <- function(x) {
  - min(x[1] ^ 2, 25 * x[2])
}
nonlinearIEPEquilibrium060m <-
  solnp(
    c(1, 1),
    complements,
    ineqfun = bc,
    ineqLB = 0, 
    ineqUB = m4[1],
    LB = c(0,0), 
    UB = c(Inf, Inf)
  )
nonlinearIEPEquilibrium080m <-
  solnp(
    c(1, 1),
    complements,
    ineqfun = bc,
    ineqLB = 0, 
    ineqUB = m4[2],
    LB = c(0,0), 
    UB = c(Inf, Inf)
  )
nonlinearIEPEquilibrium100m <-
  solnp(
    c(1, 1),
    complements,
    ineqfun = bc,
    ineqLB = 0, 
    ineqUB = m4[3],
    LB = c(0,0), 
    UB = c(Inf, Inf)
  )
nonlinearIEPEquilibrium120m <-
  solnp(
    c(1, 1),
    complements,
    ineqfun = bc,
    ineqLB = 0, 
    ineqUB = m4[4],
    LB = c(0,0), 
    UB = c(Inf, Inf)
  )
par(mfrow = c(1, 3))
plot(
  x = rep(nonlinearIEPEquilibrium060m$pars[1], 10),
  y = seq(nonlinearIEPEquilibrium060m$pars[2], 120, length.out = 10),
  type = 'l',
  col = 'dark green',
  xlim = c(0, m4[4] / p1),
  ylim = c(0, m4[4] / p2),
  xlab = expression(x[1]),
  ylab = expression(x[2])
)
segments(
  x0 = nonlinearIEPEquilibrium060m$pars[1],
  x1 = 120,
  y0 = nonlinearIEPEquilibrium060m$pars[2],
  y1 = nonlinearIEPEquilibrium060m$pars[2],
  col = 'dark green'
)
segments(
  x0 = nonlinearIEPEquilibrium080m$pars[1],
  x1 = nonlinearIEPEquilibrium080m$pars[1],
  y0 = nonlinearIEPEquilibrium080m$pars[2],
  y1 = 120,
  col = 'dark green'
)
segments(
  x0 = nonlinearIEPEquilibrium080m$pars[1],
  x1 = 120,
  y0 = nonlinearIEPEquilibrium080m$pars[2],
  y1 = nonlinearIEPEquilibrium080m$pars[2],
  col = 'dark green'
)
segments(
  x0 = nonlinearIEPEquilibrium100m$pars[1],
  x1 = nonlinearIEPEquilibrium100m$pars[1],
  y0 = nonlinearIEPEquilibrium100m$pars[2],
  y1 = 120,
  col = 'dark green'
)
segments(
  x0 = nonlinearIEPEquilibrium100m$pars[1],
  x1 = 120,
  y0 = nonlinearIEPEquilibrium100m$pars[2],
  y1 = nonlinearIEPEquilibrium100m$pars[2],
  col = 'dark green'
)
segments(
  x0 = nonlinearIEPEquilibrium120m$pars[1],
  x1 = nonlinearIEPEquilibrium120m$pars[1],
  y0 = nonlinearIEPEquilibrium120m$pars[2],
  y1 = 120,
  col = 'dark green'
)
segments(
  x0 = nonlinearIEPEquilibrium120m$pars[1],
  x1 = 120,
  y0 = nonlinearIEPEquilibrium120m$pars[2],
  y1 = nonlinearIEPEquilibrium120m$pars[2],
  col = 'dark green'
)
abline(100 / p2,-p1 / p2, col = 'red')
curve(x ^ 2 / 25, add = TRUE)
curve(
  p2 * x ^ 2 / 25 + p1 * x,
  xlim = c(0, nonlinearIEPEquilibrium120m$pars[1]),
  ylim = c(0, m4[4]),
  xlab = expression(x[1]),
  ylab = 'Income'
)
curve(
  p2 * x + 5 * p1 * sqrt(x),
  xlim = c(0, nonlinearIEPEquilibrium120m$pars[2]),
  ylim = c(0, m4[4]),
  xlab = expression(x[2]),
  ylab = 'Income'
)
par(mfrow = c(1,1))
```

```{r nonlinearIEP-check}
grade_code()
```

## Demand Curves

### Plotting Demand by Connecting the Dots

Plot the three points you derived in the previous exercise for the
prices of good 1 equal to 1.0, 1.5, and 2.0. (Orient quantity on the 
x-axis ranging from 0 to 60 and price
on the y-axis ranging from 0 to 2.5)

```{r demandPoints, exercise = TRUE, exercies.reveal_solution = TRUE}

```

```{r demandPoints-solution}
plot(
  initialEquilibrium$pars[1],
  p1,
  type = 'p',
  xlim = c(0, 60),
  ylim = c(0, 2.5),
  xlab = "Quantity of Good 1",
  ylab = "Price of Good 1"
)
points(lowPEquilibrium$pars[1], p1 - dp1)
points(highPEquilibrium$pars[1], p1 + dp1)
```

```{r demandPoints-check}
grade_code()
```

Connect the dots with line segments. 

```{r demandSegments, exercise = TRUE, exercies.reveal_solution = TRUE}
plot(
  initialEquilibrium$pars[1],
  p1,
  type = 'p',
  xlim = c(0, 60),
  ylim = c(0, 2.5),
  xlab = "Quantity of Good 1",
  ylab = "Price of Good 1"
)
points(lowPEquilibrium$pars[1], p1 - dp1)
points(highPEquilibrium$pars[1], p1 + dp1)

```

```{r demandSegments-solution}
plot(
  initialEquilibrium$pars[1],
  p1,
  type = 'p',
  xlim = c(0, 60),
  ylim = c(0, 2.5),
  xlab = "Quantity of Good 1",
  ylab = "Price of Good 1"
)
points(lowPEquilibrium$pars[1], p1 - dp1)
points(highPEquilibrium$pars[1], p1 + dp1)
segments(
  x0 = initialEquilibrium$pars[1],
  y0 = p1,
  x1 = lowPEquilibrium$pars[1],
  y1 = p1 - dp1
)
segments(
  x0 = initialEquilibrium$pars[1],
  y0 = p1,
  x1 = highPEquilibrium$pars[1],
  y1 = p1 + dp1
)
```

```{r demandSegments-check}
grade_code()
```

### Connecting Many Dots on a Demand Curve

The more points we do this for, the smoother and more complete our
demand curve becomes. Run the code below to derive the demand schedule
for $x_1$ using the parameter values in our original example (m = 100,
p2 = 1, $\alpha$ = 0.25) for the prices ranging from 0.5 to 2.5 in
increments of 0.1. 

1. Define `p1seq` as the sequence of prices and `c1` as a `NULL` object. 
2. Create a `for` loop to iterate the constrained optimization function over all of the values in the length of `p1` using the following steps:
a. Defines the initial values of the choice variables, `pars` to be equal to `c(1,1)` if `i == 1` and otherwise equal to the value of the solution from the previous iteration (which we call `c` below); 
b. Define `c` as the value of the `pars` result from the current iteration solution. 
c. Define the $i^th$ element of `c1` as the first element of the `pars` variable (the optimal level of consumption for that price) of each solution.
3. Plot the demand curve of p1 (y-axis) versus c1 (x-axis).

```{r demandCurve, exercise = TRUE, exercise.reveal_solution = TRUE}
p1seq <- seq(0.5, 2.5, 0.1)
c1 <- NULL
for (i in 1:length(p1seq)) {
  if(i == 1) {
    pars <- c(1,1)
  } else {
    pars <- c
  }
  c <- solnp(
    pars = pars,
    ux,
    ineqfun = function(x) p1seq[i] * x[1] + p2 * x[2],
    ineqLB = 0,
    ineqUB = m,
    LB = c(0, 0),
    UB = c(Inf, Inf)
  )$pars
  c1[i] <- c[1]
}
plot(c1,
     p1seq,
     type = 'l',
     xlim = c(0, 50),
     ylim = c(0, 2.5),
     xlab = 'Quantity of Good 1',
     ylab = 'Price of Good 1')
```

```{r demandCurve-solution}
p1seq <- seq(0.5, 2.5, 0.1)
c1 <- NULL
for (i in 1:length(p1seq)) {
  if(i == 1) {
    pars <- c(1,1)
  } else {
    pars <- c
  }
  c <- solnp(
    pars = pars,
    ux,
    ineqfun = function(x) p1seq[i] * x[1] + p2 * x[2],
    ineqLB = 0,
    ineqUB = m,
    LB = c(0, 0),
    UB = c(Inf, Inf)
  )$pars
  c1[i] <- c[1]
}
plot(c1,
     p1seq,
     type = 'l',
     xlim = c(0, 50),
     ylim = c(0, 2.5),
     xlab = 'Quantity of Good 1',
     ylab = 'Price of Good 1')
```

```{r demandCurve-check}
grade_code()
```

### Plotting a Demand Function

Recall that the analytical solution for the demand curve was:

$$x_1(p_1, m) = `r tex(D[1])`$$

1.  Plot this function. Note that this function expresses *quantity* as
    a function of *price*, which implies that the axes are reversed.
2.  Rearrange the demand function to get *price* as a function of
    *quantity* in order to plot the *inverse* demand function (with
    price on the y-axis as usual).

```{r demandFunction, exercise = TRUE, exercise.reveal_solution = TRUE}

```

```{r demandFunction-solution}
curve(
  a * m / x,
  from = 0,
  to = 2.5,
  ylim = c(0, 50),
  xlab = "Price of Good 1",
  ylab = "Quantity of Good 1"
)
curve(
  a * m / x,
  from = 0,
  to = 50,
  ylim = c(0, 2.5),
  xlab = "Quantity of Good 1",
  ylab = "Price of Good 1"
)
```

```{r demandFunction-check}
grade_code("Samesies!")
```

Why do you think Alfred Marshall would graph demand functions
"backwards?"

## Income and Substitution Effects

The *substitution effect* arises because the change in the price of one good encourages a consumer to buy more of the relatively cheaper good (and less of the relatively more expensive good). <br> 
Analytically, the substitution effect is the difference between (1) the quantity that minimizes the consumer's expenditures *subject to leaving the the consumer at least as well-off as the original equilibrium* ($\tilde{x}$); and (2) the original quantity consumed ($x^*$).

The *income effect* arises because when prices change, the same amount of money income can now buy a greater or lesser amount of goods and services, overall. <br> 
Analytically, the income effect is the difference between (1) $\tilde{x}$; and (2) the new quantity consumed ($x^**$)

### Finding $\tilde{x}$. 

To separate the income and substitution, the first order of business is finding $\tilde{x}$. Run the following code to use the `Rsolnp::solnp()` function to solve the *nonlinearly*-constrained optimization problem for finding $\tilde{x}$, 
$$min_x \bigg\{p_1x_1 + p_2x_2 \bigg\}$$
$$\text{s.t. } u_0^* - u(x_1, x_2) \le 0$$

```{r xTilde, exercise = TRUE, exercise.reveal_solution = TRUE}
library(Rsolnp)
u0 <- -tail(initialEquilibrium$values, 1)
x0 <- initialEquilibrium$pars
xTilde <-
  solnp(
    pars = x0,
    fun = bcLowP,
    ineqfun = ux,
    ineqLB = -Inf,
    ineqUB = -u0,
    LB = c(0, 0),
    UB = c(Inf, Inf)
  )
xTilde
```

```{r xTilde-solution}
library(Rsolnp)
u0 <- -tail(initialEquilibrium$values, 1)
x0 <- initialEquilibrium$pars
xTilde <-
  solnp(
    pars = x0,
    fun = bcLowP,
    ineqfun = ux,
    ineqLB = -Inf,
    ineqUB = -u0,
    LB = c(0, 0),
    UB = c(Inf, Inf)
  )
xTilde
```

```{r xTilde-check}
grade_code()
```

Note: several packages implement constrained optimization with nonlinear constraints; the advantage to `Rsolnp` is that it stores the Lagrange multiplier, which is economically meaningful. 

### Graphical Decomposition

```{r, fig.height = 8, fig.width = 8}
curve(
  (-tail(initialEquilibrium$values, 1)) ^ (1 / (1 - a)) / (x ^ (a / (1 - a))),
  from = 0,
  to = m / (p1 - dp1),
  xlim = c(0, m / (p1 - dp1)),
  ylim = c(0, m / p2),
  xlab = expression(x[1]),
  ylab = expression(x[2]),
  col = 'dark green'
)
abline(m / p2,-p1 / p2, col = 'red')
points(initialEquilibrium$pars[1],
       initialEquilibrium$pars[2],
       col = 'blue',
       pch = 20)
curve(
  (-tail(lowPEquilibrium$values, 1)) ^ (1 / (1 - a)) / (x ^ (a / (1 - a))),
  add = TRUE,
  col = 'dark green',
  lty = 'dashed'
)
abline(m / p2,-(p1 - dp1) / p2, col = 'red', lty = 'dashed')
points(lowPEquilibrium$pars[1],
       lowPEquilibrium$pars[2],
       col = 'blue',
       pch = 20)
abline(tail(xTilde$values, 1) / p2,-(p1 - dp1) / p2, col = 'red', lty = 'dashed')
points(xTilde$pars[1],
       xTilde$pars[2],
       col = 'blue',
       pch = 20)
segments(x0 = initialEquilibrium$pars[1], 
         y0 = initialEquilibrium$pars[2], 
         x1 = initialEquilibrium$pars[1], 
         y1 = -5, lty = 'dashed', col = 'blue')
segments(x0 = xTilde$pars[1], 
         y0 = xTilde$pars[2], 
         x1 = xTilde$pars[1], 
         y1 = -5, lty = 'dashed', col = 'blue')
segments(x0 = lowPEquilibrium$pars[1], 
         y0 = lowPEquilibrium$pars[2], 
         x1 = lowPEquilibrium$pars[1], 
         y1 = -5, lty = 'dashed', col = 'blue')
text(x = (initialEquilibrium$pars[1]+xTilde$pars[1])/2, 
     y = 0, "Substitution Effect", adj = c(0, 0.5), srt = 90)
text(x = (lowPEquilibrium$pars[1]+xTilde$pars[1])/2, 
     y = 0, "Income Effect", adj = c(0, 0.5), srt = 90)
```

<details style="line-height:105%"><summary>Click here to see the code for graphing the decomposition of the change in price.</summary>

`curve(`<br>
`  (-tail(initialEquilibrium$values, 1)) ^ (1 / (1 - a)) / (x ^ (a / (1 - a))),`<br>
`  from = 0,`<br>
`  to = m / (p1 - dp1),`<br>
`  xlim = c(0, m / (p1 - dp1)),`<br>
`  ylim = c(0, m / p2),`<br>
`  xlab = expression(x[1]),`<br>
`  ylab = expression(x[2]),`<br>
`  col = 'dark green')`<br>
`abline(m / p2,-p1 / p2, col = 'red')`<br>
`points(initialEquilibrium$pars[1],`<br>
`       initialEquilibrium$pars[2],`<br>
`       col = 'blue', pch = 20)`<br>
`curve(`<br>
`  (-tail(lowPEquilibrium$values, 1)) ^ (1 / (1 - a)) / (x ^ (a / (1 - a))),`<br>
`  add = TRUE,`<br>
`  col = 'dark green',`<br>
`  lty = 'dashed')`<br>
`abline(m / p2,-(p1 - dp1) / p2, col = 'red', lty = 'dashed')`<br>
`points(lowPEquilibrium$pars[1],`<br>
`       lowPEquilibrium$pars[2],`<br>
`       col = 'blue',`<br>
`       pch = 20)`<br>
`abline(tail(xTilde$values, 1) / p2,-(p1 - dp1) / p2, col = 'red', lty = 'dashed')`<br>
`points(xTilde$pars[1],`<br>
`       xTilde$pars[2],`<br>
`       col = 'blue', pch = 20)`<br>
`segments(x0 = initialEquilibrium$pars[1], `<br>
`         y0 = initialEquilibrium$pars[2], `<br>
`         x1 = initialEquilibrium$pars[1], `<br>
`         y1 = -5, lty = 'dashed', col = 'blue')`<br>
`segments(x0 = xTilde$pars[1], `<br>
`         y0 = xTilde$pars[2], `<br>
`         x1 = xTilde$pars[1], `<br>
`         y1 = -5, lty = 'dashed', col = 'blue')`<br>
`segments(x0 = lowPEquilibrium$pars[1], `<br>
`         y0 = lowPEquilibrium$pars[2], `<br>
`         x1 = lowPEquilibrium$pars[1], `<br>
`         y1 = -5, lty = 'dashed', col = 'blue')`<br>
`text(x = (initialEquilibrium$pars[1]+xTilde$pars[1])/2, `<br>
`     y = 0, "Substitution Effect", adj = c(0, 0.5), srt = 90)`<br>
`text(x = (lowPEquilibrium$pars[1]+xTilde$pars[1])/2, `<br>
`     y = 0, "Income Effect", adj = c(0, 0.5), srt = 90)`<br>

</details>

### Changes in Consumer Surplus

The two dashed expenditure lines (through $x^**$ and $\tilde{x}$) capture the difference in income needed to achieve the new and original maximum levels of utility, respectively, calculated at the new price levels. In other words this is the additional income the consumer "feels" by moving from the original indifference curve to the new one. It is also one of the ways to capture the change in consumer surplus across all markets when relative prices change. 

```{r installR, echo=FALSE}
question("What is the approximate dollar-denominated welfare effect (i.e. income effect) of the price change (evaluated at the new prices) for our example?",
    answer("zero"),
    answer("-9.64"),
    answer("9.64", correct = TRUE),
    answer("90.36"),
    allow_retry = TRUE
  )
```

### Calculating Consumer Surplus

Once you've found the $\tilde{x}$, finding the consumer surplus is relatively straightforward. In our case, the expenditures needed at the new prices in order to reach the original utility is the value of the objective function for the expenditure minimization problem we solved to calculate $\tilde{x}$, `xTilde$objective`, or `r xTilde$objective`; the expenditures needed to reach the new optimal utility is simply the consumer's money income, `m`. The change in consumer surplus will be the difference `r round(m - xTilde$objective, 2)`. 

Some policy analyses use this to evaluate the welfare effects of policy changes in terms of real income. The trick is estimating (with as few assumptions as possible!) the unobserved $\tilde{x}$. 
