---
title: "3.4 Monopolistic Competition"
output: 
  learnr::tutorial: 
    progressive: true
runtime: shiny_prerendered
author: 
  name: "Created by [Jim Bang](http://www.github.com/bangecon)"
  email: BangJamesT@sau.edu
  affiliation: "[St. Ambrose University](http://www.sau.edu)"
date: "`r format(Sys.Date(), '%d %b %Y')`"
---

<style type="text/css">
h1{font-size: 24pt}
h2{font-size: 20pt}
h3{font-size: 18pt}
h4,h5,h6{font-size: 16pt}
body{font-size: 16pt}
#doc-metadata {
  margin-bottom: 10px;
  font-size: 0.66em;
}
</style>

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
gradethis_setup(exercise.reveal_solution = TRUE, exercise.eval = TRUE)
knitr::opts_chunk$set(echo = FALSE, fig.height = 6, fig.width = 6)
```

## Monopolistic Competition

As the manager of a firm in a monopolistically-competitive market, you estimate demand for your product to follow the demand function: 

$$Q_d = 36 - 4P + A$$
where $A$ is the amount of expenditure the firm makes on advertising. 

The internal (total) cost function of the firm is:

$$C(Q) = 4 + 4Q + Q^2 + A$$

Solve for the profit-maximizing levels of output, price, and advertising expenditures. 

```{r advertising, exercise = TRUE}
```

```{r advertising-setup}

```

```{r advertising-solution}

```

```{r advertising-check}

```

