---
title: "3.4 Monopolistic Competition"
output: 
  learnr::tutorial: 
    progressive: true
runtime: shiny_prerendered
author: 
  name: "Created by [Jim Bang](http://www.github.com/bangecon)"
  email: BangJamesT@sau.edu
  affiliation: "[St. Ambrose University](http://www.sau.edu)"
date: "`r format(Sys.Date(), '%d %b %Y')`"
---


```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(ggcheck)
gradethis_setup(exercise.reveal_solution = TRUE, exercise.eval = TRUE)
knitr::opts_chunk$set(echo = FALSE, fig.height = 6, fig.width = 6)
```

## Monopolistic Competition

As the manager of a firm in a monopolistically-competitive market, you estimate demand for your product to follow the demand function: 

$$Q_d = 36 - 4P + 8A^{0.5}$$
where $A$ is the amount of expenditure the firm makes on advertising. 

The internal (total) cost function of the firm is:

$$C(Q) = 4 + 4Q + Q^2 + A$$

Solve for the profit-maximizing levels of output, price, and advertising expenditures. 

```{r advertising, exercise = TRUE}

```

```{r advertising-hint}
# Since cost is a function of Q, you might want to try inverting the demand function to make it price as a function of quantity to make total revenues = P(Q,A)*Q. 
```

```{r advertising-solution, message = FALSE, warning = FALSE}
library(managerialEconWithR)
pi <- function(x) {
  9*x[1] - 0.25*x[1]^2 + 2*x[1]*x[2]^0.5 - (4 + 4*x[1] + x[1]^2 + x[2])
}
eq <- solnp2(
  pars = c(1,1), 
  fun = pi, 
  LB = c(0, 0), 
  control = list(trace = 0), 
  max = TRUE
)
(Q = eq$pars[1])
(A = eq$pars[2])
(P = 9 - 0.25*Q + 2*A^0.5 )
```

```{r advertising-check}
grade_this({
  pass_if(round(.result, 1) == 26.5) 
  fail()
})
```

