---
title: "2.4 Competitive Firms' Output Choices in the Short Run"
author: Jim Bang
output: 
  learnr::tutorial: 
    progressive: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
gradethis_setup(exercise.reveal_solution = TRUE, exercise.eval = TRUE)
knitr::opts_chunk$set(echo = FALSE, fig.height = 6, fig.width = 6)
```

<style type="text/css">
h1{font-size: 24pt}
h2{font-size: 20pt}
h3{font-size: 18pt}
h4,h5,h6{font-size: 16pt}
body{font-size: 16pt}
</style>

## The Profit Function

Perfect Competition

$$\matrix{ \matrix{\text{1. Many, many sellers (and buyers)} \\ \text{2. Homogeneous product} \\ \text{3. Perfect information}} \Biggr\} \\ \text{4. Free entry/exit} \\ \text{  }} \matrix{\text{ } \\ \implies \\ \text{ } \\ \implies \\ \text{ } }  \matrix{\Biggl\{ \matrix{\text{Firms are} \\ \text{price} \\ \text{takers}} \\ \text{Profits/losses} \\ \text{are temporary}}$$

### The Producer's Problem

1. Objective: 
2. Endogenous (choice) variable: 
3. Exogenous variable(s): 

### General Form 

$$\max_Q \{ \pi(Q) = TR(Q) - C(Q) \} $$
Since price is *exogenous* (determined by the market), $TR = PQ$:
$$\pi(Q) = PQ - C(Q)$$

### Per-Unit Form

Since $C(Q) = ATC(Q) \cdot Q$, we can rewrite profits as: 

$$\pi(Q) = [P - ATC(Q)]Q$$

## Analytical Approach

Suppose that our firm's technology leads to a cost function that can be approximated by the function:

$$C(Q) = \frac{Q^3}{20} - \frac{9Q^2}{10} + 10Q + 5$$

and that market conditions imply that $P = 10$.

Solve the producer's problem analytically using `Ryacas`. 

1. Load `Ryacas`.
2. Define the profit function. 
3. Derive $d\pi/dQ$ and simplify. 
4. Solve $d\pi/dQ = 0$ for $Q$, remove the variables, and reformat to an ordinary `R` object. 

```{r profitmaxAnalytical, exercise = TRUE}

```

```{r profitmaxAnalytical-solution}
library(Ryacas)
pi    <- ysym('7*Q - (0.05*Q^3 - 0.9*Q^2 + 10*Q + 5)')
dpi   <- deriv(pi, 'Q') |>
  simplify()
qStar <- solve(dpi, 'Q')[2] |>
  y_rmvars() |>
  as_r()
```

```{r profitmaxAnalytical-check}
grade_this(
  if(round(.result) != 10) {
    fail("The solution should be negligibly close to 10.")} 
  pass()
)
```

```{r}
library(Ryacas)
pi    <- ysym('7*Q - (0.05*Q^3 - 0.9*Q^2 + 10*Q + 5)')
dpi   <- deriv(pi, 'Q') |>
  simplify()
qStar <- solve(dpi, 'Q')[2] |>
  y_rmvars() |>
  as_r()
```

The first-order condition for the profit-maximization problem is: 

$$`r Ryacas::tex(dpi)` = 0$$

## Graphical Approach

Run the following code to construct the following graphs: 

1. Total revenue and total cost versus quantity (top left);
2. Total profit versus quantity (top right);
3. Marginal revenue, marginal cost, average cost, and average variable cost versus quantity (bottom left); 
4. Marginal (change in) profit versus quantity (bottom right); 

```{r profitmaxGraph, exercise = TRUE}
par(mfrow = c(2,2))                                                   # Set a 2-by-2 plot window.
curve(
  0.05 * x ^ 3 - 0.9 * x ^ 2 + 10 * x + 5,                            # Total cost curve
  col = 'red',         # Line color
  from = 0,            # Starting x value
  to = 15,             # Ending x value
  xlab = "Quantity",   # X-axis label
  ylab = "$"           # Y-axis label
)                                           
curve(7 * x, col = 'blue', add = TRUE)                                # Total revenue curve
curve(7 * x - 5, lty = 'dashed', add = TRUE)                          # MR = MC reference line
curve(
  7 * x - (0.05 * x ^ 3 - 0.9 * x ^ 2 + 10 * x + 5),                  # Total profit curve
  col = 'black',
  from = 0,            # Starting x value
  to = 15,             # Ending x value
  xlab = "Quantity",   # X-axis label
  ylab = "$",          # Y-axis label
  ylim = c(-10, 10)    # Y-axis range
)
lines(x = c(0, 15),                                                   # Zero-profit reference line
      y = c(0, 0),
      lty = 'dashed')  # Line type
curve(
  0.15 * x ^ 2 - 1.8 * x + 10,                                        # Marginal cost curve
  col = 'red',         # Line color
  from = 0,            # Starting x value
  to = 15,             # Ending x value
  xlab = "Quantity",   # X-axis label
  ylab = "$",          # Y-axis label
  ylim = c(0, 12.5)    # Y-axis range
)
curve(0.05 * x ^ 2 - 0.9 * x + 10 + 5 / x, col = 'brown', add = TRUE) # Average total cost curve
curve(0.05 * x ^ 2 - 0.9 * x + 10,                                    # Average variable cost curve
      col = 'brown',   # Line color
      add = TRUE,
      lty = 'dashed')  # Line type
lines(x = c(0, 15),                                                   # Average/marginal revenue curve (price)
      y = c(7, 7), 
      col = 'blue')
lines(x = c(0, 15),                                                   # Shutdown reference line
      y = c(5.95, 5.95),
      lty = 'dashed')
curve(
  7 - (0.15 * x ^ 2 - 1.8 * x + 10),                                  # Marginal profit curve
  col = 'black',
  from = 0,            # Starting x value
  to = 15,             # Ending x value
  xlab = "Quantity",   # X-axis label
  ylab = "$",          # Y-axis label
  ylim = c(-10, 10)    # Y-axis range
)
lines(x = c(0, 15),                                                   # d(Profit) = 0 reference line
      y = c(0, 0),
      lty = 'dashed')  # Line type
```

```{r profitmaxGraph-solution}
par(mfrow = c(2,2))                                                   # Set a 2-by-2 plot window.
curve(
  0.05 * x ^ 3 - 0.9 * x ^ 2 + 10 * x + 5,                            # Total cost curve
  col = 'red',         # Line color
  from = 0,            # Starting x value
  to = 15,             # Ending x value
  xlab = "Quantity",   # X-axis label
  ylab = "$"           # Y-axis label
)                                           
curve(7 * x, col = 'blue', add = TRUE)                                # Total revenue curve
curve(
  7 * x - (0.05 * x ^ 3 - 0.9 * x ^ 2 + 10 * x + 5),                  # Total profit curve
  col = 'black',
  from = 0,            # Starting x value
  to = 15,             # Ending x value
  xlab = "Quantity",   # X-axis label
  ylab = "$",          # Y-axis label
  ylim = c(-10, 10)    # Y-axis range
)
lines(x = c(0, 15),                                                   # Zero-profit reference line
      y = c(0, 0),
      lty = 'dashed')  # Line type
curve(
  0.15 * x ^ 2 - 1.8 * x + 10,                                        # Marginal cost curve
  col = 'red',         # Line color
  from = 0,            # Starting x value
  to = 15,             # Ending x value
  xlab = "Quantity",   # X-axis label
  ylab = "$",          # Y-axis label
  ylim = c(0, 12.5)    # Y-axis range
)
curve(0.05 * x ^ 2 - 0.9 * x + 10 + 5 / x, col = 'brown', add = TRUE) # Average total cost curve
curve(0.05 * x ^ 2 - 0.9 * x + 10,                                    # Average variable cost curve
      col = 'brown',   # Line color
      add = TRUE,
      lty = 'dashed')  # Line type
lines(x = c(0, 15),                                                   # Average/marginal revenue curve (price)
      y = c(7, 7), col = 'blue', add = TRUE)
lines(x = c(0, 15),                                                   # Shutdown reference line
      y = c(5.95, 5.95),
      lty = 'dashed', add = TRUE)
curve(
  7 - (0.15 * x ^ 2 - 1.8 * x + 10),                                  # Marginal profit curve
  col = 'black',
  from = 0,            # Starting x value
  to = 15,             # Ending x value
  xlab = "Quantity",   # X-axis label
  ylab = "$",          # Y-axis label
  ylim = c(-10, 10)    # Y-axis range
)
lines(x = c(0, 15),                                                   # d(Profit) = 0 reference line
      y = c(0, 0),
      lty = 'dashed')  # Line type
```

```{r profitmaxGraph-check}
grade_code()
```

## Computational Approach

Find the solution to the profit-maximization problem numerically. 

Careful! Since our profit function has multiple "critical values" (points where change = 0), getting the correct solution will be sensitive to the initial value (`pars` option)

```{r profitmaxNumerical, exercise = TRUE}

```

```{r profitmaxNumerical-solution}
library(Rsolnp)
loss <- function(x) {
  - 7*x + (0.05*x^3 - 0.9*x^2 + 10*x + 5)
}
solution <- solnp(
  pars = 1,
  fun = loss,  # Solnp calculates minima so we want to minimize loss 
  LB = 0
)
```

```{r profitmaxNumerical-check}
grade_this(
  if(round(.result$pars) != 10) {
    fail("The solution should be negligibly close to 10.")} 
  pass()
)
```

