---
title: "2.2 Cost Minimization"
output: 
  learnr::tutorial: 
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
author: 
  name: "Created by [Jim Bang](http://www.github.com/bangecon)"
  email: BangJamesT@sau.edu
  affiliation: "[St. Ambrose University](http://www.sau.edu)"
date: "`r format(Sys.Date(), '%d %b %Y')`"
---

<style type="text/css">
h1{font-size: 24pt}
h2{font-size: 20pt}
h3{font-size: 18pt}
h4,h5,h6{font-size: 16pt}
body{font-size: 16pt}
#doc-metadata {
  margin-bottom: 10px;
  font-size: 0.66em;
}
</style>

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
gradethis_setup(exercise.reveal_solution = TRUE, exercise.eval = TRUE)
knitr::opts_chunk$set(echo = FALSE, fig.height = 6, fig.width = 6)
```

## Review: The Firm's Three Problems:

1. <span style="color: red;">Cost-Minimizing Combination of Inputs: For a *given* level of (expected) output (firm demand) choose the *combination* of inputs that minimizes the cost of producing that amount to derive the cost function.</span>  
2. Profit-Maximizing Level of Output: Choose the *level* of output that maximizes profits by adjusting quantity and/or price of the output.  
3. Long-Run Readjustment: Adjust the level of capital to minimize costs at the given price and quantity in the market. This might happen at the firm level (adjusting firm capacity) or at the industry level (entry/exit of firms).  

Decision sequence:\
1. Empirical Market Research: Determine the extent of your market ($\hat{Q}$).\
2. Capital Budgeting: For $\hat{Q}$ and market input prices ($w$ and $r_K$), determine the cost-minimizing combination of inputs ($\hat{L}$ and $\bar{K}$). *This fixes your choice of capital for subsequent short run decisions.*\
3. Adjust *variable* inputs ($L$) to increase or decrease output ($Q$) in the short run in response to temporary fluctuations in demand.\
4. Periodically evaluate the capital budget.

Flexibility of input choices:  \
1. Higher (relative) wages: Use relatively more capital to produce $\hat{Q}$.\
2. Lower (relative) wages: Use relatively more labor to produce $\hat{Q}$.

### Cost-Minimization Overview

Goal: Minimize costs for a (each?) given quantity 

Constraint: Quantity produced with available technology

Endogenous (Strategic) Variables: Choice of inputs

Exogenous Variables (Parameters): Market input prices, technology

### Isocost Lines

Isocost lines resemble budget lines from the consumer problem. 

$$TC = wL + rK$$

Since total cost is the *objective* the firm sets out to minimize, we do not set a given value for the left-hand side. 

The right-hand side consists of labor costs (price per unit labor, i.e. wage, times the quantity of labor hired) plus capital costs (price per unit capital, i.e. rent, times the quantity of capital rented). 

```{r costExpr, echo=FALSE}
question("What is the syntax for the `expr` argument to plot an isocost curve with the `curve()` function, where the wage is 10, rental price of capital is 20, and total cost is 1000?",
    answer("`10*L + 20*K`"),
    answer("`1000 - 10*x - 20*y`"),
    answer("`50 - 0.5*x`", correct = TRUE),
    answer("`50 - 2*L`"),
    allow_retry = TRUE
  )
```

### Plotting Isocost Lines

Solving for $K$, the isocost line becomes

$$K = \frac{TC}{r} - \frac{w}{r}L.$$

Plot the isocost lines for $TC = \{500, 750, 1000\}$ for $0 \le L \le 100$ given a market wage of 10 and a market rental price of capital of 20. Orient the quantity of labor on the horizontal axis (labeled "Labor") and quantity of capital on the vertical axis (labeled "Capital").

```{r isocost, exercise = TRUE}

```

```{r isocost-solution}
library(ggplot2)
ggplot() + 
  geom_function(fun = function(x) (1000 - 10*x) / 20, aes(color = '1000')) + 
  geom_function(fun = function(x) (750 - 10*x) / 20, aes(color = '750')) + 
  geom_function(fun = function(x) (500 - 10*x) / 20, aes(color = '500')) + 
  lims(x = c(0, 100), y = c(0,50)) + 
  labs(x = "Labor", y = "Capital", color = "Total Cost") + 
  scale_colour_manual(breaks = c("500", "750", "1000"), values = c("red", 'blue', "purple")) 
```

```{r isocost-check}
grade_this({
  if (ith_geom_is(.result, "function", i = 1) & ith_geom_is(.result, "function", i = 2)) {
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

### Isoquant *Constraint*

Suppose that the firm wants to produce 2,500 units of output. Add the isoquant for the Cobb-Douglas production function from the previous section to the graph you started above. 

$$f(L, K) = 100L^{1/3}K^{2/3} = 2500$$

```{r quantExpr, echo=FALSE}
question("Which of the following is NOT a correct syntax for the `expr` argument to plot an isoquant curve with the `curve()` function, where production is described by a Cobb-Douglass production function with labor-share of one-third, a total factor productivity of 100, and total output of 2,500?",
    answer("100*x^(`1/3`) * y^(`2/3`)", correct = TRUE),
    answer("125/(x^0.5)"),
    answer("(2500/x^(`1/3`))^(3/2)"),
    answer("25^(3/2) * x^(-3/2)"),
    allow_retry = TRUE
  )
```

### Plotting Isoquants

Add an isoquant for $Q = 2500 to the previous isocost graph (stored as `g` in the setup). 

```{r isoquant, exercise = TRUE, warning = FALSE, message = FALSE}

```

```{r isoquant-setup, warning = FALSE, message = FALSE}
library(ggplot2)
g <- ggplot() +
  geom_function(
    fun = function(x)
      (1000 - 10 * x) / 20,
    aes(color = 'Cost = 1000')
  ) +
  geom_function(
    fun = function(x)
      (750 - 10 * x) / 20,
    aes(color = 'Cost = 750')
  ) +
  geom_function(
    fun = function(x)
      (500 - 10 * x) / 20,
    aes(color = 'Cost = 500')
  ) +
  lims(x = c(0, 100), y = c(0, 50)) +
  labs(x = "Labor", y = "Capital", color = NULL) +
  scale_colour_manual(
    breaks = c("Cost = 500", "Cost = 750", "Cost = 1000"),
    values = c("red", 'blue', "purple")
  ) 
```

```{r isoquant-solution, warning = FALSE, message = FALSE}
g + geom_function(
  fun = function(x)
    (25 * x ^ (-1 / 3)) ^ (3 / 2),
  aes(color = 'Output = 2500')
) +
  scale_colour_manual(
    breaks = c("Cost = 500", "Cost = 750", "Cost = 1000", "Output = 2500"),
    values = c("red", 'blue', "purple", "black")
  ) 
```

```{r isoquant-check}
grade_this({
  if (ith_geom_is(.result, "function", i = 1) & ith_geom_is(.result, "function", i = 2)) {
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

## Graphical Approach

From our graphs, it appears that this firm would minimize its cost of producing 2500 units by hiring about 25 units of labor and renting about 25 units of capital at a cost of 750.  

Let's see if we can verify this computationally (and analytically in the appendix!). 

## Computational Approach

### Expenditure Minimization Redux

Suppose that we wanted to find the expenditure-minimizing way to achieve 2,500 "utils" of utility for a consumer whose preferences follow a Cobb-Douglas utility function and prices of $x_1$ and $x_2$ equal to 20 and 10, respectively. 

$$U(x) = 100x_1^{1/3}x_2^{2/3}$$

$$e(x) = 20x_1 + 10x_2$$

What made the computational approach to expenditure minimization tricky is the fact that the constraint is the value of the *nonlinear* utility function, while the objective is the *linear* expenditure function. The code we used to solve the problem was: 

```{r xTilde, echo = TRUE}
library(MicroWithR)
A = 100
a = 1/3
p1 = 10
p2 = 20
ux <- function(x) {
  A * x[1] ^ (a) * x[2] ^ (1 - a)
}
bc <- function(x) {
  p1*x[1] + p2*x[2]
}
xTilde <- solnp2(
  pars = c(1,1),
  fun = bc,
  ineqfun = ux,
  ineqLB = 2500,
  ineqUB = Inf,
  LB = c(0, 0),
  UB = c(Inf, Inf), 
  control = list(trace = 0)
)
```

### Cost Minimization 

Find the cost-minimizing level of labor and capital at a wage of 10 and rental price of capital of 20 for a firm using a Cobb-Douglass production technology with a total factor productivity of 100 and a labor share of one-third that expects to sell 2500 units.  

```{r costMin, exercise = TRUE}


```

```{r costMin-solution}
A = 100
a = 1/3
p1 = 10
p2 = 20
qx <- function(x) {
  A * x[1] ^ (a) * x[2] ^ (1 - a)
}
tc <- function(x) {
  p1*x[1] + p2*x[2]
}
costmin <- solnp2(
  pars = c(1,1),
  fun = tc,
  ineqfun = qx,
  ineqLB = 2500,
  ineqUB = Inf,
  LB = c(0, 0),
  UB = c(Inf, Inf), 
  control = list(trace = 0)
)
```

```{r costMin-check}
grade_this({if (!inherits(.result, c("list"))) {
    fail("Your class of your answer should be a list")}
  if (.result$pars != c(25, 25)) { 
    fail("The values of L and K should equal (25, 25).") }
  pass()
})
```

Wow! Our graphical eyeball test was right! For $Q = 2,500$ the cost-minimizing combination is 25 units of labor and 25 units of capital at a cost of 750. 

```{r lambda, echo=FALSE}
question("Finally, what does the lagrange multiplier represent here? ",
    answer("The additional output produced by hiring one more unit of labor and capital."),
    answer("The additional cost of hiring one more unit of labor and capital."),
    answer("The additional cost of producing one more unit of output.", correct = TRUE),
    answer("The additional output produced by increasing costs by one unit"),
    allow_retry = TRUE
  )
```

## Summary

1. A firm seeks to produce any given quantity with a combination of inputs that minimizes its costs. 
2. The combination of inputs that minimizes per-unit costs *for a given quantity of output* represents a point on the long-run average cost curve. 
3. Choosing (or anticipating) the right *quantity* of output consumers will buy at a given price determines whether the firm will maximize its profits. 

## Appendix: Analytical Approach

### Consumer Expenditure Minimization Redux

The solution to the cost-minimization problem mirrors the expenditure minimization problem we used to help decompose the effect of a price change into income and substitution effects. In that context, we defined $\tilde{x}$ to be the consumption bundle that minimized expenditures subject to remaining on the same level of utility as the original equilibrium before the price change. 

$$min_x \bigg\{p_1x_1 + p_2x_2 \bigg\}$$

$$\text{s.t. } u_0^* - u(x_1, x_2) \le 0$$

### Cost Minimization - Setup

In the case of cost minimization, the basic structure is similar, except that the firm uses its "expenditures" (costs) to "buy" (hire/rent) inputs to gain a benefit in the form of output. Formally, the firm solves: 

$$min_{L,K} \bigg\{wL + rK \bigg\}$$

$$\text{s.t. } y^* - f(L, K) \le 0$$

Recall the five key steps for constrained optimization problems: 

1. Rearrange the constraint to be something equal to *zero* when it binds. 
2. Define the Lagrangean (since this is minimization the constraint portion of the Lagrangean will be *less than* or equal to zero).
3. Take derivatives with respect to the choice variables (including $\lambda$). 
4. Set the derivatives equal to zero. 
5. Solve for all choice variables. 

The Lagrangean for the firm with parameters defined from our example is: 

$$\mathcal{L} = 10L + 20K + \lambda(2500 - 100L^{1/3}K^{2/3})$$

The code chunk below obtains the first-order conditions for the cost minimization problem. 

```{r foc, echo = TRUE}
library(Ryacas)
library(Deriv)
w = 10
r = 20
G <- ysym('10*L + 20*K + lambda*(2500 - 100*L^(1/3) * K^(2/3))')
dG <- deriv(G, c('L', 'K', 'lambda'))
dG
```

Setting the derivatives of the Lagrangean equal to zero finds the "first-order conditions." These conditions find the "critical values" where the Lagrangean switches from increasing to decreasing (or vice-versa) corresponding to maximum (minimum) values. 

$$\pmatrix{\frac{\partial\mathcal{L}}{\partial L} \\ \frac{\partial\mathcal{L}}{\partial K} \\ \frac{\partial\mathcal{L}}{\partial \lambda}} = 
\pmatrix{`r tex(dG[[1]])` \\ `r tex(dG[[2]])` \\`r tex(dG[[3]])`} = \pmatrix{0 \\ 0 \\ 0} $$

### The Marginal Rate of Technical Substitution (MRTS)

Combining the first-order conditions with respect to $x_1$ and $x_2$, we get the "Marginal Resource Cost per Marginal Product" condition: 

$$\frac{w}{MP_L} = \frac{r}{MP_K} = \lambda \text{, or}$$

We can rewrite this condition as the ratio of the marginal products equalling the ratio of the marginal resource costs. 

$$MRTS_{LK} = \frac{MP_L}{MP_K} = \frac{w}{r}.$$

Rewriting it this way gives the tangency condition we saw in the graphical approach: The slope of the isoquant is the ratio of the marginal products, while the slope of the isocost line is the ratio of the resource prices; when these are equal, the firm minimizes the cost of producing the level of output described by the isoquant. 

The code chunk below derives the $MRTS$ of the example: 

```{r MRTS, echo = TRUE}
MPLinv <- solve(dG[[1]], 'lambda') |>
  y_rmvars()/w 
MPL <- yac_str(1/MPLinv) |>
  Simplify() |>
  ysym()
MPKinv <- solve(dG[[2]], 'lambda') |>
  y_rmvars()/r
MPK <- yac_str(1/MPKinv) |>
  Simplify() |>
  ysym()
MRTS <- yac_str(MPL/MPK) |> 
  Simplify() |> 
  ysym()
MRTS
```

$$`r tex(MRTS)` = w/r$$

### Solving for Input Levels

The following chunc finishes the solution. 

```{r solution, echo = TRUE}
Lstar <- with_value(dG[[3]], 'K',  simplify(y_rmvars(solve(MRTS, w/r, 'K')))) |>
  solve('L') |>
  y_rmvars() |> 
  as_r()
Kstar <- with_value(solve(MRTS, w/r, 'K'), 'L', Lstar) |> 
  y_rmvars() |> 
  as_r()
lambda <- with_value(dG[[1]], 'K', as_r(y_rmvars(Kstar))) |>
  with_value('L', as_r(y_rmvars(Lstar))) |>
  solve('lambda') |>
  y_rmvars() |>
  as_r()
```

$$\pmatrix{L^* \\ K^* \\ \lambda} = \pmatrix{`r Lstar` \\ `r Kstar` \\`r lambda`}$$

What does $\lambda$ *represent*?
