---
title: "3.5 Oligopoly: Sweezy and Bertrand Oligopoly"
output: 
  learnr::tutorial: 
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
author: 
  name: "Created by [Jim Bang](http://www.github.com/bangecon)"
  email: BangJamesT@sau.edu
  affiliation: "[St. Ambrose University](http://www.sau.edu)"
date: "`r format(Sys.Date(), '%d %b %Y')`"
---

<style type="text/css">
h1{font-size: 24pt}
h2{font-size: 20pt}
h3{font-size: 18pt}
h4,h5,h6{font-size: 16pt}
body{font-size: 16pt}
#doc-metadata {
  margin-bottom: 10px;
  font-size: 0.66em;
}
</style>

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
gradethis_setup(exercise.reveal_solution = TRUE, exercise.eval = TRUE)
knitr::opts_chunk$set(echo = FALSE, fig.height = 6, fig.width = 6, message = FALSE)
```

## Oligopoly

### Characteristics of Oligopoly

1.  Few Large Producers (often one dominant firm)
2.  Products may be homogeneous or differentiated
3.  Mutual Interdependence (Some Control over Price)
4.  Barriers to Entry

### Basic Types of Oligopoly

|             | Sweezy    | Bertrand     | Cournot      | Stackelberg |
|-------------|-----------|--------------|--------------|-------------|
| Firms       | Few       | Few          | Few?         | 2           |
| Strategy    | Price?    | Price        | Quantity     | Price       |
| Products    | Different | Same         | Same         | Same        |
| Timing      | Unclear   | Simultaneous | Simultaneous | Sequential  |
| Information | Unclear   | Perfect      | Perfect      | Perfect     |
| Rationality | Unclear   | Full         | Full         | Full        |

## Stackelberg (Kinked Demand Curve) Oligopoly



## Review: A Simple Monopoly Example

Market (inverse) demand:

$$P = 10 - Q$$

Firm Costs: 

$$TC = 6Q$$

Solve for the monopoly equilibrium price and quantity. 

Suppose that the firms entered a legally-binding trust that enabled them to collude. What would the quantity of each individual firm be? 

### Solution

If you chose to try doing this using R (which you might not need to since demand, marginal revenue, and marginal cost are all linear), the code might look like this:

```{r, echo=TRUE}
library(MicroWithR)
Dinv = function(x) {
  10 - x
}
pi <- function(x) {
  Dinv(x)*x - 6*x
}
eq0 <- solnp2(
  pars = 1, 
  fun = pi, 
  LB = 0, 
  control = list(trace = 0), 
  max = TRUE
) 
p0 = Dinv(eq0$pars)
qi = eq0$pars/2
```

The total monopolist's quantity would be: 

$$(Q,P) = (`r eq0$pars`, `r p0`)$$

This makes each of the two firms' share of the total quantity: 

$$(Q_1, Q_2) = (`r eq0$pars/2`, `r eq0$pars/2`)$$ 

## Bertrand Duopoly

A Bertrand oligopoly is a "winner take all" market in which the firm that charges the lowest price "wins" the opportunity to supply the entire market. If there is a tie, then firms charging the lowest price split the quantity equally. 

$$Q_1 = D(P_1,P_2) = \Biggl\{ \matrix{0 \text{ if } P_1 > P_2 \\ Q/2 \text{ if } P_1 = P_2 \\ Q \text{ if } P_1 < P_2}$$

$$Q_2 = D(P_1,P_2) = \Biggl\{ \matrix{0 \text{ if } P_2 > P_1 \\ Q/2 \text{ if } P_2 = P_1 \\ Q \text{ if } P_2 < P_1}$$

### "Analytical" Solution

Four Non-Cases: 

1. $p_1 \ne p_2$ and $min\{p_1, p_2\} < MC$
2. $p_1 = p_2$ and $min\{p_1, p_2\} < MC$
3. $p_1 \ne p_2$ and $min\{p_1, p_2\} > MC$
4. $p_1 = p_2$ and $min\{p_1, p_2\} > MC$

Suppose two identical firms choose their price to maximize profits subject to a "winner take all" market structure. What will the profit-maximizing price of the winning firm(s) be? 

### Simulation of Bertrand Duopoly

Use the following link to choose your strategy under the demand and cost conditions for the example. 

[Bertrand Duopoloy Activity](https://forms.gle/kRFtVWAQgcCTZbDp6)

<iframe src="https://docs.google.com/forms/d/e/1FAIpQLSe8Z5TTc29MeRST2_zhCmpVedXVmNQnoItNg8Mn38jKvzzoWA/viewform?embedded=true" width="640" height="1123" frameborder="0" marginheight="0" marginwidth="0">Loading…</iframe>

