---
title: "1.3 Utility Functions - Practice"
output: 
  learnr::tutorial: 
    progressive: true
runtime: shiny_prerendered
author: 
  name: "Created by [Jim Bang](http://www.github.com/bangecon)"
  email: BangJamesT@sau.edu
  affiliation: "[St. Ambrose University](http://www.sau.edu)"
date: "`r format(Sys.Date(), '%d %b %Y')`"
---

<style type="text/css">
#doc-metadata {
  margin-bottom: 10px;
  font-size: 0.8em;
}
</style>

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(ggcheck)
library(Deriv)
library(Ryacas)
gradethis_setup()
tutorial_options(exercise.eval = TRUE, exercise.reveal_solution = FALSE)
knitr::opts_chunk$set(echo = FALSE, fig.align = "center", fig.height = 6, fig.width = 6)
knitr::include_graphics
```

## Constant Elasticity of Substitution

Suppose that we can describe a consumer's preferences between two goods using a constant elasticity of substitution utility function with $r = 1/2$: 

$$U(x) = (x_1^{1/2} + x_2^{1/2})^2, $$ 

so that the elasticity of substitution is $\sigma = \frac{1}{1-r} = 2$. 

1. Graph the utility function in 3 dimensions. 

```{r ces, exercise=TRUE, exercise.reveal_solution = TRUE}

```

```{r ces-hint}
# Use seq() define $x$ and $y$ as sequences; and define $z$ as the `outer()` product of $x^{\alpha}$ and $y^{(1-\alpha)}$.
```

```{r ces-solution, messages = FALSE, warnings = FALSE}
library(plotly)
alpha = 0.25
x <- seq(0, 10, length.out = 101)
y <- seq(0, 10, length.out = 101)
z <- outer(x, y, FUN = function(x, y) (x^0.5 + y^0.5)^2)
plot_ly(z = ~z, type = 'surface') 
```

```{r ces-check}
grade_this({
  if (!inherits(.result, c("plotly"))) {
    fail("Your class of your answer should be a plotly graph")}
  if (length(.result) != 8) { 
    fail("The result should a length of 8") }
  pass()
})
```

2. Graph the indifference curves for this utility function for $U(*) = 10$ and $U(*) = 25$. 

```{r cesIndifference, exercise=TRUE, exercise.reveal_solution = TRUE}

```

```{r cesIndifference-setup, messages = FALSE, warnings = FALSE}
library(ggplot2) 
```

```{r cesIndifference-solution, messages = FALSE, warnings = FALSE}
ggplot() +
  geom_function(fun = function(x) (10^0.5 - x^0.5)^2, color = 'red') + 
  geom_function(fun = function(x) (25^0.5 - x^0.5)^2, color = 'violet') + 
  lims(x = c(0,10), y = c(0, 10)) + 
  labs(x = "Good 1", y = "Good 2")
```

```{r cesIndifference-check}
grade_this({
  if (ith_geom_is(.result, "function", i = 1) & ith_geom_is(.result, "function", i = 2)) {
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

## Marginal Utility

The *marginal utility* of each good is the amount that utility changes as a result of a (very small?) change in the amount of that good. Mathematically, for *inifinitesimal* changes, marginal utility is the *derivative* of $u(x)$ with respect to $x_i$, or: 
$$MU_i = \frac{d}{dx_i}u(x_1, x_2, ...,x_n)$$

3. Run the following exercise that finds the marginal utility functions with respect to goods 1 and 2. 

```{r marginalUtility, exercise = TRUE}
u <- ysym('(x1^0.5 + x2^0.5)^2')
MU <- deriv(u, c('x1', 'x2'))
MU
```

```{r marginalUtility-solution, messages = FALSE, warnings = FALSE}
u <- ysym('(x1^0.5 + x2^0.5)^2')
MU <- deriv(u, c('x1', 'x2'))
MU
```

```{r marginalUtility-check}
grade_code()
```

4. Run this exercise that finds $MRS_{12} = \frac{MU_1}{MU_2}$ and simplifies the result. 

```{r marginalRateOfSubstitution, exercise = TRUE}
MRS <- (MU[1]/MU[2]) |>
  yac_str() |> 
  Simplify() |>
  ysym()
MRS
```

```{r marginalRateOfSubstitution-setup, messages = FALSE, warnings = FALSE}
u <- ysym('(x1^0.5 + x2^0.5)^2')
MU <- deriv(u, c('x1', 'x2'))
```

```{r marginalRateOfSubstitution-solution, messages = FALSE, warnings = FALSE}
MRS <- (MU[1]/MU[2]) |>
  yac_str() |> 
  Simplify() |>
  ysym()
MRS
```

```{r marginalRateOfSubstitution-check}
grade_code()
```
