---
title: "3.3 Monopoly"
output: 
  learnr::tutorial: 
    progressive: true
runtime: shiny_prerendered
author: 
  name: "Created by [Jim Bang](http://www.github.com/bangecon)"
  email: BangJamesT@sau.edu
  affiliation: "[St. Ambrose University](http://www.sau.edu)"
date: "`r format(Sys.Date(), '%d %b %Y')`"
---

<style type="text/css">
h1{font-size: 24pt}
h2{font-size: 20pt}
h3{font-size: 18pt}
h4,h5,h6{font-size: 16pt}
body{font-size: 16pt}
#doc-metadata {
  margin-bottom: 10px;
  font-size: 0.66em;
}
</style>

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
gradethis_setup(exercise.reveal_solution = TRUE, exercise.eval = TRUE)
knitr::opts_chunk$set(echo = FALSE, fig.height = 6, fig.width = 6)
```

## Natural Monopoly

As the manager of a power company, your firm faces a demand function estimated as: 

$$Q_d = 350 - 20Q$$

Internal cost analysis estimates the company's long-run cost function to be: 

$$C(Q) = \frac{Q^2}{200} + 10Q + \frac{10000}{Q}$$

### Graphical Approach

Plot the inverse demand, marginal revenue, long-run average cost, and long-run marginal cost functions. 

```{r naturalMonopoly, exercise = TRUE}

```

```{r naturalMonopoly-hint}

```

```{r naturalMonopoly-solution}
library(ggplot2)
ggplot() + 
  geom_function(fun = function(x) 17.5 - 0.05*x, aes(color = "D")) + 
  geom_function(fun = function(x) 17.5 - 0.1*x, aes(color = "MR")) + 
  geom_function(fun = function(x) x/200 + 10 + 10000/(x^2) - 10000/(x^3), aes(color = "LRAC")) + 
  geom_function(fun = function(x) x/100 + 10 - 10000/(x^2) + 20000/(x^3), aes(color = "LRMC")) +
  lims(x = c(0,200), y = c(0, 20)) +
  labs(x = "Quantity", y = "Price", color = "") + 
  scale_color_manual(breaks = c("D", "MR", "LRAC", "LRMC"), values =c("blue", "skyblue3", "maroon", "red"))
```

```{r naturalMonopoly-check}

```

### Computational Approach

Solve computationally for the profit-maximizing quantity, price, and profit.  

```{r equilibrium, exercise = TRUE}
```

```{r equilibrium-hint1}
# Be careful! 
# The profit function is cubic and has multiple quantities where the slope equals zero. 
# Set your starting `pars` option wisely!
```

```{r equilibrium-hint2}
# The "first" critical value is a (local) minumum at about q = 49. 
# Set the starting `pars` greater than this to get the correct profit-maximizing quantity. 
```

```{r equilibrium-solution}
library(MicroWithR)
pi <- function(x) {
  17.5*x - 0.05*x^2 - (x^2/200 + 10*x + 10000/x)
}
eq <- solnp2(
  pars = 50, 
  fun = pi, 
  LB = 0, 
  control = list(trace = 0), 
  max = TRUE
)
eq$pars
```

```{r equilibrium-check}
```

