---
title: "2.4 Firm Output Choice - Practice"
output: 
  learnr::tutorial: 
    progressive: true
runtime: shiny_prerendered
author: 
  name: "Created by [Jim Bang](http://www.github.com/bangecon)"
  email: BangJamesT@sau.edu
  affiliation: "[St. Ambrose University](http://www.sau.edu)"
date: "`r format(Sys.Date(), '%d %b %Y')`"
---

<style type="text/css">
#doc-metadata {
  margin-bottom: 10px;
  font-size: 0.8em;
}
</style>

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
gradethis_setup(exercise.reveal_solution = FALSE, exercise.eval = TRUE)
knitr::opts_chunk$set(echo = FALSE, fig.height = 6, fig.width = 6)
```

## Perfectly Competitive 

Suppose that the following supply and demand functions characterize the perfectly competitive market in which a firm produces: 

$$Q_d = 6000 - 1000P$$
$$Q_s = 2000P$$

1. Solve for the market equilibrium price using `nleqslv::nleqslv()`. 

```{r market, exercise = TRUE}

```

```{r market-solution}
library(nleqslv)
market <- function(x) {
  demand <- 6000 - 1000*x[1] - x[2]
  supply <- 2000*x[1] - x[2]
  return(c(demand, supply))
}
equilibrium <- nleqslv(c(1,1), market)
equilibrium
```


```{r market-check}
grade_this({if (!inherits(.result, c("list"))) {
    fail("Your class of your answer should be a list")}
  if (round(.result$x, 4) != c(20, 4000)) { 
    fail("The equilibrium quantity and price are 20 and 4000)") }
  pass()
})
```

2. Find a firm's profit-maximizing quantity if its cost function is given by: 

$$C(Q) = 0.125Q^3 - Q^2 + 3Q + 5$$

```{r firm, exercise = TRUE}

```

```{r firm-setup}
library(nleqslv)
market <- function(x) {
  demand <- 6000 - 1000*x[1] - x[2]
  supply <- 2000*x[1] - x[2]
  return(c(demand, supply))
}
equilibrium <- nleqslv(c(1,1), market)
```

```{r firm-solution}
library(MicroWithR)
pi <- function(x) {
  2*x - (0.125*x^3 - x^2 + 3*x + 5) 
}
solution <- solnp2(par = 10, fun = pi, LB = 0, max = TRUE)
solution
```

```{r firm-check}
grade_this({if (!inherits(.result, c("list"))) {
    fail("Your class of your answer should be a list")}
  if (round(.result$x, 3) != 4.775) { 
    fail("The equilibrium quantity for the firm is about 4.775)") }
  pass()
})
```

3. Graph the price, average cost, and marginal cost curves for the representative firm. 

```{r graph, exercise = TRUE}

```

```{r graph-setup}
library(nleqslv)
market <- function(x) {
  demand <- 6000 - 1000*x[1] - x[2]
  supply <- 2000*x[1] - x[2]
  return(c(demand, supply))
}
equilibrium <- nleqslv(c(1,1), market)
library(MicroWithR)
pi <- function(x) {
  2*x - (0.125*x^3 - x^2 + 3*x + 5) 
}
solution <- solnp2(par = 10, fun = pi, LB = 0, max = TRUE)
```

```{r graph-solution}
p <- equilibrium$x[1]
mc <- function(x) 0.375*x^2 - 2*x + 3
atc <- function(x) 0.125*x ^ 2 - x + 3 + 5 / x 
avc <- function(x) 0.125*x ^ 2 - x + 3
ggplot() + 
  geom_function(fun = mc, aes(color = "MC")) + 
  geom_function(fun = atc, aes(color = "ATC")) + 
  geom_function(fun = avc, aes(color = "AVC")) + 
  geom_hline(aes(yintercept = p, color = "P")) + 
  lims(x = c(0, 8), y = c(0, 10)) + 
  labs(x = "Quantity", "$", color = "") + 
  scale_color_manual(breaks = c("MC", "ATC", "AVC", "P"), 
                     values = c("red", "brown", "chocolate1", "blue"))
```

```{r graph-check}
grade_this({
  if (uses_geoms(.result, "function") & 
      uses_geoms(.result, "hline")) {
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

### Extra Practice

Try solving the problem with the following supply, demand, and cost functions: 

$$Q_d = 1,200 - 20P$$

$$Q_s = 100P$$

$$C(Q) = \frac{Q^3}{20} - \frac{9Q^2}{10} + 10Q + 5$$
