---
title: "2.1 Production and Production Functions  Practice"
output: 
  learnr::tutorial: 
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
author: 
  name: "Created by [Jim Bang](http://www.github.com/bangecon)"
  email: BangJamesT@sau.edu
  affiliation: "[St. Ambrose University](http://www.sau.edu)"
date: "`r format(Sys.Date(), '%d %b %Y')`"
---

<style type="text/css">
#doc-metadata {
  margin-bottom: 10px;
  font-size: 0.8em;
}
</style>

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(ggcheck)
library(Ryacas)
library(Deriv)
gradethis_setup(exercise.reveal_solution = FALSE)
knitr::opts_chunk$set(echo = FALSE, fig.height = 6, fig.width = 6)
```

## Increasing Returns to Scale

Suppose we have a production technology with diminishing returns to each individual input, but increasing returns to scale.  

$$ y = AL^{\alpha}K^{\beta},\text{ where}$$
$$ \alpha, \beta < 1 \text{, and}$$
$$\alpha + \beta > 1.$$
Suppose that $A = 100$ and $\alpha = \beta = 2/3$. 

1. Plot the short-run production function for $0 \le L \le 50$ setting $K = 50$.

```{r shortrun, exercise = TRUE}

```

```{r shortrun-solution}
library(ggplot2)
ggplot() +
  geom_function(fun = function(x) 100*x^(2/3)*50^(2/3)) + 
  lims(x = c(0, 50)) + 
  labs(x = "Labor", y = "Output")
```

```{r shortrun-check}
grade_this({
  if (ith_geom_is(.result, "function", i = 1) ) {
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

2. Calculate the output for the following combinations of inputs: 
  a. $(L, K) = (25, 25)$
  b. $(L, K) = (50, 50)$
  c. $(L, K) = (75, 75)$

```{r scale, exercise = TRUE}

```

```{r scale-solution}
y = function(L, K) {
  100*L^(2/3)*K^(2/3)
}
y(25,25)
y(50,50)
y(75,75)
```

```{r scale-check}
grade_this({if (round(.result, 2) != 31628.72) { 
    fail("The output for L = K = 75 should equal 31,628.72.") }
  pass("Notice that when we double or triple the inputs, production more than doubles or triples!")
})
```

3. Plot the isoquants for $Q = 25000$, $Q = 50000$, and $Q = 75000$

```{r isoquants, exercise = TRUE}

```

```{r isoquants-setup}
library(ggplot2)
```

```{r isoquants-solution}
ggplot() + 
  geom_function(fun = function(x) (25000 / (100*x^(2/3)))^(3/2), aes(color = "25000")) + 
  geom_function(fun = function(x) (50000 / (100*x^(2/3)))^(3/2), aes(color = "50000")) + 
  geom_function(fun = function(x) (75000 / (100*x^(2/3)))^(3/2), aes(color = "75000")) + 
  lims(x = c(0,200), y = c(0,200)) + 
  labs(x = "Labor", y = "Capital", color = "Output") + 
  scale_color_manual(values = c('red', 'blue', 'purple'))
```

```{r isoquants-check}
grade_this({
  if (ith_geom_is(.result, "function", i = 1) & ith_geom_is(.result, "function", i = 2) & ith_geom_is(.result, "function", i = 3)) {
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

