---
title: "2.4 Competitive Firms' Output Choices in the Short Run"
output: 
  learnr::tutorial: 
    progressive: true
runtime: shiny_prerendered
author: 
  name: "Created by [Jim Bang](http://www.github.com/bangecon)"
  email: BangJamesT@sau.edu
  affiliation: "[St. Ambrose University](http://www.sau.edu)"
date: "`r format(Sys.Date(), '%d %b %Y')`"
---

<style type="text/css">
h1{font-size: 24pt}
h2{font-size: 20pt}

h3{font-size: 18pt}
h4,h5,h6{font-size: 16pt}
body{font-size: 16pt}
#doc-metadata {
  margin-bottom: 10px;
  font-size: 0.66em;
}
</style>

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
gradethis_setup(exercise.reveal_solution = TRUE)
knitr::opts_chunk$set(echo = FALSE, fig.height = 6, fig.width = 6)
```

## Review: The Firm's Three Problems:

1. ~~Cost-Minimizing Combination of Inputs.~~
2. <span style="color: red;">Profit-Maximizing Level of Output. </span>
3. Long-Run Readjustment.

### Perfectly Competitive Output Markets

$$\matrix{ \matrix{\text{1. Many sellers and buyers} \\ \text{2. Homogeneous product} \\ \text{3. Perfect information}} \Biggr\} \\ \text{4. Free entry and exit} \\ \text{  }} \matrix{\text{ } \\ \implies \\ \text{ } \\ \implies \\ \text{ } }  \matrix{\Biggl\{ \matrix{\text{Firms are} \\ \text{price} \\ \text{takers}} \\ \text{Profits/losses} \\ \text{are temporary}}$$

### The Producer's Problem

1. Objective: 
2. Endogenous (choice) variable: 
3. Exogenous variable(s): 

### The Profit Function: General Form 

$$\max_Q \{ \pi(Q) = TR(Q) - C(Q) \} $$
Since price is *exogenous* (determined by the market), $TR = PQ$:
$$\pi(Q) = PQ - C(Q)$$

### The Profit Function: Per-Unit Form

Since $C(Q) = ATC(Q) \cdot Q$, we can rewrite profits as: 

$$\pi(Q) = [P - ATC(Q)]Q$$

### Market Equilibrium

$$Q_d = 6,000,000 - 2,000,000P$$
$$Q_s = 10,000,000*P$$ 

Solve for the market equilibrium price and quantity.

```{r market, exercise = TRUE}

```

```{r market-solution}
library(nleqslv) 
market <- function(x) {
    supply <- 10000000*x[1] - x[2]
    demand <- 6000000 - 2000000*x[1] - x[2]
    return(c(supply, demand))
}
equilibrium <- nleqslv(c(2,2), market)
P <- equilibrium$x[1]
P
```

```{r market-hint}
# In our introduction module, we solved the following example where:
# Q_s = P 
# Q_d = 10 - P 
market <- function(x) {                 # x[1] is price; x[2] is quantity
    supply <- x[1] - x[2]               # nleqslv finds "zeros": P - Q_s = 0
    demand <- 10 - x[1] - x[2]          # 10 - P - Q_d = 0
    return(c(supply, demand))           # Return a vector function
}
equilibrium <- nleqslv(c(2,2), market)  # Solve where both equations equal 0.
```

```{r market-check}
grade_code()
```

## Profit Maximization

When we left our representative firm, we could express the short-run production function (with capacity optimized for $Q = 2500$) in terms of the cube-root of the labor input, $f(L|K=25) = 100*25^{2/3}*L^{1/3}$. The cost function is:

$$C(Q) = 10L(Q) + 20K.$$

Substituting $K=25$, and solving $f(L|K = 25)$ for $L$, we have: 

$$L(Q) = \frac{Q^3}{625,000,000}$$

$$C(Q) = \frac{Q^3}{62,500,000} + 500$$

Note that the firm chose (optimized) its capacity on expecting to sell 2,500 units of output. Suppose that after building its capital input, market conditions yielded the following supply and demand functions: 

### Analytical Approach

Run the following code to solve the producer's problem analytically using `Ryacas`. 

<details style="line-height:105%"><summary>Click here to view the steps for the analytical solution.</summary>
1. Load `Ryacas`.
2. Define the profit function. 
3. Derive $d\pi/dQ$ and simplify. 
4. Solve $d\pi/dQ = 0$ for $Q$, remove the variables, and reformat to an ordinary `R` object. 
</details>

```{r profitmaxAnalytical, exercise = TRUE}
library(Ryacas)
pi <- ysym('0.5*Q - 500 - Q^3 / 62500000') # Define the profit function as a yac_symbol
dpi   <- deriv(pi, 'Q') |>                 # Take the derivative (and simplify)
  simplify()
qStar <- solve(dpi, 'Q')[1] |>          # Solve for Q (take the first [>0] value)
  y_rmvars()                               # Isolate the solution.
```

```{r profitmaxAnalytical-solution}
library(Ryacas)
pi <- ysym('0.5*Q - 500 - Q^3 / 62500000') # 
dpi   <- deriv(pi, 'Q') |>
  simplify()
qStar <- solve(dpi, 'Q')[1] |>
  y_rmvars() 
```

```{r profitmaxAnalytical-check}
grade_code()
```

```{r, message=FALSE}
library(Ryacas)
pi <- ysym('0.5*Q - 500 - Q^3 / 62500000') # 
dpi   <- deriv(pi, 'Q') |>
  simplify()
qStar <- solve(dpi, 'Q')[1] |>
  y_rmvars() 
```

The first-order condition for the profit-maximization problem is: 

$$`r Ryacas::tex(dpi)` = 0$$

This is the condition that sets $MR = MC$. 

The profit-maximizing quantity is: 

$$Q^* = `r tex(qStar)` \approx `r round(as_r(qStar), 3)`$$

### Graphical Approach

Run the following code to construct the following graphs: 

1. Total revenue and total cost versus quantity (top left);
2. Total profit versus quantity (top right);
3. Marginal revenue, marginal cost, average cost, and average variable cost versus quantity (bottom left); 
4. Marginal (change in) profit versus quantity (bottom right); 

```{r profitmaxGraph, exercise = TRUE}
par(mfrow = c(2,2))                                  # Set a 2-by-2 plot window.
curve(500 + x^3 / 62500000,                          # Total cost curve
  from = 0, to = 5000, ylim = c(0,2500),             # Start/end x, y-axis limits
  xlab = "Q", ylab = "TR/TC", col = 'red')           # Axis labels, line color
curve(x / 2, col = 'blue', add = T)                  # Total revenue curve
b = 500 + as_r(qStar)^3 / 62500000 - 0.5*as_r(qStar) # Tangent line intercept
curve(b + 0.5 * x,                                   # Cost tangent at Qstar
      lty = 'dashed', col = 'red', add = T)
curve(0.5 * x - (500 + x^3 / 62500000),              # Total profit curve
  from = 0, to = 5000, ylim = c(-500, 600), xlab = "Q", ylab = "Profit")
lines(x = c(0, 5000), y = c(0, 0), lty = 'dashed')   # Zero-profit reference line
curve(3*x^2 / 62500000,                              # Marginal cost curve
  from = 0, to = 5000, ylim = c(0, 1), xlab = "Q", ylab = "MC/AC", col = 'red')
curve(x^2/62500000 + 500/x, col = 'brown', add = T)  # Average total cost curve
curve(x^2 / 62500000,                                # Average variable cost curve
      col = 'brown', lty = 'dashed', add = T) 
lines(x = c(0, 5000), y = c(0.5, 0.5), col = 'blue') # Price/marginal revenue curve
curve(0.5 - (3*x^2 / 62500000),                      # Marginal profit curve
  from = 0, to = 5000, ylim = c(-0.6, 0.6), xlab = "Q", ylab = expression(DeltaProfit))
lines(x = c(0, 5000), y = c(0, 0), lty = 'dashed')   # dProfit = 0 reference line
```

```{r profitmaxGraph-setup}
library(Ryacas)
pi <- ysym('0.5*Q - 500 - Q^3 / 62500000') # 
dpi   <- deriv(pi, 'Q') |>
  simplify()
qStar <- solve(dpi, 'Q')[1] |>
  y_rmvars() 
```

```{r profitmaxGraph-solution}
par(mfrow = c(2,2))                                  # Set a 2-by-2 plot window.
curve(500 + x^3 / 62500000,                          # Total cost curve
  from = 0, to = 5000, ylim = c(0,2500),             # Start/end x, y-axis limits
  xlab = "Q", ylab = "TR/TC", col = 'red')           # Axis labels, line color
curve(x / 2, col = 'blue', add = T)                  # Total revenue curve
b = 500 + as_r(qStar)^3 / 62500000 - 0.5*as_r(qStar) # Tangent line intercept
curve(b + 0.5 * x,                                   # Cost tangent at Qstar
      lty = 'dashed', col = 'red', add = T)
curve(0.5 * x - (500 + x^3 / 62500000),              # Total profit curve
  from = 0, to = 5000, ylim = c(-500, 600), xlab = "Q", ylab = "Profit")
lines(x = c(0, 5000), y = c(0, 0), lty = 'dashed')   # Zero-profit reference line
curve(3*x^2 / 62500000,                              # Marginal cost curve
  from = 0, to = 5000, ylim = c(0, 1), xlab = "Q", ylab = "MC/AC", col = 'red')
curve(x^2/62500000 + 500/x, col = 'brown', add = T)  # Average total cost curve
curve(x^2 / 62500000,                                # Average variable cost curve
      col = 'brown', lty = 'dashed', add = T) 
lines(x = c(0, 5000), y = c(0.5, 0.5), col = 'blue') # Price/marginal revenue curve
curve(0.5 - (3*x^2 / 62500000),                      # Marginal profit curve
  from = 0, to = 5000, ylim = c(-0.6, 0.6), xlab = "Q", ylab = expression(DeltaProfit))
lines(x = c(0, 5000), y = c(0, 0), lty = 'dashed')   # dProfit = 0 reference line
```

```{r profitmaxGraph-check}
grade_code()
```

### Computational Approach

Find the solution to the profit-maximization problem numerically. 

Careful! Since our profit function has multiple "critical values" (points where change = 0), getting the correct solution will be sensitive to the initial value (`pars` option)

```{r profitmaxNumerical, exercise = TRUE}

```

```{r profitmaxNumerical-solution}
library(Rsolnp)
loss <- function(x) -(0.5*x - (x^3 / 62500000 + 500))  # Define losses
profitMax <- solnp(pars = 1, fun = loss, LB = 0)
```

```{r profitmaxNumerical-hint}
# Don't forget to load your libraries!

# Define losses (minus-profits) as the function(x) because 'solnp()' finds a min 

# Since there's no constraint, we only need three arguments: 
  # (1) a starting value of Q; 
  # (2) the objective function; 
  # (3) the lower bound for Q, "x" (unnecessary with an interior solution). 
```

```{r profitmaxNumerical-check}
grade_this(
  if(round(.result$pars, 1) != 3227.5) {
    fail("The solution should be about 3227.5.")} 
  pass("Same as we got from the analytical approach")
)
```
