---
title: "2.4 Firm Output Choice"
output: 
  learnr::tutorial: 
    progressive: true
runtime: shiny_prerendered
author: 
  name: "Created by [Jim Bang](http://www.github.com/bangecon)"
  email: BangJamesT@sau.edu
  affiliation: "[St. Ambrose University](http://www.sau.edu)"
date: "`r format(Sys.Date(), '%d %b %Y')`"
---

<style type="text/css">
h1{font-size: 24pt}
h2{font-size: 20pt}
h3{font-size: 18pt}
h4,h5,h6{font-size: 16pt}
body{font-size: 16pt}
#doc-metadata {
  margin-bottom: 10px;
  font-size: 0.66em;
}
</style>

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(ggcheck)
library(Ryacas)
gradethis_setup(exercise.reveal_solution = TRUE)
knitr::opts_chunk$set(echo = FALSE, fig.align = 'center', fig.height = 6, fig.width = 6, message = FALSE, warning = FALSE)
```

## Review: The Firm's Three Problems:

1. ~~Cost-Minimizing Combination of Inputs.~~
2. <span style="color: red;">Profit-Maximizing Level of Output. </span>
3. Long-Run Readjustment.

### Perfectly Competitive Output Markets

$$\matrix{ \matrix{\text{1. Many sellers and buyers} \\ \text{2. Homogeneous product} \\ \text{3. Perfect information}} \Biggr\} \\ \text{4. Free entry and exit} \\ \text{  }} \matrix{\text{ } \\ \implies \\ \text{ } \\ \implies \\ \text{ } }  \matrix{\Biggl\{ \matrix{\text{Firms are} \\ \text{price} \\ \text{takers}} \\ \text{Profits/losses} \\ \text{are temporary}}$$

### The Producer's Problem

1. Objective: 
2. Endogenous (choice) variable: 
3. Exogenous variable(s): 

### The Profit Function: General Form 

$$\max_Q \{ \pi(Q) = TR(Q) - C(Q) \} $$
Since price is *exogenous* (determined by the market), $TR = PQ$:
$$\pi(Q) = PQ - C(Q)$$

### The Profit Function: Per-Unit Form

Since $C(Q) = ATC(Q) \cdot Q$, we can rewrite profits as: 

$$\pi(Q) = [P - ATC(Q)]Q$$

### Market Equilibrium

$$Q_d = 6,000,000 - 2,000,000P$$
$$Q_s = 10,000,000*P$$ 

Solve for the market equilibrium price and quantity.

```{r market, exercise = TRUE}

```

```{r market-solution}
library(nleqslv) 
market <- function(x) {
  supply <- 10000000*x[1] - x[2]
  demand <- 6000000 - 2000000*x[1] - x[2]
  return(c(supply, demand))
}
equilibrium <- nleqslv(c(2,2), market)
P <- equilibrium$x[1]
P
```

```{r market-hint}
# In our introduction module, we solved the following example where:
# Q_s = P 
# Q_d = 10 - P 
market <- function(x) {                 # x[1] is price; x[2] is quantity
    supply <- x[1] - x[2]               # nleqslv finds "zeros": P - Q_s = 0
    demand <- 10 - x[1] - x[2]          # 10 - P - Q_d = 0
    return(c(supply, demand))           # Return a vector function
}
equilibrium <- nleqslv(c(2,2), market)  # Solve where both equations equal 0.
```

```{r market-check}
grade_code()
```

## Profit Maximization

When we left our representative firm, we could express the short-run production function (with capacity optimized at $K = 25$ for $Q = 2500$) in terms of the cube-root of the labor input as: 

$$f(L|K=25) = 100(25^{2/3})(L^{1/3})$$. 

Solving $f(L|K = 25)$ for $L$, we have: 

$$L(Q) = \Bigg( \frac{Q}{100 \cdot 25^{2/3}} \Bigg)^3 =\frac{Q^3}{625,000,000}$$

The cost function is:

$$C(Q | K = 25) = 10L(Q) + 20(25)$$


$$C(Q) = \frac{Q^3}{62,500,000} + 500$$

### Computational Approach

Find the solution to the profit-maximization problem numerically. 

Careful! Since our profit function has multiple "critical values" (points where change = 0), getting the correct solution will be sensitive to the initial value (`pars` option)

```{r profitmaxNumerical, exercise = TRUE}

```

```{r profitmaxNumerical-solution}
library(MicroWithR)
P = 0.5
pi <- function(x)
  P * x - (x ^ 3 / 62500000 + 500)
profitMax <-
  solnp2(
    pars = 1,
    fun = pi,
    LB = 0,
    max = TRUE,
    control = list(trace = 0)
  )
```

```{r profitmaxNumerical-hint}
# Don't forget to load your libraries!

# Define losses (minus-profits) as the function(x) because 'solnp()' finds a min 

# Since there's no constraint, we only need three arguments: 
  # (1) a starting value of Q; 
  # (2) the objective function; 
  # (3) the lower bound for Q, "x" (unnecessary with an interior solution). 
```

```{r profitmaxNumerical-check}
grade_this(
  if(round(.result$pars, 1) != 3227.5) {
    fail("The solution should be about 3227.5.")} 
  pass("Same as we got from the analytical approach")
)
```

### Graphical Approach

Construct the following graphs: 

1. Total revenue and total cost versus quantity
```{r trtcGraph, exercise = TRUE, fig.align='center', fig.height=6}

```

```{r trtcGraph-setup}
library(MicroWithR)
P = 0.5
pi <- function(x)
  P * x - (x ^ 3 / 62500000 + 500)
profitMax <-
  solnp2(
    pars = 1,
    fun = pi,
    LB = 0,
    max = TRUE,
    control = list(trace = 0)
  )
```

```{r trtcGraph-solution}
library(ggplot2)
tcGraph <- ggplot() + 
  geom_function(fun = function(x) 500 + x^3 / 62500000) + 
  geom_function(fun = function(x) P*x) + 
  lims(x = c(0, 5000), y = c(0, 2500)) 
```

```{r trtcGraph-check}
grade_this({
  if (ith_geom_is(.result, "function", i = 1) & 
      ith_geom_is(.result, "function", i = 2) & 
      ith_geom_is(.result, "function", i = 3)) {
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

2. Marginal revenue, marginal cost, average cost, and average variable cost versus quantity; 
```{r mrmcGraph, exercise = TRUE, fig.align='center', fig.height=6}

```

```{r mrmcGraph-setup}
library(MicroWithR)
library(ggplot2)
P = 0.5
pi <- function(x)
  P * x - (x ^ 3 / 62500000 + 500)
profitMax <-
  solnp2(
    pars = 1,
    fun = pi,
    LB = 0,
    max = TRUE,
    control = list(trace = 0)
  )
```

```{r mrmcGraph-solution}
ggplot() + 
  geom_function(fun = function(x) x^2/62500000 + 500/x) + 
  geom_function(fun = function(x) x^2 / 62500000) + 
  geom_function(fun = function(x) 3*x^2 / 62500000) + 
  geom_hline(yintercept = P) +
  lims(x = c(0, 5000), y = c(0, 1)) 
```

```{r mrmcGraph-check}
grade_this({
  if (ith_geom_is(.result, "function", i = 1) & 
      ith_geom_is(.result, "function", i = 2) & 
      ith_geom_is(.result, "function", i = 3) & 
      ith_geom_is(.result, "hline", i = 4)) {
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

3. Total profit versus quantity;

```{r piGraph, exercise = TRUE, fig.align='center', fig.height=6}

```

```{r piGraph-setup}
library(MicroWithR)
library(ggplot2)
P = 0.5
pi <- function(x)
  P * x - (x ^ 3 / 62500000 + 500)
profitMax <-
  solnp2(
    pars = 1,
    fun = pi,
    LB = 0,
    max = TRUE,
    control = list(trace = 0)
  )
```

```{r piGraph-solution}
piGraph <- ggplot() + 
  geom_function(fun = function(x) P * x - (500 + x^3 / 62500000)) + 
  geom_hline(yintercept = 0) +
  lims(x = c(0, 5000), y = c(-500, 600)) 
```

```{r piGraph-check}
grade_this({
  if (ith_geom_is(.result, "function", i = 1) & 
      ith_geom_is(.result, "function", i = 2) & 
      ith_geom_is(.result, "function", i = 3) & 
      ith_geom_is(.result, "hline", i = 4)) {
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

4. Marginal (change in) profit versus quantity (bottom right); 

```{r dpiGraph, exercise = TRUE, fig.align='center', fig.height=6}

```

```{r dpiGraph-setup}
library(MicroWithR)
library(ggplot2)
P = 0.5
pi <- function(x)
  P * x - (x ^ 3 / 62500000 + 500)
profitMax <-
  solnp2(
    pars = 1,
    fun = pi,
    LB = 0,
    max = TRUE,
    control = list(trace = 0)
  )
```

```{r dpiGraph-solution}
ggplot() + 
  geom_function(fun = function(x) 0.5 - (3*x^2 / 62500000)) + 
  geom_hline(yintercept = 0) +
  lims(x = c(0, 5000), y = c(-0.6, 0.6)) 
```

```{r dpiGraph-check}
grade_this({
  if (ith_geom_is(.result, "function", i = 1) & 
      ith_geom_is(.result, "hline", i = 2)) {
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

### Comparison of the Graphical Equilibria

```{r profitmaxGraphSetup, echo = FALSE}
library(MicroWithR)
P = 0.5
pi <- function(x)
  P * x - (x ^ 3 / 62500000 + 500)
profitMax <-
  solnp2(
    pars = 1,
    fun = pi,
    LB = 0,
    max = TRUE,
    control = list(trace = 0)
  )
qStar <- profitMax$pars
library(ggplot2)
tcGraph <- ggplot() + 
  geom_function(fun = function(x) 500 + x^3 / 62500000, aes(color = "TC")) + 
  geom_function(fun = function(x) P*x, aes(color = "TR")) + 
  geom_function(fun = function(x) 500 + qStar^3 / 62500000 - 0.5*qStar + 0.5 * x, 
                color = 'red', linetype = 'dashed') +
  lims(x = c(0, 5000), y = c(0, 2500)) + 
  labs(x = "Quantity", y = "$", color = "") +
  scale_colour_manual(breaks = c("TC", "TR"), values = c("red","blue")) + 
  theme(legend.position = "bottom")
piGraph <- ggplot() + 
  geom_function(fun = function(x) P * x - (500 + x^3 / 62500000), 
                aes(linetype = "Profit")) + 
  geom_hline(aes(yintercept = 0, linetype = 'Profit=0')) +
  lims(x = c(0, 5000), y = c(-500, 600)) + 
  labs(x = "Quantity", y = "Profit", linetype = "") +
  scale_linetype_manual(breaks = c("Profit", "Profit=0"), values = c("solid","dashed")) + 
  theme(legend.position = "bottom")
acGraph <- ggplot() + 
  geom_function(fun = function(x) x^2/62500000 + 500/x, aes(color = "ATC")) + 
  geom_function(fun = function(x) x^2 / 62500000, aes(color = "AVC")) + 
  geom_function(fun = function(x) 3*x^2 / 62500000, aes(color = "MC")) + 
  geom_hline(aes(yintercept = P, color = 'P')) +
  lims(x = c(0, 5000), y = c(0, 1)) + 
  labs(x = "Quantity", y = "$", color = "") +
  scale_color_manual(breaks = c("ATC", "AVC", "MC", "P"), values = c("brown", "chocolate1","red", "blue")) + 
  theme(legend.position = "bottom")
dpiGraph <- ggplot() + 
  geom_function(fun = function(x) 0.5 - (3*x^2 / 62500000), aes(linetype = 'dProfit')) + 
  geom_hline(aes(yintercept = 0, linetype = 'dProfit=0')) +
  lims(x = c(0, 5000), y = c(-0.6, 0.6)) + 
  labs(x = "Quantity", y = "dProfit", linetype = "") + 
  scale_linetype_manual(breaks = c("dProfit", "dProfit=0"), values = c("solid","dashed")) + 
  theme(legend.position = "bottom")
```

We can name the plots and combine them using `ggpubr::ggarrange()` as I've done below (with some extra aesthetics. For details see below. 

```{r, echo = TRUE}
library(ggpubr)
ggarrange(tcGraph, piGraph, acGraph, dpiGraph)
```

<details style="line-height:105%">

<summary>Click here to see the code for adding all of the fancy color, linetype, and legend aesthetics.</summary>

```{r, echo = TRUE}
library(MicroWithR)
P = 0.5
pi <- function(x)
  P * x - (x ^ 3 / 62500000 + 500)
profitMax <-
  solnp2(
    pars = 1,
    fun = pi,
    LB = 0,
    max = TRUE,
    control = list(trace = 0)
  )
qStar <- profitMax$pars
library(ggplot2)
tcGraph <- ggplot() + 
  geom_function(fun = function(x) 500 + x^3 / 62500000, aes(color = "TC")) + 
  geom_function(fun = function(x) P*x, aes(color = "TR")) + 
  geom_function(fun = function(x) 500 + qStar^3 / 62500000 - 0.5*qStar + 0.5 * x, 
                color = 'red', linetype = 'dashed') +
  lims(x = c(0, 5000), y = c(0, 2500)) + 
  labs(x = "Quantity", y = "$", color = "") +
  scale_colour_manual(breaks = c("TC", "TR"), values = c("red","blue")) + 
  theme(legend.position = "bottom")
piGraph <- ggplot() + 
  geom_function(fun = function(x) P * x - (500 + x^3 / 62500000), 
                aes(linetype = "Profit")) + 
  geom_hline(aes(yintercept = 0, linetype = 'Profit=0')) +
  lims(x = c(0, 5000), y = c(-500, 600)) + 
  labs(x = "Quantity", y = "Profit", linetype = "") +
  scale_linetype_manual(breaks = c("Profit", "Profit=0"), values = c("solid","dashed")) + 
  theme(legend.position = "bottom")
acGraph <- ggplot() + 
  geom_function(fun = function(x) x^2/62500000 + 500/x, aes(color = "ATC")) + 
  geom_function(fun = function(x) x^2 / 62500000, aes(color = "AVC")) + 
  geom_function(fun = function(x) 3*x^2 / 62500000, aes(color = "MC")) + 
  geom_hline(aes(yintercept = P, color = 'P')) +
  lims(x = c(0, 5000), y = c(0, 1)) + 
  labs(x = "Quantity", y = "$", color = "") +
  scale_color_manual(breaks = c("ATC", "AVC", "MC", "P"), values = c("brown", "chocolate1","red", "blue")) + 
  theme(legend.position = "bottom")
dpiGraph <- ggplot() + 
  geom_function(fun = function(x) 0.5 - (3*x^2 / 62500000), aes(linetype = 'dProfit')) + 
  geom_hline(aes(yintercept = 0, linetype = 'dProfit=0')) +
  lims(x = c(0, 5000), y = c(-0.6, 0.6)) + 
  labs(x = "Quantity", y = "dProfit", linetype = "") + 
  scale_linetype_manual(breaks = c("dProfit", "dProfit=0"), values = c("solid","dashed")) + 
  theme(legend.position = "bottom")
```
</details>

### Analytical Approach

The analytical solution undertakes the following steps:  

1. Load `Ryacas`.  
2. Define the profit function. 
3. Derive $d\pi/dQ$ and simplify. 
4. Solve $d\pi/dQ = 0$ for $Q$, remove the variables, and reformat as an `R` object. 

The following code solves the producer's problem analytically using `Ryacas`.  

```{r profitmaxAnalytical, echo = TRUE}
library(Ryacas)
pi <- ysym('0.5*Q - 500 - Q^3 / 62500000') # Define the profit function as a yac_symbol
dpi   <- deriv(pi, 'Q') |>                 # Take the derivative (and simplify)
  simplify()
qStar <- solve(dpi, 'Q')[1] |>          # Solve for Q (take the first [>0] value)
  y_rmvars()                               # Isolate the solution.
```

The first-order condition for the profit-maximization problem is: 

$$`r Ryacas::tex(dpi)` = 0$$

This is the condition that sets $MR = MC$. 

The profit-maximizing quantity is: 

$$Q^* = `r tex(qStar)` \approx `r round(as_r(qStar), 3)`$$
