---
title: "1.6 Comparative Statics II: Price Changes"
editor_options: 
  markdown: 
    wrap: 72
output: 
  learnr::tutorial: 
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
author: 
  name: "Created by [Jim Bang](http://www.github.com/bangecon)"
  email: BangJamesT@sau.edu
  affiliation: "[St. Ambrose University](http://www.sau.edu)"
date: "`r format(Sys.Date(), '%d %b %Y')`"
---

<style type="text/css">
h1{font-size: 24pt}
h2{font-size: 20pt}
h3{font-size: 18pt}
h4,h5,h6{font-size: 16pt}
body{font-size: 16pt}
#doc-metadata {
  margin-bottom: 10px;
  font-size: 0.66em;
}
</style>

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(Ryacas)
gradethis_setup()
tutorial_options(exercise.eval = TRUE, exercise.reveal_solution = TRUE)
knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE, 
  fig.align = "center", fig.height = 6, fig.width = 6)
knitr::include_graphics
par(mar = c(4, 4, 1, 1))
```

## Price Change 

Let's analyze the effect of a $\$0.50$ increase in price, and a $\$0.50$ decrease in price for good 1.

### Computational Approach

1.  For a consumer with the same preferences (Cobb-Douglass with $\alpha = 0.25$), and facing the price of good 2 and income ($p_2 = 1$), compute the new equilibrium for an increase in the price of good 1 from $\$1.50$ to $\$2.00$ (`dp1 = 0.5`).
2.  Print the new optimal bundles.

```{r priceChange, exercise = TRUE}


```

```{r priceChange-setup}
library(MicroWithR)
ux <- function(x)
  x[1] ^ (a) * x[2] ^ (1 - a)
bc <- function(x)
  p1 * x[1] + p2 * x[2]
a = 0.25
p1 = 1.5
p2 = 1
m = 100 
eq <-
  solnp2(
    c(1, 1),
    ux,
    ineqfun = bc,
    ineqLB = 0,
    ineqUB = m,
    LB = c(0, 0),
    UB = c(Inf, Inf), 
    control = list(trace = 0),
    max = TRUE
  )
```

```{r priceChange-solution}
dp1 = 0.5
bc2 <- function(x)
  (p1 + dp1) * x[1] + p2 * x[2]
dpEq <-
  solnp2(
    c(1, 1),
    ux,
    ineqfun = bc2,
    ineqLB = 0,
    ineqUB = m,
    LB = c(0, 0),
    UB = c(Inf, Inf), 
    control = list(trace = 0),
    max = TRUE
  )
```

```{r priceChange-check}
grade_this({if (!inherits(.result, c("list"))) {
    fail("Your class of your answer should be a list")}
  if (.result$pars != c(17.5, 78.75)) { 
    fail("The optimal amount of goods one and two after the change should be 17.5 and 78.75") }
  pass()
})
```

```{r dpEquilibrium}
dp1 = 0.5
bc2 <- function(x)
  (p1 + dp1) * x[1] + p2 * x[2]
dpEq <-
  solnp2(
    c(1, 1),
    ux,
    ineqfun = bc2,
    ineqLB = 0,
    ineqUB = m,
    LB = c(0, 0),
    control = list(trace = 0),
    UB = c(Inf, Inf), 
    max = TRUE
  )
```

Consumption of good 1 after the price increase is: 

$$x_1^{\dagger\dagger} = `r round(dpEq$pars[1], 3)`$$

$$\frac{\Delta x_1}{\Delta p_1} \Bigg|_{x_1 = x_1^*} \approx `r round((dpEq$pars[1] - eq$pars[1])/(dp1), 3)`$$

### Graphical Approach

Add the `geom_function()` layers to the graph to add the budget line and indifference curve for the equilibrium after the increase in price. Include the `geom_point()` for the new optimal bundle and label it using `geom_text()`. 

```{r priceChangeGraph, exercise = TRUE, exercise.setup = 'dpEquilibrium'}
# We solved for consumption (and utility at that consumption) for the original price, the price after an increase, and after a decrease. 


```

```{r priceChangeGraph-setup}
library(MicroWithR)
ux <- function(x)
  x[1] ^ (a) * x[2] ^ (1 - a)
bc <- function(x)
  p1 * x[1] + p2 * x[2]
a = 0.25
p1 = 1.5
p2 = 1
m = 100 
eq <-
  solnp2(
    c(1, 1),
    ux,
    ineqfun = bc,
    ineqLB = 0,
    ineqUB = m,
    LB = c(0, 0),
    UB = c(Inf, Inf), 
    control = list(trace = 0),
    max = TRUE
  )
dp1 = 0.5
bc2 <- function(x)
  (p1 + dp1) * x[1] + p2 * x[2]
dpEq <-
  solnp2(
    c(1, 1),
    ux,
    ineqfun = bc2,
    ineqLB = 0,
    ineqUB = m,
    LB = c(0, 0),
    UB = c(Inf, Inf), 
    control = list(trace = 0),
    max = TRUE
  )
```

```{r priceChangeGraph-solution}
library(ggplot2)
p <- ggplot() +
  geom_function(
    fun = function(x)
      ux(eq$pars) ^ (1 / (1 - a)) / (x ^ (a / (1 - a))),
    color = 'darkgreen'
  ) +
  geom_function(
    fun = function(x)
      m / p2 - x * p1 / p2,
    color = 'red'
  ) +
  geom_point(aes(x = eq$pars[1], y = eq$pars[2]), color = 'blue') +
  geom_text(
    aes(x = eq$pars[1], y = eq$pars[2]),
    label = paste0(
      'x* = (',
      round(eq$pars[1], 2),
      ', ',
      round(eq$pars[2], 2),
      ') \n u(x*) = ',
      round(ux(eq$pars), 2)
    ),
    hjust = 0,
    vjust = 0
  ) +
  lims(x = c(0, 100), y = c(0, 100)) +
  labs(x = "Good 1", y = "Good 2")
p + geom_function(
  fun = function(x)
    ux(dpEq$pars) ^ (1 / (1 - a)) / (x ^ (a / (1 - a))),
  color = 'darkgreen'
) +
  geom_function(
    fun = function(x)
      (m + dm) / p2 - x * p1 / p2,
    color = 'red'
  ) +
  geom_point(aes(x = dpEq$pars[1], y = dpEq$pars[2]), color = 'blue') +
  geom_text(
    aes(x = dpEq$pars[1], y = dpEq$pars[2]),
    label = paste0(
      'x* = (',
      round(dpEq$pars[1], 2),
      ', ',
      round(dpEq$pars[2], 2),
      ') \n u(x*) = ',
      round(ux(dpEq$pars), 2)
    ),
    hjust = 0,
    vjust = 0
  )
```

```{r priceChangeGraph-check}
grade_this({
  if (uses_geoms(.result, "function") & uses_geoms(.result, "point")) {
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

## Demand Curves

Recall from the previous section (and its appendix) that the general form for the demand function for good 1 for a consumer with Cobb-Douglas preferences is: 

$$x_1(p_1, \alpha, m) = \frac{\alpha m}{p_1}$$

Plot the demand curve for good 1 from the example ($\alpha = 0.25$, $p_2 = 1$, and $m = 100$).

```{r cdDemand, exercise = TRUE}

```

```{r cdDemand-setup}
library(ggplot2)
a = 0.25
p1 = 1.5
p2 = 1
m = 100 
```

```{r cdDemand-solution}
ggplot() + 
  geom_function(fun = function(x) a * m / x) + 
  lims(x = c(0, 150)) +
  labs(x = "Income", y = "Consumption of Good 1") 
```

```{r cdDemand-check}
grade_this({
  if (uses_geoms(.result, "function")) {
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

Note that this plots consumption of the good on the y axis and price on the x axis, which is the opposite of how we usually see it. We get the convention of plotting price (the independent variable?) on the y axis from Alfred Marshall. Why do you think Alfred Marshall would graph demand functions "backwards?"

It turns out this function is not too hard to invert using normal algebra. We can alsu do it using the `coord_flip()` function from `ggplot2`. Do this to plot the demand curve in the usual, Marshallian way.  

```{r coodrFlip, exercise = TRUE}

```

```{r coodrFlip-setup}
library(ggplot2)
a = 0.25
p1 = 1.5
p2 = 1
m = 100 
```

```{r coodrFlip-solution}
ggplot() + 
  geom_function(fun = function(x) a * m / x) + 
  lims(x = c(0, 150)) + 
  labs(x = "Income", y = "Consumption of Good 1") + 
  coord_flip()
```

```{r coodrFlip-check}
grade_this({
  if (uses_geoms(.result, "function")) {
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

### Plotting Demand by Connecting the Dots

Plot the three points you derived in the previous exercise for the prices of good 1 equal to 1.5 and 2.0. (Orient quantity on the x-axis ranging from 0 to 60 and price on the y-axis ranging from 0 to 2.5)

The solutions (for $p_1 = 1.5$ and $p_1 = 2$, respectively) are stored in the setup for the exercise as `eq` and `dpEq`, respectively. All you need to do is plot the points for price and quantity of good 1 using `geom_point()`. 

```{r demandPoints, exercise = TRUE}

```

```{r demandPoints-setup}
library(MicroWithR)
library(ggplot2)
ux <- function(x)
  x[1] ^ (a) * x[2] ^ (1 - a)
bc <- function(x)
  p1 * x[1] + p2 * x[2]
a = 0.25
p1 = 1.5
p2 = 1
m = 100 
eq <-
  solnp2(
    c(1, 1),
    ux,
    ineqfun = bc,
    ineqLB = 0,
    ineqUB = m,
    LB = c(0, 0),
    UB = c(Inf, Inf), 
    max = TRUE
  )
dp1 = 0.5
bc2 <- function(x)
  (p1 + dp1) * x[1] + p2 * x[2]
dpEq <-
  solnp2(
    c(1, 1),
    ux,
    ineqfun = bc2,
    ineqLB = 0,
    ineqUB = m,
    LB = c(0, 0),
    UB = c(Inf, Inf), 
    control = list(trace = 0),
    max = TRUE
  )
```

```{r demandPoints-solution}
ggplot() + 
  geom_point(aes(x = eq$pars[1], y = p1)) + 
  geom_point(aes(x = dpEq$pars[1], y = p1 + dp1)) + 
  lims(x = c(0, 20), y = c(0,2.5)) + 
  labs(x = "Quantity of Good 1", y = "Price of Good 1")
```

```{r demandPoints-check}
grade_this({
  if (uses_geoms(.result, "point")) {
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

Connect the dots with line segments. 

```{r demandSegments, exercise = TRUE}
ggplot() + 
  geom_point(aes(x = eq$pars[1], y = p1)) + 
  geom_point(aes(x = dpEq$pars[1], y = p1 + dp1)) + 
  lims(x = c(0, 20), y = c(0,2.5)) + 
  labs(x = "Quantity of Good 1", y = "Price of Good 1") 
  # Add a layer to connect the points here (don't forget to add "+" to the previous line).

```

```{r demandSegments-setup}
library(MicroWithR)
library(ggplot2)
ux <- function(x)
  x[1] ^ (a) * x[2] ^ (1 - a)
bc <- function(x)
  p1 * x[1] + p2 * x[2]
a = 0.25
p1 = 1.5
p2 = 1
m = 100 
eq <-
  solnp2(
    c(1, 1),
    ux,
    ineqfun = bc,
    ineqLB = 0,
    ineqUB = m,
    LB = c(0, 0),
    UB = c(Inf, Inf), 
    max = TRUE
  )
dp1 = 0.5
bc2 <- function(x)
  (p1 + dp1) * x[1] + p2 * x[2]
dpEq <-
  solnp2(
    c(1, 1),
    ux,
    ineqfun = bc2,
    ineqLB = 0,
    ineqUB = m,
    LB = c(0, 0),
    control = list(trace = 0),
    UB = c(Inf, Inf), 
    max = TRUE
  )
```

```{r demandSegments-solution}
ggplot() + 
  geom_point(aes(x = eq$pars[1], y = p1)) + 
  geom_point(aes(x = dpEq$pars[1], y = p1 + dp1)) + 
  lims(x = c(0, 20), y = c(0,2.5)) + 
  labs(x = "Quantity of Good 1", y = "Price of Good 1") + 
  geom_segment(aes(x = eq$pars[1], xend = dpEq$pars[1], y = p1, yend = p1 + dp1))
```

```{r demandSegments-check}
grade_this({
  if (uses_geoms(.result, "segment")) {
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

### Connecting Many Dots on a Demand Curve

The more points we do this for, the smoother and more complete our
demand curve becomes. Run the code below to derive the demand schedule
for $x_1$ using the parameter values in our original example (m = 100,
p2 = 1, $\alpha$ = 0.25) for the prices ranging from 0.5 to 2.5 in
increments of 0.1. 

1. Define `p1seq` as the sequence of prices and `c1` as a `NULL` object. 
2. Create a `for` loop to iterate the constrained optimization function over all of the values in the length of `p1` using the following steps:
a. Defines the initial values of the choice variables, `pars` to be equal to `c(1,1)` if `i == 1` and otherwise equal to the value of the solution from the previous iteration (which we call `c` below); 
b. Define `c` as the value of the `pars` result from the current iteration solution. 
c. Define the $i^th$ element of `c1` as the first element of the `pars` variable (the optimal level of consumption for that price) of each solution.
3. Plot the demand curve of p1 (y-axis) versus c1 (x-axis).

```{r demandCurve, echo = TRUE}
library(MicroWithR)
ux <- function(x)
  x[1] ^ (a) * x[2] ^ (1 - a)
bc <- function(x)
  p1 * x[1] + p2 * x[2]
a = 0.25
p1 = 1.5
p2 = 1
m = 100 
p1seq <- seq(0.5, 2.5, 0.1)
c1 <- NULL
for(i in 1:length(p1seq)) {
    pars <- c(1,1)
  c <- solnp2(
    pars = pars,
    ux,
    ineqfun = function(x) p1seq[i] * x[1] + p2 * x[2],
    ineqLB = 0,
    ineqUB = m,
    LB = c(0, 0),
    UB = c(Inf, Inf), 
    control = list(trace = 0), 
    max = TRUE
  )$pars
  c1[i] <- c[1]
}
ggplot() + 
  geom_line(aes(x = c1, y = p1seq)) +
  lims(x = c(0, max(c1)), y = c(0, max(p1seq))) +
  labs(x = "Quantity of Good 1", y = "Price of Good 1")
```

### Plotting a Demand Function

Recall that the analytical solution for the demand curve was:

```{r demand, echo = TRUE}
library(Ryacas)
u <- ysym('x1^(alpha) * x2^(1-alpha)')
L <- u + ysym('lambda*(m - p1*x1 - p2*x2)')
dL <- deriv(L, c('x1', 'x2', 'lambda'))
MU <- c(
  p1*solve(dL[1], 'lambda'), 
  p2*solve(dL[2], 'lambda')) |>
  y_rmvars()
MRS <- with_value(MU[1]/MU[2], 'p1', p1) |>
  with_value('p2', p2) |> yac_str() |>
  Deriv::Simplify() |> ysym()
D <- solve(c(MRS, dL[[3]]), c(ysym('p1/p2'), 0), c('x1', 'x2')) |> 
  y_rmvars() |>
  simplify()
D[1] <- with_value(D[1], 'x2', D[2]) |> 
  simplify() |> suppressWarnings()
```

$$x_1(p_1, m) = `r tex(D[1])`$$

1.  Plot this function. Note that this function expresses *quantity* as
    a function of *price*, which implies that the axes are reversed.
2.  Rearrange the demand function to get *price* as a function of
    *quantity* in order to plot the *inverse* demand function (with
    price on the y-axis as usual).

```{r demandFunction, exercise = TRUE}

```

```{r demandFunction-setup}
library(ggplot2)
a = 0.25
p1 = 1.5
p2 = 1
m = 100 
```

```{r demandFunction-solution}
ggplot() + 
  geom_function(fun = function(x) a * m / x) +
  lims(x = c(0, 2.5), y = c(0,50)) + 
  labs(x = "Price of Good 1", y = "Quantity of Good 1")
ggplot() + 
  geom_function(fun = function(x) a * m / x) +
  lims(x = c(0, 2.5), y = c(0,50)) + 
  labs(x = "Price of Good 1", y = "Quantity of Good 1") + 
  coord_flip()
```

```{r demandFunction-check}
grade_this({
  if (uses_geoms(.result, "function")) {
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

## Income and Substitution Effects

The *substitution effect* arises because the change in the price of one good encourages a consumer to buy more of the relatively cheaper good (and less of the relatively more expensive good) in order to keep the same level of satisfaction.  

Analytically, the substitution effect is the difference between (1) the quantity that minimizes the consumer's expenditures *subject to leaving the the consumer at least as well-off as the original equilibrium* ($\tilde{x}$); and (2) the original quantity consumed ($x^*$).  

The *income effect* arises because when prices change, the same amount of money income can now buy a greater or lesser amount of goods and services, overall.  

Analytically, the income effect is the difference between (1) $\tilde{x}$; and (2) the new quantity consumed ($x^**$)

### Finding $\tilde{x}$. 

To separate the income and substitution, the first order of business is finding $\tilde{x}$. Run the following code to use the `Rsolnp::solnp()` function to solve the *nonlinearly*-constrained optimization problem for finding $\tilde{x}$, 
$$min_x \bigg\{p_1x_1 + p_2x_2 \bigg\}$$
$$\text{s.t. } u_0^* - u(x_1, x_2) \le 0$$

```{r xTilde, echo = TRUE}
library(Rsolnp)
library(MicroWithR)
ux <- function(x)
  x[1] ^ (a) * x[2] ^ (1 - a)
bc <- function(x)
  p1 * x[1] + p2 * x[2]
a = 0.25
p1 = 1.5
p2 = 1
m = 100 
eq <-
  solnp2(
    c(1, 1),
    ux,
    ineqfun = bc,
    ineqLB = 0,
    ineqUB = m,
    LB = c(0, 0),
    UB = c(Inf, Inf), 
    max = TRUE
  )
dp1 = 0.5
bc2 <- function(x)
  (p1 + dp1) * x[1] + p2 * x[2]
xTilde <-
  solnp2(
    pars = eq$pars,
    fun = bc2,
    ineqfun = ux,
    ineqLB = ux(eq$pars),
    ineqUB = Inf,
    LB = c(0, 0),
    UB = c(Inf, Inf), 
    max = FALSE
  )
xTilde$pars
```

### Graphical Decomposition

```{r}
dpEq <-
  solnp2(
    c(1, 1),
    ux,
    ineqfun = bc2,
    ineqLB = 0,
    ineqUB = m,
    LB = c(0, 0),
    control = list(trace = 0),
    UB = c(Inf, Inf), 
    max = TRUE
  )
```

```{r substitutionEffect, echo = TRUE, fig.height = 8, fig.width = 8}
subEffect <- ggplot() + 
  geom_function(
    fun = function(x) ux(eq$pars) ^ (1 / (1 - a)) / (x ^ (a / (1 - a))), 
    color = 'darkgreen') + 
  geom_function(
    fun = function(x) m/p2 - (p1/p2)*x, 
    color = 'red') + 
  geom_function(
    fun = function(x) bc2(xTilde$pars)/p2 - ((p1 + dp1)/p2)*x, 
    color = 'red', linetype = 'dashed') + 
  geom_point(aes(x = xTilde$pars[1], y = xTilde$pars[2])) +
  geom_point(aes(x = eq$pars[1], y = eq$pars[2]), ) +
  geom_text(aes(
    x = xTilde$pars[1], y = xTilde$pars[2], 
    label = paste0("(", round(xTilde$pars[1], 2), ", ", round(xTilde$pars[2], 2), ")")), 
    hjust = 0, vjust = 0) +
  geom_text(aes(
    x = eq$pars[1], y = eq$pars[2], 
    label = paste0("(", round(eq$pars[1], 2), ", ", round(eq$pars[2], 2), ")")), 
    hjust = 0, vjust = 0) +
  lims(x = c(0, m/p1), y = c(0, m/p2)) + 
  labs(x = "Consumption of Good 1", y = "Consumption of Good 2")
subEffect
```

```{r incomeEffect, fig.height = 8, fig.width = 8}
incEffect <- subEffect + 
  geom_function(
    fun = function(x) ux(dpEq$pars) ^ (1 / (1 - a)) / (x ^ (a / (1 - a))), 
    color = 'darkgreen', linetype = 'dashed') + 
  geom_function(
    fun = function(x) m/p2 - ((p1 + dp1)/p2)*x, 
    color = 'red', linetype = 'dashed') + 
  geom_point(aes(x = dpEq$pars[1], y = dpEq$pars[2]), ) +
  geom_text(aes(
    x = dpEq$pars[1], y = dpEq$pars[2], 
    label = paste0("(", round(dpEq$pars[1], 2), ", ", round(dpEq$pars[2], 2), ")")), 
    hjust = -1, vjust = 0)
incEffect
```

### Changes in Consumer Surplus

The two dashed expenditure lines (through $x^**$ and $\tilde{x}$) capture the difference in income needed to achieve the new and original maximum levels of utility, respectively, calculated at the new price levels. In other words this is the additional income the consumer "feels" by moving from the original indifference curve to the new one. It is also one of the ways to capture the change in consumer surplus across all markets when relative prices change. 

```{r sC, echo=FALSE}
question("What is the approximate dollar-denominated welfare effect (i.e. income effect) of the price change (evaluated at the new prices) for our example?",
    answer("zero"),
    answer("-9.64"),
    answer("9.64", correct = TRUE),
    answer("90.36"),
    allow_retry = TRUE
  )
```

### Calculating Consumer Surplus

Once you've found $\tilde{x}$, finding the consumer surplus is relatively straightforward. In our case, the expenditures needed at the new prices in order to reach the original utility is the value of the expenditure function at $\tilde{x}$, which equals the budget constraintafter the price change evaluated at `xTilde$pars`, or `r round(bc2(xTilde$pars), 2)`; the expenditures needed to reach the new optimal utility is simply the consumer's money income, `m`. The change in consumer surplus will be the difference `r round(m - bc2(xTilde$pars), 2)`. 

Some policy analyses use this to evaluate the welfare effects of policy changes in terms of real income. The trick is estimating (with as few assumptions as possible!) the unobserved $\tilde{x}$. 
