---
title: "3.1 Market Demand in Competitive Markets - Practice"
output: 
  learnr::tutorial: 
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
author: 
  name: "Created by [Jim Bang](http://www.github.com/bangecon)"
  email: BangJamesT@sau.edu
  affiliation: "[St. Ambrose University](http://www.sau.edu)"
date: "`r format(Sys.Date(), '%d %b %Y')`"
---

<style type="text/css">
h1{font-size: 24pt}
h2{font-size: 20pt}
h3{font-size: 18pt}
h4,h5,h6{font-size: 16pt}
body{font-size: 16pt}
#doc-metadata {
  margin-bottom: 10px;
  font-size: 0.66em;
}
</style>

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(ggcheck)
gradethis_setup(exercise.reveal_solution = TRUE, exercise.eval = TRUE)
knitr::opts_chunk$set(echo = FALSE, fig.height = 6, fig.width = 6)
```

## Deriving Market Demand

The example from the tutorial used a Cobb-Douglas utility function with budget shares 0.25 and 0.75 on goods 1 and 2, respectively; prices 0.75 and 1.0; and income equal to 100. 

$$\max_x \bigg\{ U(x_1, x_2) = x_1^{0.25}x_2^{0.75} \bigg\}$$

$$\text{s.t.: }0.75x_1+x_2 \le 100$$

### Individual Consumer Demand

Replicate the approach I demonstrated in the tutorial to get the demand for *good 2* at prices from 0.25 to 2.0 in increments of 0.05. 

```{r individualDemand, exercise = TRUE}

```

```{r individualDemand-setup}

```

```{r individualDemand-solution}
library(MicroWithR)
a = 0.25
p1 = 1.5 
p2 = seq(0.25, 2.0, 0.05)
m = 30    
ux <- function(x)
  x[1] ^ a * x[2] ^ (1 - a)
bc <- function(x)
  p1 * x[1] + p2[i] * x[2] 
c2 <- NULL
for (i in 1:length(p2)) {
  if(i == 1) {
    pars <- c(1,1)
  } else {
    pars <- c
  }
  c <- solnp2(
    pars = pars,
    ux,
    ineqfun = bc,
    ineqLB = 0,
    ineqUB = m,
    LB = c(0, 0),
    UB = c(Inf, Inf), 
    control = list(trace = 0), 
    max = TRUE
  )$pars
  c2[i] <- c[2]
}
head(c2)
```

```{r individualDemand-check}
grade_this({if (.result[1:2] != c(90, 75)) { 
    fail("The utility-maximizing bundle is 5 units of good 1 and 22.5 units of good 2.") }
  pass()
})
```

Plot the graph of the consumer's demand for good 2 (stored as `c2` in the setup) for prices between 0.25 and 2.0. 

```{r individualDemandCurve, exercise = TRUE}

```

```{r individualDemandCurve-setup}
library(MicroWithR)
a = 0.25
p1 = 1.5 
p2 = seq(0.25, 2.0, 0.05)
m = 30    
ux <- function(x)
  x[1] ^ a * x[2] ^ (1 - a)
bc <- function(x)
  p1 * x[1] + p2[i] * x[2] 
c2 <- NULL
for (i in 1:length(p2)) {
  if(i == 1) {
    pars <- c(1,1)
  } else {
    pars <- c
  }
  c <- solnp2(
    pars = pars,
    ux,
    ineqfun = bc,
    ineqLB = 0,
    ineqUB = m,
    LB = c(0, 0),
    UB = c(Inf, Inf), 
    control = list(trace = 0), 
    max = TRUE
  )$pars
  c2[i] <- c[2]
}
```

```{r individualDemandCurve-hint}
# In the computational solution we defined numerical vectors containing the values for the prices of good 2 ranging from 0.25 to 2 (`p2`) and the corresponding consumption levels of good 2 (`c2`). All you need to do is plot the connection of these points using `geom_line`. 
```

```{r individualDemandCurve-solution}
library(ggplot2)
ggplot() + 
  geom_line(aes(x = c2, y = p2, color = "D")) + 
  lims(x = c(0, 75),y = c(0, 2.0)) + 
  labs(x = "Quantity", y = "Price", color = "") + 
  scale_color_manual(breaks = c("D"), values = c("blue"))
```

```{r individualDemandCurve-check}
grade_this({
  if (uses_geoms(.result, "line")) {
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

The analytical solution for the demand curve was:

<details><summary>
Click here to see the code for deriving the demand function using `Ryacas` functions.
</summary>
```{r demand, echo = TRUE, message = FALSE, warning = FALSE}
library(Ryacas)
library(Deriv)
u <- ysym('x1^(alpha) * x2^(1-alpha)')
L <- u + ysym('lambda*(m - p1*x1 - p2*x2)')
dL <- deriv(L, c('x1', 'x2', 'lambda'))
MU <- deriv(u, c('x1', 'x2'))
MRS <- yac_str(MU[1]/MU[2]) |> 
  Simplify() |> 
  ysym()
D <- solve(c(MRS, dL[[3]]), c(ysym('p1/p2'), 0), c('x1', 'x2')) |> 
  y_rmvars() |>
  simplify()
D[1] <- with_value(D[1], 'x2', D[2]) |> 
  simplify()
```
</details>

$$x_2(p_2, \alpha, m) = `r tex(D[2])`$$

### Many (Identical) Consumers

Plot the market demand curve for *good 2* for a market with 200,000 consumers who are all identical to the one for whom we solved the equilibrium in the previous exercises. 

```{r marketDemand, exercise = TRUE}
```

```{r marketDemand-setup}
library(ggplot2)
```

```{r marketDemand-hint}
# First you will need to add up the individual demands, but then you will need to invert the demand curve to get price as the "dependent" (y) variable.
# Never mind the fact that price isn't really the dependent variable that the consumer chooses. Blame Marshall!
```

```{r marketDemand-solution}
ggplot() + 
  geom_function(fun = function(x) 4.5e6/x, aes(color = "D")) + 
  lims(x = c(0,9e6), y = c(0, 2)) +
  labs(x = "Quantity", y = "Price", color = "") + 
  scale_color_manual(breaks = c("D"), values = c("blue"))
```

```{r marketDemand-check}
grade_this({
  if (uses_geoms(.result, "function") | uses_geoms(.result, "line")) {
    pass(random_praise())
  }
  fail(random_encouragement())
})
```
