---
title: "1.7 Other Applications"
editor_options: 
  markdown: 
    wrap: 72
output: 
  learnr::tutorial: 
    progressive: true
runtime: shiny_prerendered
author: 
  name: "Created by [Jim Bang](http://www.github.com/bangecon)"
  email: BangJamesT@sau.edu
  affiliation: "[St. Ambrose University](http://www.sau.edu)"
date: "`r format(Sys.Date(), '%d %b %Y')`"
---

<style type="text/css">
h1{font-size: 24pt}
h2{font-size: 20pt}
h3{font-size: 18pt}
h4,h5,h6{font-size: 16pt}
body{font-size: 16pt}
 # doc-metadata {
  margin-bottom: 10px;
  font-size: 0.66em;
}
</style>

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
gradethis_setup()
tutorial_options(exercise.eval = TRUE, exercise.reveal_solution = FALSE)
knitr::opts_chunk$set(echo = FALSE, fig.width = 6, fig.height = 8,
  fig.align='center', out.width = '75%')
```

## Overview

There are many, many specific applications for consumer choice. 
We have already modeled a few using our utility maximization framework: 

1. Insurance and Risk  
2. Excise Taxes  
3. In-Kind versus Cash Transfers  

Here, we consider a few more examples: 

4. Inter-Temporal Choice  
5. Labor-Leisure Tradeoffs 

## Intertemporal Choice

### Overview

Intertemporal choice involves choosing between consuming *now* or consuming *later*. In that respect, it resembles the standard consumer between different *goods*, except that the "different goods" differ in terms of the state - the present versus the future - during which the consumer consumes them. This application illustrates the intuition that (1) consumers tend to prefer current consumption over consuming in the future; and (2) variety-loving consumers will choose a "smooth" consumption path that tries to balance current and future consumption. 

### Budget Constraint

The budget constraint requires that the present-value of consumption be less than or equal to the present-value of income (at market interest rates). 

$$c_1 + \frac{c_2}{1+r} \le m_1 + \frac{m_2}{1+r}$$

In other words, the second-period consumption must be less than or equal to second-period income plus the future value of *net* savings from the first period. 

$$c_2 \le m_2 + (1+r)(m_1-c_1)$$

Graph the budget constraint for a consumer who receives an income of 80 in period 1 and 20 in period 2 at a market interest rate of 20%. Position consumption for period 1 on the x-axis. Use the intercepts of the budget line as your x and y limits. 

```{r intertemporalBC, exercise = TRUE}

```

```{r intertemporalBC-solution}
m1 = 80
m2 = 20
r = 0.20
ggplot() + 
  geom_function(fun = function(x)   m2 + (1 + r) * (m1 - x), color = 'red') +
  lims(x = c(0, m1 + m2/(1+r)), y = c(0, m2 + (1+r)*m1)) + 
  labs(x = "Period 1 Consumption", y = "Period 2 Consumption")
```

```{r intertemporalBC-check}
grade_this({if (!inherits(.result, c("list"))) {
    fail("Your class of your answer should be a list")}
  if (.result$pars != c(17.5, 78.75)) { 
    fail("The optimal amount of goods one and two after the change should be 17.5 and 78.75") }
  pass()
})
```

### Preferences

Most intertemporal choice problems model utility as the sum of the (diminishing-returns) utilities from consumption in each time period. At the same time, consumers *subjectively* discount future consumption. For our example, let's assume a log utility function as given by: 

$$ u(c_1, c_2) = \sum_{t=1}^2 \bigg\{\delta^{t-1}u(c_t)\bigg\} = ln(c_1) + \delta ln(c_2)$$

Add the the indifference curves for the intertemporal preferences for total utility levels of 6 and 7. Assume a subjective discount rate of $\beta = 2/3$. Plot the graph over the values of the intercepts of the budget constraint above. 

```{r intertemporalPreferences, exercise = TRUE, exercise.reveal_solution = TRUE}

```

```{r intertemporalPreferences-setup}
m1 = 80
m2 = 20
r = 0.20
```

```{r intertemporalPreferences-solution}
d = 2/3
ggplot() + 
  geom_function(fun = function(x) exp((6 - log(x)) / d), color = 'darkgreen') + 
  lims(x = c(0, m1 + m2/(1+r)), y = c(0, m2 + (1+r)*m1)) + 
  labs(x = "Consumption in Period 1", y = "Consumption in Period 2")
```

```{r intertemporalPreferences-check}
grade_this({
  if (uses_geoms(.result, "function") & uses_geoms(.result, "point")) {
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

### Equilibrium

Calculate the equilibrium numerically. 

```{r intertemporalEquilibrium, exercise = TRUE, exercise.reveal_solution = TRUE}

```

```{r intertemporalEquilibrium-setup}
m1 = 80
m2 = 20
r = 0.20
d = 2 / 3
```

```{r intertemporalEquilibrium-solution}
library(MicroWithR)
ux <- function(x)
  log(x[1]) + d * log(x[2])
bc <- function(x)
  (1 + r) * x[1] + x[2]
eq <- solnp2(
  c(1, 1),
  ux,
  ineqfun = bc,
  ineqLB = 0,
  ineqUB = (1 + r) * m1 + m2,
  LB = c(0, 0),
  UB = c(Inf, Inf), 
  max = TRUE
)
```

```{r intertemporalEquilibrium-check}
grade_this({if (!inherits(.result, c("list"))) {
    fail("Your class of your answer should be a list")}
  if (round(.result$pars, 1) != c(58, 46.4)) { 
    fail("The optimal amount of goods one and two after the change should be about 58 and 46.4") }
  pass()
})
```

### Intertemporal Graph

Graph the equilibrium using `ggplot`. 

```{r intertemporalGraph, exercise = TRUE}

```

```{r intertemporalGraph-setup}
m1 = 80
m2 = 20
r = 0.20
d = 2 / 3
library(MicroWithR)
ux <- function(x)
  log(x[1]) + d * log(x[2])
bc <- function(x)
  (1 + r) * x[1] + x[2]
eq <- solnp2(
  c(1, 1),
  ux,
  ineqfun = bc,
  ineqLB = 0,
  ineqUB = (1 + r) * m1 + m2,
  LB = c(0, 0),
  UB = c(Inf, Inf), 
  max = TRUE
)
```

```{r intertemporalGraph-solution}
ggplot() + 
  geom_function(fun = function(x) m2 + (1 + r) * (m1 - x), color = 'red') + 
  geom_function(fun = function(x) exp((ux(eq$pars) - log(x))/d), color = 'darkgreen') + 
  geom_point(aes(x = eq$pars[1], y = eq$pars[2])) +
  lims(x = c(0, m2/(1+r) + m1), y = c(0, m2 + (1+r)*m1)) + 
  labs(x = "Consumption in Period 1", y = "Consumption in Period 2")
```

```{r intertemporalGraph-check}
grade_code()
```

### Effect of a Rate Hike

One thing we can see is how much the allocation changes in response to a change in the interest rate *to 200%*. Run the following code to simulate this change.

```{r intertemporalRateHike, exercise = TRUE}

```

```{r intertemporalRateHike-setup}
m1 = 80
m2 = 20
r = 0.20
d = 2 / 3
library(MicroWithR)
ux <- function(x)
  log(x[1]) + d * log(x[2])
bc <- function(x)
  (1 + r) * x[1] + x[2]
eq <- solnp2(
  c(1, 1),
  ux,
  ineqfun = bc,
  ineqLB = 0,
  ineqUB = (1 + r) * m1 + m2,
  LB = c(0, 0),
  UB = c(Inf, Inf)
)
```

```{r intertemporalRateHike-solution}
r = 2
eq2 <- solnp2(
  c(1, 1),
  ux,
  ineqfun = bc,
  ineqLB = 0,
  ineqUB = (1 + r) * m1 + m2,
  LB = c(0, 0),
  UB = c(Inf, Inf),
  max = TRUE
)
```

```{r intertemporalRateHike-check}
grade_this({if (!inherits(.result, c("list"))) {
    fail("Your class of your answer should be a list")}
  if (round(.result$pars, 1) != c(58, 46.4)) { 
    fail("The optimal amount of goods one and two after the change should be about 58 and 46.4") }
  pass()
})
```

### Graphical Effect of a Rate Hike

Add the rate-hike equilibrium to the original graph. The original equilibrium solution, its graph, and the new equilibrium solution are included in the setup as `eq`, `g`, and `eq2` to get you started so that all you need to do is add layers for the new budget line and the new indifference curve through the rate-hike equilibrium.

Note: In order to fit the full budget constraint you might need to change the `lims` layer. 

```{r intertemporalRateHikeGraph, exercise = TRUE}

```

```{r intertemporalRateHikeGraph-setup}
m1 = 80
m2 = 20
r = 0.20
d = 2 / 3
library(MicroWithR)
ux <- function(x)
  log(x[1]) + d * log(x[2])
bc <- function(x)
  (1 + r) * x[1] + x[2]
eq <- solnp2(
  c(1, 1),
  ux,
  ineqfun = bc,
  ineqLB = 0,
  ineqUB = (1 + r) * m1 + m2,
  LB = c(0, 0),
  UB = c(Inf, Inf), 
  max = TRUE
)
r0 = 0.2
g <- ggplot() + 
  geom_function(fun = function(x) m2 + (1 + r0) * (m1 - x), color = 'red') + 
  geom_function(fun = function(x) exp((ux(eq$pars) - log(x))/d), color = 'darkgreen') + 
  geom_point(aes(x = eq$pars[1], y = eq$pars[2])) +
  lims(x = c(0, m2/(1+r0) + m1), y = c(0, m2 + (1+r)*m1)) + 
  labs(x = "Consumption in Period 1", y = "Consumption in Period 2")
r = 2
eq2 <- solnp2(
  c(1, 1),
  ux,
  ineqfun = bc,
  ineqLB = 0,
  ineqUB = (1 + r) * m1 + m2,
  LB = c(0, 0),
  UB = c(Inf, Inf),
  max = TRUE
)
```

```{r intertemporalRateHikeGraph-solution}
r = 2
g + 
  geom_function(fun = function(x) m2 + (1 + r) * (m1 - x), 
                color = 'red', linetype = 'dashed') + 
  geom_function(fun = function(x) exp((ux(eq2$pars) - log(x))/d), 
                color = 'darkgreen', linetype = 'dashed') + 
  geom_point(aes(x = eq2$pars[1], y = eq2$pars[2])) 
```

```{r intertemporalRateHikeGraph-check}
grade_code()
```

Note: In order to keep the original budget line in its old location for the graph, I created a new value, $r0$ equal to the original interest rate *only for the purpose of building the graphs. 

From this we can see that *current* consumption is not all that sensitive to the interest rate. Notice that a "price change" in the form of increasing the interest rate has an effect that is less clear than a simple static price change. The budget line does not simply pivot at the x or y axis because increasing $r$ increases the opportunity *cost* of current consumption *and* the future *value* of current income on *both* sides of the budget constraint: 

$$(1+r)c_1 + c_2 \le (1+r)m_1 + m_2$$

## Labor-Leisure Tradeoffs

Alternatives for describing choice of labor supply:

1. Choose amounts of consumption (a "good") and *labor* (a "bad"); 
2. Choose amounts of consumption and *leisure* (both "goods").

Really, the two are equivalent, but the former is a little cleaner to model since the indifference curves look "normal." 

### Manager Effort Decision

Suppose that a manager derives utility from consuming (1) goods and services out of their income, $y$; and (2) on-the-job leisure ("shirking"), $\ell$ according to the following utility function: 

$$u(y,l) = y^{0.75}\ell^{0.25}$$
Their total on-the-job time is 50 hours per week, 50 weeks per year for a total of 2,500 hours per year. 

With minimal managerial effort ($\ell = 2,500$) the firm's profits reach \$1,000,000. Each hour of managerial effort, $e$, while on the job returns profit to the firm of \$2,500. 

Suppose that the firm pays the manager a fixed salary of \$150,000 per year. What would be the total number of hours of effort versus on-the-job leisure that the manager will choose? 

Suppose now that the firm pays the manager a fixed salary of just \$100,000 with an incentive bonus equal to 4% of the firm's total profits. Solve for the optimal amount of leisure, income, and effort. 

```{r effort, exercise = TRUE}

```

```{r effort-hint}

```

```{r effort-solution}
library(MicroWithR)
ux = function(x)
  x[1] ^ 0.75 * x[2] ^ 0.25
H = 2500
S = 100000
s = 0.04
m = 2500
h = function(x)
  H - x[2]
pi0 = 100000
pi = function(x)
  m * h(x) + pi0
bx <- function(x)
  x[1] + s * m * x[2]
eq <- solnp2(
  pars = c(1, 1),
  fun = ux,
  ineqfun = bx,
  ineqUB = S + s * (m * H + pi0),
  ineqLB = -Inf,
  max = TRUE
)
```

```{r effort-check}
grade_this({if (!inherits(.result, c("list"))) {
    fail("Your class of your answer should be a list")}
  if (round(.result$pars, 1) != c(58, 46.4)) { 
    fail("The optimal amount of goods one and two after the change should be about 58 and 46.4") }
  pass()
})
```

### Shareholders' Incentive Decision. 
