---
title: "1.7 Other Applications: Labor-Leisure Trade-Off - Practice"
editor_options: 
  markdown: 
    wrap: 72
output: 
  learnr::tutorial: 
    progressive: true
runtime: shiny_prerendered
author: 
  name: "Created by [Jim Bang](http://www.github.com/bangecon)"
  email: BangJamesT@sau.edu
  affiliation: "[St. Ambrose University](http://www.sau.edu)"
date: "`r format(Sys.Date(), '%d %b %Y')`"
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
gradethis_setup()
tutorial_options(exercise.eval = TRUE, exercise.reveal_solution = FALSE)
knitr::opts_chunk$set(echo = FALSE, fig.width = 6, fig.height = 8,
  fig.align='center', out.width = '75%')
```

## Overview

There are many, many specific applications for consumer choice. 
We have already modeled a few using our utility maximization framework: 

1. Insurance and Risk  
2. Excise Taxes  
3. In-Kind versus Cash Transfers  
4. Inter-Temporal Choice  

Here, we consider another examples: 

5. Labor-Leisure Tradeoffs 

## Labor-Leisure Tradeoffs

Alternatives for describing choice of labor supply:

1. Choose amounts of consumption (a "good") and *labor* (a "bad"); 
2. Choose amounts of consumption and *leisure* (both "goods").

Really, the two are equivalent, but the former is a little cleaner to model since the indifference curves look "normal." 

### Manager Effort Decision

Suppose that a manager derives utility from consuming (1) goods and services out of their income, $y$; and (2) on-the-job leisure ("shirking"), $\ell$ according to the following utility function: 

$$u(y,l) = y^{0.75}\ell^{0.25}$$
Their total on-the-job time is 50 hours per week, 50 weeks per year for a total of 2,500 hours per year. 

With minimal managerial effort ($\ell = 2,500$) the firm's profits reach \$1,000,000. Each hour of managerial effort, $e$, while on the job returns profit to the firm of \$2,500. 

Suppose that the firm pays the manager a fixed salary of \$150,000 per year. What would be the total number of hours of effort versus on-the-job leisure that the manager will choose? 

Suppose now that the firm pays the manager a fixed salary of just \$100,000 with an incentive bonus equal to 4% of the firm's total profits. Solve for the optimal amount of leisure, income, and effort. 

```{r effort, exercise = TRUE}

```

```{r effort-hint}

```

```{r effort-solution}
library(MicroWithR)
ux = function(x)
  x[1] ^ 0.75 * x[2] ^ 0.25
H = 2500
S = 100000
s = 0.04
m = 2500
h = function(x)
  H - x[2]
pi0 = 1000000
pi = function(x)
  m * h(x) + pi0
bx <- function(x)
  x[1] + s * m * x[2]
eq <- solnp2(
  pars = c(1, 1),
  fun = ux,
  ineqfun = bx,
  ineqUB = S + s * (m * H + pi0),
  ineqLB = -Inf,
  max = TRUE
)
```

```{r effort-check}
grade_this({if (!inherits(.result, c("list"))) {
    fail("Your class of your answer should be a list")}
  if (round(.result$pars, 1) != c(58, 46.4)) { 
    fail("The optimal amount of goods one and two after the change should be about 58 and 46.4") }
  pass()
})
```
