---
title: "1.6 Comparative Statics II: Price Changes Application - Practice"
editor_options: 
  markdown: 
    wrap: 72
output: 
  learnr::tutorial: 
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
author: 
  name: "Created by [Jim Bang](http://www.github.com/bangecon)"
  email: BangJamesT@sau.edu
  affiliation: "[St. Ambrose University](http://www.sau.edu)"
date: "`r format(Sys.Date(), '%d %b %Y')`"
---

<style type="text/css">
#doc-metadata {
  margin-bottom: 10px;
  font-size: 0.8em;
}
</style>

```{r setup, include = FALSE}
library(learnr)
library(gradethis)
gradethis_setup()
tutorial_options(exercise.eval = TRUE, exercise.reveal_solution = FALSE)
knitr::opts_chunk$set(echo = FALSE, fig.align = "center", fig.height = 6, fig.width = 6)
knitr::include_graphics
```

## Alcoholic Beverage (Excise) Tax

### Initial Solution

Recall the beer and donuts example from our budget constraint section.  
Consider a consumer with the income ($m = 20$) and price ($p_b = 2$ and $p_d = 1$) parameters from that example along with preferences described by a Cobb-Douglas utility function with equal budget shares on each good. 

$$\max_x \bigg\{ U(x_b,x_d) = x_b^{0.5}x_d^{0.5} \bigg\}$$
$$\text{s.t. } 2x_b + x_d \le 20$$

Calculate and graph the initial equilibrium 
(with beer on the x-axis and donuts on the y-axis as before). 

```{r taxInitialEquilibrium, exercise = TRUE}

```

```{r taxInitialEquilibrium-solution}
library(MicroWithR)
ux <- function(x)
  x[1] ^ (a) * x[2] ^ (1 - a)
bc <- function(x)
  pb * x[1] + pd * x[2]
a = 0.5
m = 20
pb = 2
pd = 1
control = list(trace = 0)
eq <- solnp2(
  c(1, 1),
  ux,
  ineqfun = bc,
  ineqLB = 0,
  ineqUB = m,
  LB = c(0, 0),
  UB = c(Inf, Inf),
  max = TRUE
)
```

```{r taxInitialEquilibrium-check}
grade_this({if (!inherits(.result, c("list"))) {
    fail("Your class of your answer should be a list")}
  if (.result$pars != c(5, 10)) { 
    fail("The optimal amount of goods one and two should be 5 and 10") }
  pass()
})
```

### Alcoholic-Beverage Tax Solution

Suppose the government levies a $0.50/unit tax on beer. 
Calculate the new equilibrium.  

```{r taxEquilibrium, exercise = TRUE}

```

```{r taxEquilibrium-setup}
library(MicroWithR)
a = 0.5
m = 20
pb = 2
pd = 1
t = 0.5
control = list(trace = 0)
ux <- function(x)
  - x[1] ^ (a) * x[2] ^ (1 - a)
bc <- function(x)
  pb * x[1] + pd * x[2]
```

```{r taxEquilibrium-solution}
t = 0.5
bcT <- function(x)
  (pb + t) * x[1] + pd * x[2]
eqT <- solnp2(
  c(1, 1),
  ux,
  ineqfun = bcT,
  ineqLB = 0,
  ineqUB = m,
  LB = c(0, 0),
  UB = c(Inf, Inf), 
  max = TRUE
)
```

```{r taxEquilibrium-check}
grade_this({if (!inherits(.result, c("list"))) {
    fail("Your class of your answer should be a list")}
  if (.result$pars != c(5, 10)) { 
    fail("The optimal amount of goods one and two should be 4 and 10") }
  pass()
})
```

### Alcoholic-Beverage Tax Graph

Graph the equilibrium (indifference curve and budget line) before and after the tax. 
Include a point for the equilbrim labeled with the equilibrium values $(Q_beer, Q_Donuts)$. 
Place beer on the x-axis (labeled "Beer") and donuts on the y-axis (labeled "Donuts). 

```{r taxGraph, exercise = TRUE}

```

```{r taxGraph-setup}
library(MicroWithR)
a = 0.5
m = 20
pb = 2
pd = 1
t = 0.5
control = list(trace = 0)
bc <- function(x)
  pb * x[1] + pd * x[2]
ux <- function(x)
  - x[1] ^ (a) * x[2] ^ (1 - a)
eq <-
  solnp2(
    c(1, 1),
    ux,
    ineqfun = bc,
    ineqLB = 0,
    ineqUB = m,
    LB = c(0, 0),
    UB = c(Inf, Inf),
    max = TRUE
  )
bcT <- function(x)
  (pb + t) * x[1] + pd * x[2]
eqT <- solnp2(
  c(1, 1),
  ux,
  ineqfun = bcT,
  ineqLB = 0,
  ineqUB = m,
  LB = c(0, 0),
  UB = c(Inf, Inf),
  max = TRUE
)
```

```{r taxGraph-solution}
library(ggplot2)
ggplot() + 
  geom_function(
    fun = function(x) ux(eq$pars) ^ (1 / (1 - a)) / (x ^ (a / (1 - a))), 
    color = 'darkgreen') +
  geom_function(fun = function(x) m / pd - pb * x / pd, color = 'red') +
  geom_function(
    fun = function(x) ux(eqT$pars) ^ (1 / (1 - a)) / (x ^ (a / (1 - a))), 
    color = 'darkgreen', linetype = 'dashed') +
  geom_function(
    fun = function(x) m / pd - (pb + t) * x / pd, 
    color = 'red', linetype = 'dashed') +
  geom_point(aes(x = eq$pars[1], y = eq$pars[2]), color = "blue") +
  geom_point(aes(x = eqT$pars[1], y = eqT$pars[2]), color = "blue") +
  geom_text(
    aes(x = eq$pars[1], y = eq$pars[2]), 
    label = paste0('(', round(eq$pars[1], 2), ', ', round(eq$pars[2], 2), ')'),
    hjust = 0, vjust = 0) + 
  geom_text(
    aes(x = eqT$pars[1], y = eqT$pars[2]), 
    label = paste0('(', round(eqT$pars[1], 2), ', ', round(eqT$pars[2], 2), ')'), 
    hjust = 0, vjust = 0) + 
  lims(x = c(0, m/pb), y = c(0, m/pd)) + 
  labs(x = "Quantity of Beer", y = "Quantity of Donuts")
```

```{r taxGraph-check}
grade_this({
  if (uses_geoms(.result, "function") & uses_geoms(.result, "point")) {
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

Note that the tax reduces beer consumption and consumer utility. 
However, since the budget share is constant, the tax does *not* affect donut consumption in this example. A consumer with different preferences would respond differently.

