---
title: "2.3 Cost Curves"
output: 
  learnr::tutorial: 
    progressive: true
runtime: shiny_prerendered
author: 
  name: "Created by [Jim Bang](http://www.github.com/bangecon)"
  email: BangJamesT@sau.edu
  affiliation: "[St. Ambrose University](http://www.sau.edu)"
date: "`r format(Sys.Date(), '%d %b %Y')`"
---

<style type="text/css">
#doc-metadata {
  margin-bottom: 10px;
  font-size: 0.8em;
}
</style>

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
gradethis_setup(exercise.reveal_solution = FALSE, exercise.eval = TRUE)
knitr::opts_chunk$set(echo = FALSE, fig.height = 6, fig.width = 6)
```

## Constant Marginal Productivity 

1. Compute the values of the cost-minimizing combinations of labor and capital to produce at least ten different quantities of output using a production function where $MP_L = MP_K = 100$ for all levels of the inputs $L$ and $K$ where $w = 1$ and $r = 2$. 

Plot the long-run total and long-run average cost curves. 

```{r constantMPLs, exercise = TRUE}

```

```{r constantMPLs-solution}
library(Ryacas)
library(Deriv)
library(Rsolnp)
w = 1; r = 2
costSchedule <-
  data.frame(
    Quantity = seq(from = 0, to = 5000, by = 100),
    Labor = NA,
    Capital = NA,
    Total_Cost = NA,
    Average_Cost = NA
  )
for (i in 1:nrow(costSchedule)) {
  costMinTemp <- solnp(
    pars = c(1, 1), 
    fun = function(x) w*x[1] + r*x[2],
    ineqfun = function(x) 100*(x[1] + x[2]),
    ineqLB = costSchedule$Quantity[i], 
    ineqUB = Inf, 
    LB = c(0, 0))
  costSchedule$Labor[i] <- costMinTemp$pars[1]
  costSchedule$Capital[i] <- costMinTemp$pars[2]
  costSchedule$Total_Cost[i] <- tail(costMinTemp$values, 1)
}
costSchedule$Average_Cost <- costSchedule$Total_Cost / costSchedule$Quantity
par(mfrow = c(1, 2))
plot(costSchedule$Quantity, costSchedule$Total_Cost, type = 'l', ylim = c(0,50), xlab = "Quantity", ylab = "Total Cost")
plot(costSchedule$Quantity, costSchedule$Average_Cost, type = 'l', ylim = c(0,0.02), xlab = "Quantity", ylab = "Average Cost")
```

```{r constantMPLs-check}
grade_code()
```

2. Fix $K = 10$ in the previous question. Compute the short-run cost-minimizing input combinations for at least 10 different quantities of output. 

Plot the total and total variable cost curves on one graph and the marginal, average, and average variable cost curves on another. 

```{r SRconstantMPLs, exercise = TRUE}

```

```{r SRconstantMPLs-solution}
SRcostSchedule <-
  data.frame(
    Quantity = seq(from = 0, to = 5000, by = 20),
    Labor = NA,
    Capital = 10,
    Total_Cost = NA,
    Average_Cost = NA
  )
for (i in 1:nrow(SRcostSchedule)) {
  costMinTemp <- solnp(
    pars = 1, 
    fun = function(x) w*x + r*10,
    ineqfun = function(x) 100*(x + 10),
    ineqLB = SRcostSchedule$Quantity[i], 
    ineqUB = Inf, 
    LB = 0)
  SRcostSchedule$Labor[i] <- costMinTemp$pars[1]
  SRcostSchedule$Variable_Cost[i] <- w*SRcostSchedule$Labor[i]
  SRcostSchedule$Total_Cost[i] <- tail(costMinTemp$values, 1)
}
SRcostSchedule$Average_Cost <- SRcostSchedule$Total_Cost / SRcostSchedule$Quantity
SRcostSchedule$Average_Variable_Cost <- 
  SRcostSchedule$Variable_Cost / SRcostSchedule$Quantity
SRcostSchedule$Marginal_Cost <- 
  c(NA,diff(SRcostSchedule$Total_Cost, lag = 2)/40, NA)
par(mfrow = c(1, 2))
plot(SRcostSchedule$Quantity, SRcostSchedule$Total_Cost, type = 'l', ylim = c(0,50), xlab = "Quantity", ylab = "Total Cost")
lines(SRcostSchedule$Quantity, SRcostSchedule$Variable_Cost, col = 'darkorange')
plot(SRcostSchedule$Quantity, SRcostSchedule$Average_Cost, type = 'l', ylim = c(0,0.05), xlab = "Quantity", ylab = "Average Cost")
lines(SRcostSchedule$Quantity, SRcostSchedule$Average_Variable_Cost, col = 'darkorange')
lines(SRcostSchedule$Quantity, SRcostSchedule$Marginal_Cost, col = 'red')
```

```{r SRconstantMPLs-check}
grade_code()
```

