---
title: "1.5 Comparative Statics"
editor_options: 
  markdown: 
    wrap: 72
output: 
  learnr::tutorial: 
    progressive: true
runtime: shiny_prerendered
author: 
  name: "Created by [Jim Bang](http://www.github.com/bangecon)"
  email: BangJamesT@sau.edu
  affiliation: "[St. Ambrose University](http://www.sau.edu)"
date: "`r format(Sys.Date(), '%d %b %Y')`"
---

<style type="text/css">
#doc-metadata {
  margin-bottom: 10px;
  font-size: 0.8em;
}
</style>

```{r setup, include = FALSE}
library(learnr)
library(gradethis)
library(Ryacas)
library(Deriv)
library(Rsolnp)
gradethis_setup()
tutorial_options(exercise.eval = TRUE, exercise.reveal_solution = FALSE)
knitr::opts_chunk$set(echo = FALSE, fig.align = "center", fig.height = 6, fig.width = 6)
knitr::include_graphics
```

## Quasi-Linear Utility

Consider a consumer who initially budgets an income of 100 for consuming goods 1 and 2; faces prices of good 1 and good 2 equal to 1.5 and 1.0, respectively; and exhibits quasi-linear preferences described by the following utility function:  

$$U(x) = x_1^{0.25} + x_2,$$
$$\text{s.t.: }1.5x_1+x_2 \le 100$$

1. Plot the Engel Curve for this consumer for good 1 by computationally solving and connecting at least 5 points of the curve. You may choose any changes in income you like; the "solution" uses $\pm 5$ and $\pm 10$.

```{r engelComputational, exercise = TRUE}

```

```{r engelComputational-solution}
ux <- function(x) -x[1] ^ (a) + x[2] 
bc <- function(x) p1 * x[1] + p2 * x[2]
a = 0.25; p1 = 1.5; p2 = 1; m = 100; dm = 5
initialEquilibrium <- solnp(c(1, 1), ux, ineqfun = bc, ineqLB = 0, ineqUB = m, 
                            LB = c(0, 0), UB = c(Inf, Inf))
dmEquilibrium1 <- solnp(c(1, 1), ux, ineqfun = bc, ineqLB = 0, ineqUB = m + dm, 
                            LB = c(0, 0), UB = c(Inf, Inf))
dmEquilibrium2 <- solnp(c(1, 1), ux, ineqfun = bc, ineqLB = 0, ineqUB = m - dm, 
                            LB = c(0, 0), UB = c(Inf, Inf))
dmEquilibrium3 <- solnp(c(1, 1), ux, ineqfun = bc, ineqLB = 0, ineqUB = m + 2*dm, 
                            LB = c(0, 0), UB = c(Inf, Inf))
dmEquilibrium4 <- solnp(c(1, 1), ux, ineqfun = bc, ineqLB = 0, ineqUB = m - 2*dm, 
                            LB = c(0, 0), UB = c(Inf, Inf))
x1 <- c(initialEquilibrium$pars[1], dmEquilibrium1$pars[1], dmEquilibrium2$pars[1],
        dmEquilibrium3$pars[1], dmEquilibrium4$pars[1])
m1 <- c(initialEquilibrium$ineqx0, dmEquilibrium1$ineqx0, dmEquilibrium2$ineqx0,
        dmEquilibrium3$ineqx0, dmEquilibrium4$ineqx0)
plot(m1, x1, xlim = c(0,120), ylim = c(0, 100))
segments(m1[1], x1[1], m1[2], x1[2])
segments(m1[2], x1[2], m1[3], x1[3])
segments(m1[3], x1[3], m1[4], x1[4])
segments(m1[4], x1[4], m1[5], x1[5])
```

```{r engelComputational-check}

```

2. Plot the Demand Function for $x_1$. for this consumer for good 1 by computationally solving and connecting at least 5 points of the curve. You may choose any changes in price you like; the "solution" uses $\pm 0.5$ and $\pm 1.0$.

```{r demandComputational, exercise = TRUE}

```

```{r demandComputational-solution}
dp1 = 0.5
bcLowP1  <- function(x) (p1 - dp1) * x[1] + p2 * x[2]
bcHighP1 <- function(x) (p1 + dp1) * x[1] + p2 * x[2]
bcLowP2  <- function(x) (p1 - 2*dp1) * x[1] + p2 * x[2]
bcHighP2 <- function(x) (p1 + 2*dp1) * x[1] + p2 * x[2]
lowPEquilibrium1 <- solnp(c(1, 1), ux, ineqfun = bcLowP1, ineqLB = 0, ineqUB = m,
    LB = c(0, 0), UB = c(Inf, Inf))
highPEquilibrium1 <- solnp(c(1, 1), ux, ineqfun = bcHighP1, ineqLB = 0, ineqUB = m,
    LB = c(0, 0), UB = c(Inf, Inf))
lowPEquilibrium2 <- solnp(c(1, 1), ux, ineqfun = bcLowP2, ineqLB = 0, ineqUB = m,
    LB = c(0, 0), UB = c(Inf, Inf))
highPEquilibrium2 <- solnp(c(1, 1), ux, ineqfun = bcHighP2, ineqLB = 0, ineqUB = m,
    LB = c(0, 0), UB = c(Inf, Inf))
p1 <- c(p1-2*dp1, p1-dp1, p1, p1+dp1, p1+2*dp1)
x1 <- c(lowPEquilibrium2$pars[1], lowPEquilibrium1$pars[1], 
        initialEquilibrium$pars[1], highPEquilibrium1$pars[1], 
        highPEquilibrium2$pars[1])
plot(x1, p1, xlim = c(0, 200), ylim = c(0, 2.5))
segments(x1[1], p1[1], x1[2], p1[2])
segments(x1[2], p1[2], x1[3], p1[3])
segments(x1[3], p1[3], x1[4], p1[4])
segments(x1[4], p1[4], x1[5], p1[5])
```

```{r demandComputational-check}

```
