---
title: "3.4 Monopolistic Competition"
output: 
  learnr::tutorial: 
    progressive: true
runtime: shiny_prerendered
author: 
  name: "Created by [Jim Bang](http://www.github.com/bangecon)"
  email: BangJamesT@sau.edu
  affiliation: "[St. Ambrose University](http://www.sau.edu)"
date: "`r format(Sys.Date(), '%d %b %Y')`"
---

<style type="text/css">
h1{font-size: 24pt}
h2{font-size: 20pt}
h3{font-size: 18pt}
h4,h5,h6{font-size: 16pt}
body{font-size: 16pt}
#doc-metadata {
  margin-bottom: 10px;
  font-size: 0.66em;
}
</style>

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
gradethis_setup(exercise.reveal_solution = TRUE, exercise.eval = TRUE)
knitr::opts_chunk$set(echo = FALSE, fig.height = 6, fig.width = 6, message = FALSE)
```

## Monopolistic Competition

### The Firm's Three Problems

1. Input cost minimization
2. Output profit maximization
3. Input profit maximization

### Monopolistically Competitive Output Markets

$$\matrix{ \matrix{\text{1. Differentiated Products} \\ \text{2. "Relatrively Many" Sellers} }  \\ \text{3. Free Entry and Exit}} \matrix{\implies \\ \implies \\ \implies } \matrix{\text{Firms Influence Price} \\ \text{Firm Price Influence is limited} \\ \text{Profits}\to 0}$$

### The Producer's Input-Hiring Problem

1. Objective: 
2. Endogenous (choice) variable(s): 
3. Exogenous variable(s): 

## Solving for Equilibrium

As the manager of a firm in a monopolistically-competitive industry, your firm faces a demand function estimated as: 

$$Q_d = 600 - \frac{20P}{3}$$

Internal cost analysis estimates the company's long-run cost function to be: 

$$C(Q) = \frac{Q^3}{1600} - \frac{2Q^2}{5} + 110Q$$

### Graphical Approach

Plot the inverse demand, marginal revenue, long-run average cost, and long-run marginal cost functions. 

```{r naturalMonopoly, exercise = TRUE}

```

```{r naturalMonopoly-hint}

```

```{r naturalMonopoly-solution}
library(ggplot2)
ggplot() + 
  geom_function(fun = function(x) 90 - 0.15*x, aes(color = "D")) + 
  geom_function(fun = function(x) x^2/1600 - 0.4*x + 110, aes(color = "LRAC")) + 
  geom_function(fun = function(x) 90 - 0.3*x, aes(color = "MR")) + 
  geom_function(fun = function(x) 3*x^2/1600 - 0.8*x + 110, aes(color = "LRMC")) +
  lims(x = c(0,500), y = c(0, 100)) +
  labs(x = "Quantity", y = "Price", color = "") + 
  scale_color_manual(breaks = c("D", "MR", "LRAC", "LRMC"), values =c("blue", "deepskyblue", "firebrick", "red"))
```

```{r naturalMonopoly-check}
grade_this({
  pass_if(ith_geom_is(.result, "function", i = 1) & 
      ith_geom_is(.result, "function", i = 4))
  fail(random_encouragement())
  })
```

```{r longRun}
question("What would we expect to happen in the long run in this monopolistically-competitive industry?", 
         answer("Firms will exit."), 
         answer("Firms will enter", correct = TRUE), 
         answer("Nothing will happen because of barriers to entry and exit."), 
         answer("Nothing will happen because the industry is in long-run equilibrium."), 
         incorrect = random_encouragement(), 
         correct = random_praise())
```

### Computational Approach

Solve computationally for the profit-maximizing quantity, price, and profit.  

```{r equilibrium, exercise = TRUE}
```

```{r equilibrium-hint1}
# Be careful! 
# The profit function is cubic and has multiple quantities where the slope equals zero. 
# Set your starting `pars` option wisely!
```

```{r equilibrium-hint2}
# The "first" critical value is a (local) minumum at about q = 49. 
# Set the starting `pars` greater than this to get the correct profit-maximizing quantity. 
```

```{r equilibrium-solution}
library(MicroWithR)
pi <- function(x) {
  90*x - 0.15*x^2 - (x^3/1600 - 0.4*x^2 + 110*x)
}
eq <- solnp2(
  pars = 50, 
  fun = pi, 
  LB = 0, 
  control = list(trace = 0), 
  max = TRUE
)
eq$pars
```

```{r equilibrium-check}
grade_this({if (round(.result, 2) != 217.66) { 
    fail("The profit-maximizing quantity is about 217.66") }
  pass()
})
```
