---
title: "1.3 Utility Functions"
author: Jim Bang
output: 
  learnr::tutorial: 
    progressive: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(Deriv)
library(Ryacas)
gradethis_setup()
tutorial_options(exercise.eval = TRUE, exercise.reveal_solution = TRUE)
knitr::opts_chunk$set(echo = FALSE, fig.align = "center", fig.height = 6, fig.width = 6)
knitr::include_graphics
```

## Constant Elasticity of Substitution

Suppose that we can describe a consumer's preferences between two goods using a constant elasticity of substitution utility function with $r = 1/2$: 

$$U(x) = (x_1^{1/2} + x_2^{1/2})^2, $$ 

so that the elasticity of substitution is $\sigma = \frac{1}{1-r} = 2$. 

1. Graph the indifference curves for this utility function for $U(*) = 10$ and $U(*) = 25$. 

```{r plotCES, exercise=TRUE}

```

```{r plotCES-solution}

```

```{r simpleUtility-check}
grade_this({if (!inherits(.result, c("list"))) {
    fail("Your class of your answer should be a list")}
  if (.result$y[1] != Inf) { 
    fail("The y-intercept should approach infinity.") }
  pass()
})
```


### Marginal Utility

The *marginal utility* of each good is the amount that utility changes as a result of a (very small?) change in the amount of that good. Mathematically, for *inifinitesimal* changes, marginal utility is the *derivative* of $u(x)$ with respect to $x_i$, or: 
$$MU_i = \frac{d}{dx_i}u(x_1, x_2, ...,x_n)$$

2. Find the marginal utility functions with respect to goods 1 and 2. 

```{r marginalUtility, exercise = TRUE}

```

```{r marginalUtility-solution}
u <- ysym('(x1^0.5 + x2^0.5)^2')
MU <- deriv(u, c('x1', 'x2'))
```

```{r marginalUtility-check}
grade_this({if (!inherits(.result, c("yac_symbol"))) {
    fail("Your class of your answer should be a yac_symbol")}
  if (.result != "{x1^(-0.5)*(x1^0.5+x2^0.5), x2^(-0.5)*(x1^0.5+x2^0.5)}") { 
    fail("The marginal utilities should be x1^(-0.5)*(x1^0.5+x2^0.5) and
         x2^(-0.5)*(x1^0.5+x2^0.5).") }
  pass()
})
```

3. Find $MRS_{12} = \frac{MU_1}{MU_2}$. Simplify your result. 

```{r marginalRateOfSubstitution, exercise = TRUE}

```

```{r marginalRateOfSubstitution-solution}
MRS <- (MU[1]/MU[2]) |>
  yac_str() |> 
  Simplify() |>
  ysym()
MRS
```

```{r marginalRateOfSubstitution-check}
grade_this({if (!inherits(.result, c("yac_symbol"))) {
    fail("Your class of your answer should be a yac_symbol")}
  if (.result$yacas_cmd != "sqrt(x2)/sqrt(x1)") { 
    fail("The MRS should be sqrt(x2)/sqrt(x1)") }
  pass()
})
```

