---
title: "1.5 Comparative Statics I: Income Changes"
editor_options: 
  markdown: 
    wrap: 72
output: 
  learnr::tutorial: 
    progressive: true
runtime: shiny_prerendered
author: 
  name: "Created by [Jim Bang](http://www.github.com/bangecon)"
  email: BangJamesT@sau.edu
  affiliation: "[St. Ambrose University](http://www.sau.edu)"
date: "`r format(Sys.Date(), '%d %b %Y')`"
---

<style type="text/css">
h1{font-size: 24pt}
h2{font-size: 20pt}
h3{font-size: 18pt}
h4,h5,h6{font-size: 16pt}
body{font-size: 16pt}
#doc-metadata {
  margin-bottom: 10px;
  font-size: 0.66em;
}
</style>

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(ggcheck)
gradethis_setup()
tutorial_options(exercise.eval = TRUE, exercise.reveal_solution = TRUE)
knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE, 
  fig.align = "center", fig.height = 6, fig.width = 6)
knitr::include_graphics
par(mar = c(4, 4, 1, 1))
```

## Introduction

### Basic Idea

Make a *hypothesis* of what would happen to demand for a good when *one* of the exogenous determinants changes.

1.  Ceteris paribus
2.  Small changes  

The most general way to do this is to derive a general form for the demand function *analytically* and analyze the changes in that function with respect to the parameters (preferences, income, prices). This analysis is known as *comparative statics*. Traditionally, we would do this using calculus; more complex models do this using a combination of calculus and computational simulation based on numerical values for the parameters that seem "plausible" based on empirical studies; here, we will focus almost solely on computational solutions based on made-up parameters in order to develop the intuiton.

### Determinants

In the Cobb-Douglas example, the demand function for good $i$ depends on: 

1. Preferences (the budget-share parameters, $\alpha$ and $1 - \alpha$);
2. Income ($m$); 
3. The price of good $i$ ($p_i$);
4. The price of another good ($p_j$).

We will focus mainly on income and price of good 1 (which is the price of the good for good 1 and the price of another good for good 2). 

### Types of Hypotheses

1.  Qualitative: Which direction?
2.  Quantitative: How much?

### Approaches

1.  Graphical (as long as precise values aren't important)
2.  Computational
3.  Anayltical

### Initial Equilibrium

The following code computes the initial equilibrium (which we derived last time): 

```{r initialEquilibrium, echo = TRUE}
library(MicroWithR)
a = 0.25
p1 = 1.5
p2 = 1
m = 100
ux <- function(x)
  x[1] ^ (a) * x[2] ^ (1 - a)
bc <- function(x)
  p1 * x[1] + p2 * x[2]
eq <- solnp2(
  c(1, 1),
  ux,
  ineqfun = bc,
  ineqLB = 0,
  ineqUB = m,
  LB = c(0, 0),
  UB = c(Inf, Inf),
  control = list(trace = 0),
  max = TRUE
)
```

$$x_1^* = 50/3$$ 

$$x_2^* = 75$$

The next bit of code graphs the equilibrium: 

```{r initialGraph, echo = TRUE}
library(ggplot2)
ggplot() +
  geom_function(
    fun = function(x)
      ux(eq$pars) ^ (1 / (1 - a)) / (x ^ (a / (1 - a))),
    color = 'darkgreen'
  ) +
  geom_function(
    fun = function(x)
      m / p2 - x * p1 / p2,
    color = 'red'
  ) +
  geom_point(aes(x = eq$pars[1], y = eq$pars[2]), color = 'blue') +
  geom_text(
    aes(x = eq$pars[1], y = eq$pars[2]),
    label = paste0(
      'x* = (',
      round(eq$pars[1], 2),
      ', ',
      round(eq$pars[2], 2),
      ') \n u(x*) = ',
      round(ux(eq$pars), 2)
    ),
    hjust = 0,
    vjust = 0
  ) +
  lims(x = c(0, 100), y = c(0, 100)) +
  labs(x = "Good 1", y = "Good 2")
```

Now let's see how the equilibrium changes when we change income. 

## Income Change

### Computational Approach

To measure the effect computationally, we have to specify a numerical
amount for the income change, say \$5.

1.  For a consumer with the same preferences (Cobb-Douglass with
    $\alpha = 0.25$), and facing the same prices ($p_1 = 1.5$;
    $p_2 = 1$), compute the new equilibrium for an increase in income
    from \$100 to \$105 (`dm = 5`). Print the (already-stored) initial
    and new optimal bundles and the initial and new values of the
    utility function.

```{r cdIncomeChange, exercise = TRUE}

```

```{r cdIncomeChange-setup}
library(MicroWithR)
ux <- function(x)
  x[1] ^ (a) * x[2] ^ (1 - a)
bc <- function(x)
  p1 * x[1] + p2 * x[2]
a = 0.25
p1 = 1.5
p2 = 1
m = 100 
eq <-
  solnp2(
    c(1, 1),
    ux,
    ineqfun = bc,
    ineqLB = 0,
    ineqUB = m,
    LB = c(0, 0),
    UB = c(Inf, Inf), 
    control = list(trace = 0),
    max = TRUE
  )
```

```{r cdIncomeChange-solution}
dm = 5
dmEq <-
  solnp2(
    c(1, 1),
    ux,
    ineqfun = bc,
    ineqLB = 0,
    ineqUB = m + dm,
    LB = c(0, 0),
    UB = c(Inf, Inf), 
    control = list(trace = 0),
    max = TRUE
  )
dmEq[c(1, 3)]
```

```{r cdIncomeChange-check}
grade_this({if (!inherits(.result, c("list"))) {
    fail("Your class of your answer should be a list")}
  if (.result$pars != c(17.5, 78.75)) { 
    fail("The optimal amount of goods one and two after the change should be 17.5 and 78.75") }
  pass()
})
```

```{r dmEquilibrium}
dm = 5
dmEq <-
  solnp2(
    c(1, 1),
    ux,
    ineqfun = bc,
    ineqLB = 0,
    ineqUB = m + dm,
    LB = c(0, 0),
    UB = c(Inf, Inf), 
    control = list(trace = 0),
    max = TRUE
  )
```

$$x_1^{**} = `r round(dmEq$pars[1], 3)`$$ 

$$x_2^{**} = `r round(dmEq$pars[2], 3)`$$

$$\frac{\Delta x_1}{\Delta m} \Bigg|_{x_1 = x_1^*} \approx `r round((dmEq$pars[1] - eq$pars[1])/dm, 3)`$$

Using the old and new equilibrium values, we can graph the effect of the change. 

### Graphical Approach

Add the `geom_function()` layers to the graph to add the budget line and indifference curve for the equilibrium after the increase in income. Include a `geom_point()` for the new equilibrium bundle, and label the equilibrium using `geom_text()`. 

```{r incomeChangeGraph, exercise = TRUE, exercise.setup = 'dmEquilibrium'}
p <- ggplot() +
  geom_function(
    fun = function(x)
      ux(eq$pars) ^ (1 / (1 - a)) / (x ^ (a / (1 - a))),
    color = 'darkgreen'
  ) +
  geom_function(
    fun = function(x)
      m / p2 - x * p1 / p2,
    color = 'red'
  ) +
  geom_point(aes(x = eq$pars[1], y = eq$pars[2]), color = 'blue') +
  geom_text(
    aes(x = eq$pars[1], y = eq$pars[2]),
    label = paste0(
      'x* = (',
      round(eq$pars[1], 2),
      ', ',
      round(eq$pars[2], 2),
      ') \n u(x*) = ',
      round(ux(eq$pars), 2)
    ),
    hjust = 0,
    vjust = 0
  ) +
  lims(x = c(0, 100), y = c(0, 100)) +
  labs(x = "Good 1", y = "Good 2")
# Add the new budget line and indifference curve to `p` below: 

```

```{r incomeChangeGraph-setup}
library(MicroWithR)
ux <- function(x)
  x[1] ^ (a) * x[2] ^ (1 - a)
bc <- function(x)
  p1 * x[1] + p2 * x[2]
a = 0.25
p1 = 1.5
p2 = 1
m = 100 
eq <-
  solnp2(
    c(1, 1),
    ux,
    ineqfun = bc,
    ineqLB = 0,
    ineqUB = m,
    LB = c(0, 0),
    UB = c(Inf, Inf), 
    control = list(trace = 0),
    max = TRUE
  )
m = 100
dm = 5
dmEq <-
  solnp2(
    c(1, 1),
    ux,
    ineqfun = bc,
    ineqLB = 0,
    ineqUB = m + dm,
    LB = c(0, 0),
    UB = c(Inf, Inf),
    control = list(trace = 0),
    max = TRUE
  )
```

```{r incomeChangeGraph-solution}
p <- ggplot() +
  geom_function(
    fun = function(x)
      ux(eq$pars) ^ (1 / (1 - a)) / (x ^ (a / (1 - a))),
    color = 'darkgreen'
  ) +
  geom_function(
    fun = function(x)
      m / p2 - x * p1 / p2,
    color = 'red'
  ) +
  geom_point(aes(x = eq$pars[1], y = eq$pars[2]), color = 'blue') +
  geom_text(
    aes(x = eq$pars[1], y = eq$pars[2]),
    label = paste0(
      'x* = (',
      round(eq$pars[1], 2),
      ', ',
      round(eq$pars[2], 2),
      ') \n u(x*) = ',
      round(ux(eq$pars), 2)
    ),
    hjust = 0,
    vjust = 0
  ) +
  lims(x = c(0, 100), y = c(0, 100)) +
  labs(x = "Good 1", y = "Good 2")
p + geom_function(
  fun = function(x)
    ux(dmEq$pars) ^ (1 / (1 - a)) / (x ^ (a / (1 - a))),
  color = 'darkgreen'
) +
  geom_function(
    fun = function(x)
      (m + dm) / p2 - x * p1 / p2,
    color = 'red'
  ) +
  geom_point(aes(x = dmEq$pars[1], y = dmEq$pars[2]), color = 'blue') +
  geom_text(
    aes(x = dmEq$pars[1], y = dmEq$pars[2]),
    label = paste0(
      'x* = (',
      round(dmEq$pars[1], 2),
      ', ',
      round(dmEq$pars[2], 2),
      ') \n u(x*) = ',
      round(ux(dmEq$pars), 2)
    ),
    hjust = 0,
    vjust = 0
  )
```

```{r incomeChangeGraph-check}
grade_this({
  if (uses_geoms(.result, "function") & uses_geoms(.result, "point")) {
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

## Engel Curves

The *Engel Curve* plots the demand for a good against the level of income. In our Cobb-Douglas example, this "curve" is a straight line. In the appendix, we derive the demand curve for good 1 to be: 

$$x_1(p_1, \alpha, m) = \frac{\alpha m}{p_1}$$

Plot the Engel curve for good 1 from the example ($\alpha = 0.25$, $p_1 = 1.5$, and $p_2 = 1$).

```{r cdEngel, exercise = TRUE}

```

```{r cdEngel-solution}
library(ggplot2)
a = 0.25
p1 = 1.5
p2 = 1
m = 100 
ggplot() + 
  geom_function(fun = function(x) a * x / p1) + 
  lims(x = c(0, 150)) +
  labs(x = "Income", y = "Consumption of Good 1") 
```

```{r cdEngel-check}
grade_this({
  if (uses_geoms(.result, "function")) {
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

We can check this by looking at the "income expansion path" from the indifference map and corresponding solutions for different levels of income. 

```{r linearIEP, echo = TRUE}
library(MicroWithR)
library(ggplot2)
a = 0.25
p1 = 1.5
p2 = 1
m = 100 
m4 = seq(60, 120, 20)
ux <- function(x) {x[1] ^ (a) * x[2] ^ (1 - a)}
bc <- function(x) {p1 * x[1] + p2 * x[2]}
eq060 <- solnp2(c(1, 1), ux, ineqfun = bc, ineqLB = 0, ineqUB = m4[1],
    LB = c(0,0), UB = c(Inf, Inf), max = TRUE)
eq080 <- solnp2(c(1, 1), ux, ineqfun = bc, ineqLB = 0, ineqUB = m4[2],
    LB = c(0,0), UB = c(Inf, Inf), max = TRUE)
eq100 <- solnp2(c(1, 1), ux, ineqfun = bc, ineqLB = 0, ineqUB = m4[3],
    LB = c(0,0), UB = c(Inf, Inf), max = TRUE)
eq120 <- solnp2(c(1, 1), ux, ineqfun = bc, ineqLB = 0, ineqUB = m4[4],
    LB = c(0,0), UB = c(Inf, Inf), max = TRUE)
g <- ggplot() + 
  geom_function(fun = function(x)
    ux(eq060$pars) ^ (1 / (1 - a)) / (x ^ (a / (1 - a))), color = 'red') +
  geom_function(fun = function(x)
    ux(eq080$pars) ^ (1 / (1 - a)) / (x ^ (a / (1 - a))), color = 'orange') +
  geom_function(fun = function(x)
    ux(eq100$pars) ^ (1 / (1 - a)) / (x ^ (a / (1 - a))), color = 'green') +
  geom_function(fun = function(x)
    ux(eq120$pars) ^ (1 / (1 - a)) / (x ^ (a / (1 - a))),color = 'blue') +
  geom_function(fun = function(x) m / p2 - (p1 / p2)*x) + 
  geom_function(fun = function(x) p1*(1-a)*x / (p2*a)) + 
  geom_hline(aes(yintercept = eq060$pars[2]), color = 'red', linetype = 'dashed') + 
  geom_hline(aes(yintercept = eq080$pars[2]), color = 'orange', linetype = 'dashed') + 
  geom_hline(aes(yintercept = eq100$pars[2]), color = 'green', linetype = 'dashed') + 
  geom_hline(aes(yintercept = eq120$pars[2]), color = 'blue', linetype = 'dashed') + 
  geom_vline(aes(xintercept = eq060$pars[1]), color = 'red', linetype = 'dashed') + 
  geom_vline(aes(xintercept = eq080$pars[1]), color = 'orange', linetype = 'dashed') + 
  geom_vline(aes(xintercept = eq100$pars[1]), color = 'green', linetype = 'dashed') + 
  geom_vline(aes(xintercept = eq120$pars[1]), color = 'blue', linetype = 'dashed') + 
  lims(x = c(0, m / p1), y = c(0, m / p2)) + labs(x = "Beer", y = "Donuts")
engel1 <- ggplot() + 
  geom_function(fun = function(x) p1 * x / a) + 
  geom_hline(aes(yintercept = m4[1]), color = 'red', linetype = 'dashed') + 
  geom_hline(aes(yintercept = m4[2]), color = 'orange', linetype = 'dashed') + 
  geom_hline(aes(yintercept = m4[3]), color = 'green', linetype = 'dashed') + 
  geom_hline(aes(yintercept = m4[4]), color = 'blue', linetype = 'dashed') + 
  geom_vline(aes(xintercept = eq060$pars[1]), color = 'red', linetype = 'dashed') + 
  geom_vline(aes(xintercept = eq080$pars[1]), color = 'orange', linetype = 'dashed') + 
  geom_vline(aes(xintercept = eq100$pars[1]), color = 'green', linetype = 'dashed') + 
  geom_vline(aes(xintercept = eq120$pars[1]), color = 'blue', linetype = 'dashed') + 
  lims(x = c(0, m/p1), y = c(0, m4[4])) + labs(x = "Beer Consumption", y = "Income")
engel2 <- ggplot() + 
  geom_function(fun = function(x) p2 * x / (1-a)) + 
  geom_hline(aes(yintercept = m4[1]), color = 'red', linetype = 'dashed') + 
  geom_hline(aes(yintercept = m4[2]), color = 'orange', linetype = 'dashed') + 
  geom_hline(aes(yintercept = m4[3]), color = 'green', linetype = 'dashed') + 
  geom_hline(aes(yintercept = m4[4]), color = 'blue', linetype = 'dashed') + 
  geom_vline(aes(xintercept = eq060$pars[2]), color = 'red', linetype = 'dashed') + 
  geom_vline(aes(xintercept = eq080$pars[2]), color = 'orange', linetype = 'dashed') + 
  geom_vline(aes(xintercept = eq100$pars[2]), color = 'green', linetype = 'dashed') + 
  geom_vline(aes(xintercept = eq120$pars[2]), color = 'blue', linetype = 'dashed') + 
  lims(x = c(0, m/p2), y = c(0, m4[4])) + labs(x = "Donut Consumption", y = "Income") + 
  coord_flip()
ggpubr::ggarrange(g, engel2, engel1)
```

### Nonlinear Income Expansion Path Example 

A "simple" example of a nonlinear income expansion path comes from the
special case of perfect complements, where:

$$u(x) = min\{x_1^2, x_2\}.$$

In this example, for any nonzero set of prices, the consumer would
choose to set $x_2 = x_1^2$ (to put consumption in the corner of the
"L". The following code shows the analytical solution and its graph
for different levels of income.

```{r nonlinearIEP, echo = TRUE}
library(MicroWithR)
library(ggplot2)
p1 = 1.5
p2 = 1
m4 = seq(60, 120, 20)
complements <- function(x) {min(x[1] ^ 2, 25 * x[2])}
bc <- function(x)
  p1 * x[1] + p2 * x[2]
eq060 <- solnp2(c(1, 1), complements, ineqfun = bc, ineqLB = 0, ineqUB = m4[1],
    LB = c(0,0),  UB = c(Inf, Inf), max = TRUE)
eq080 <- solnp2(c(1, 1), complements, ineqfun = bc, ineqLB = 0, ineqUB = m4[2],
    LB = c(0,0), UB = c(Inf, Inf), max = TRUE)
eq100 <- solnp2(c(1, 1), complements, ineqfun = bc, ineqLB = 0, ineqUB = m4[3],
    LB = c(0,0), UB = c(Inf, Inf), max = TRUE)
eq120 <- solnp2(c(1, 1), complements, ineqfun = bc, ineqLB = 0, ineqUB = m4[4],
    LB = c(0,0), UB = c(Inf, Inf), max = TRUE)
g <- ggplot() + 
  geom_segment(aes(x = eq060$pars[1], xend = eq060$pars[1], 
                   y = eq060$pars[2], yend = m / p2), color = 'red') +
  geom_segment(aes(x = eq060$pars[1], xend = m / p1, 
                   y = eq060$pars[2], yend = eq060$pars[2]), color = 'red') +
  geom_segment(aes(x = eq080$pars[1], xend = eq080$pars[1], 
                   y = eq080$pars[2], yend = m / p2), color = 'orange') +
  geom_segment(aes(x = eq080$pars[1], xend = m / p1, 
                   y = eq080$pars[2], yend = eq080$pars[2]), color = 'orange') +
  geom_segment(aes(x = eq100$pars[1], xend = eq100$pars[1], 
                   y = eq100$pars[2], yend = m / p2), color = 'green') +
  geom_segment(aes(x = eq100$pars[1], xend = m / p1, 
                   y = eq100$pars[2], yend = eq100$pars[2]), color = 'green') +
  geom_segment(aes(x = eq120$pars[1], xend = eq120$pars[1], 
                   y = eq120$pars[2], yend = m / p2), color = 'blue') +
  geom_segment(aes(x = eq120$pars[1], xend = m / p1, 
                   y = eq120$pars[2], yend = eq120$pars[2]), color = 'blue') +
  geom_function(fun = function(x) m / p2 - (p1 / p2)*x) + 
  geom_function(fun = function(x) (x^2)/25) + 
  geom_hline(aes(yintercept = eq060$pars[2]), color = 'red', linetype = 'dashed') + 
  geom_hline(aes(yintercept = eq080$pars[2]), color = 'orange', linetype = 'dashed') + 
  geom_hline(aes(yintercept = eq100$pars[2]), color = 'green', linetype = 'dashed') + 
  geom_hline(aes(yintercept = eq120$pars[2]), color = 'blue', linetype = 'dashed') + 
  geom_vline(aes(xintercept = eq060$pars[1]), color = 'red', linetype = 'dashed') + 
  geom_vline(aes(xintercept = eq080$pars[1]), color = 'orange', linetype = 'dashed') + 
  geom_vline(aes(xintercept = eq100$pars[1]), color = 'green', linetype = 'dashed') + 
  geom_vline(aes(xintercept = eq120$pars[1]), color = 'blue', linetype = 'dashed') + 
  lims(x = c(0, m / p1), y = c(0, m / p2)) +
  labs(x = "Beer", y = "Donuts")
engel1 <- ggplot() + 
  geom_function(fun = function(x) p2 * x ^ 2 / 25 + p1 * x) + 
  geom_hline(aes(yintercept = m4[1]), color = 'red', linetype = 'dashed') + 
  geom_hline(aes(yintercept = m4[2]), color = 'orange', linetype = 'dashed') + 
  geom_hline(aes(yintercept = m4[3]), color = 'green', linetype = 'dashed') + 
  geom_hline(aes(yintercept = m4[4]), color = 'blue', linetype = 'dashed') + 
  geom_vline(aes(xintercept = eq060$pars[1]), color = 'red', linetype = 'dashed') + 
  geom_vline(aes(xintercept = eq080$pars[1]), color = 'orange', linetype = 'dashed') + 
  geom_vline(aes(xintercept = eq100$pars[1]), color = 'green', linetype = 'dashed') + 
  geom_vline(aes(xintercept = eq120$pars[1]), color = 'blue', linetype = 'dashed') + 
  lims(x = c(0, m/p1), y = c(0, m4[4])) + 
  labs(x = "Beer Consumption", y = "Income")
engel2 <- ggplot() + 
  geom_function(fun = function(x) p2 * x + 5 * p1 * sqrt(x)) + 
  geom_hline(aes(yintercept = m4[1]), color = 'red', linetype = 'dashed') + 
  geom_hline(aes(yintercept = m4[2]), color = 'orange', linetype = 'dashed') + 
  geom_hline(aes(yintercept = m4[3]), color = 'green', linetype = 'dashed') + 
  geom_hline(aes(yintercept = m4[4]), color = 'blue', linetype = 'dashed') + 
  geom_vline(aes(xintercept = eq060$pars[2]), color = 'red', linetype = 'dashed') + 
  geom_vline(aes(xintercept = eq080$pars[2]), color = 'orange', linetype = 'dashed') + 
  geom_vline(aes(xintercept = eq100$pars[2]), color = 'green', linetype = 'dashed') + 
  geom_vline(aes(xintercept = eq120$pars[2]), color = 'blue', linetype = 'dashed') + 
  lims(x = c(0, m/p2), y = c(0, m4[4])) + labs(x = "Donut Consumption", y = "Income") + 
  coord_flip()
ggpubr::ggarrange(g, engel2, engel1)
```

## Appendix: Analytical Approach

The *one* place that analytical methods help is in deriving the functions that relate the demand for a good with the consumer's income, the price of the good, and the price(s) of other good(s). It allows us to express demand for a good in terms of the model parameters (preferences, income, and prices) and derive *general* forms for the comparative statics.  

Basic approach: 

1. Find the demand functions in general form, without specific parameter values for $m$, $\alpha$, $p_1$, and $p_2$. 
2. Take the derivative of the demand functions with respect to each variable. 

R can actually do this super-fast (by vectorizing the derivatives)!

Solving the first order conditions for $x_1$ and $x_2$ in terms of the parameters
(the general-form solution from last time), we get the demand functions. The following code derives this result.

```{r demand, echo = TRUE}
library(Ryacas)
u <- ysym('x1^(alpha) * x2^(1-alpha)')
L <- u + ysym('lambda*(m - p1*x1 - p2*x2)')
dL <- deriv(L, c('x1', 'x2', 'lambda'))
MU <- c(
  p1*solve(dL[1], 'lambda'), 
  p2*solve(dL[2], 'lambda')) |>
  y_rmvars()
MRS <- with_value(MU[1]/MU[2], 'p1', p1) |>
  with_value('p2', p2) |> yac_str() |>
  Deriv::Simplify() |> ysym()
D <- solve(c(MRS, dL[[3]]), c(ysym('p1/p2'), 0), c('x1', 'x2')) |> 
  y_rmvars() |>
  simplify()
D[1] <- with_value(D[1], 'x2', D[2]) |> 
  simplify() |> suppressWarnings()
D
```

In a more reader friendly typesetting format using `Ryacas::tex()` we can write this as:
$$x_1(p_1, m) = `r tex(D[1])`$$
$$x_2(p_2, m) = `r tex(D[2])`$$
Notice that *for these preferences* the increase in consumption for $x_1$ is linear in the amount of income (it only depends on the exogenous parameters $\alpha$ and $p_1$).

Run the following code to take the derivative of demand for $x_1$ and $x_2$ with respect to income and prices. 

```{r analyticalStatics, echo = TRUE}
(statics <- deriv(D, c('m', 'p1', 'p2')))
```

substituting our parameter values, we have: 

$$\nabla x_1(m, p_1, p_2) = \begin{pmatrix} \frac{d}{dm}x_1(m, p_1) \\ \frac{d}{dp_1}x_1(m, p_1) \\ \frac{d}{dp_2}x_1(m, p_1) \end{pmatrix} 
 = \begin{pmatrix} `r tex(statics[1,1])` \\  `r tex(statics[2,1])` \\ `r tex(statics[3,1])` \end {pmatrix} 
 = \begin{pmatrix} \frac{0.25}{1.50} \\ \frac{-25}{1.50^2} \\ 0 \end{pmatrix}
 \approx \begin{pmatrix} 0.16667 \\ -11.11 \\ 0 \end{pmatrix}$$

and 

$$\nabla x_2(m, p_1, p_2) = \begin{pmatrix} \frac{d}{dm}x_2(m, p_1) \\ \frac{d}{dp_1}x_2(m, p_1) \\ \frac{d}{dp_2}x_2(m, p_1) \end{pmatrix} 
 = \begin{pmatrix} `r tex(statics[1,2])` \\  `r tex(statics[2,2])` \\ `r tex(statics[3,2])` \end {pmatrix} 
 = \begin{pmatrix} \frac{0.75}{1} \\ 0 \\ \frac{-75}{1^2} \end{pmatrix}
 \approx \begin{pmatrix} 0.75 \\ 0 \\ -75 \end{pmatrix}$$

