---
title: "3.2 Market Supply and Equilibrium in Competitive Markets"
output: 
  learnr::tutorial: 
    progressive: true
runtime: shiny_prerendered
author: 
  name: "Created by [Jim Bang](http://www.github.com/bangecon)"
  email: BangJamesT@sau.edu
  affiliation: "[St. Ambrose University](http://www.sau.edu)"
date: "`r format(Sys.Date(), '%d %b %Y')`"
---

<style type="text/css">
h1{font-size: 24pt}
h2{font-size: 20pt}
h3{font-size: 18pt}
h4,h5,h6{font-size: 16pt}
body{font-size: 16pt}
#doc-metadata {
  margin-bottom: 10px;
  font-size: 0.66em;
}
</style>

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
gradethis_setup(exercise.reveal_solution = TRUE, exercise.eval = TRUE)
knitr::opts_chunk$set(echo = FALSE, fig.height = 6, fig.width = 6)
```

## Deriving Market Supply

Market supply is the "horizontal sum" (adding up the quantities along the x-axis) of the individual supply curves. 

In perfectly competitive markets we implicitly assume identical technologies (same production function and same parameters).

### Review: Cost Minimization and Short-Run Marginal Cost

Suppose that the production function for that firm took a Cobb-Douglas form with labor share equal to $1/3$ and total factor productivity parameter $50$. It also faced input prices of $w = 25$ and $r = 50$. The firm planned its initial capital input (and planned labor input) based on an expected quantity of $100$. 

$$\min_{L,K} \bigg\{C(L, K) = 10L + 20K \bigg\}$$

$$\text{s.t.: } f(L,K) = 50L^{1/3}K^{2/3} \ge 2500$$

The "Lagrangian" function for our example is: 

$$\mathcal{L} = wL + rK + \lambda (2500 - 50L^{1/3}K^{2/3})$$

Solve for the cost-minimizing combination of labor and capital to produce **100** units.  

```{r costMin, exercise = TRUE}

```

```{r costMin-solution}
library(Rsolnp)
A = 50
a = 1/3
Q0 = 100
w = 25
r = 50
tc <- function(x) w*x[1] + r*x[2]
qx <- function(x) A * x[1]^a * x[2]^(1-a)
costMin <- solnp(
  pars = c(1, 1),
  fun = tc,
  ineqfun = qx,
  ineqLB = Q0,
  ineqUB = Inf,
  LB = c(0, 0),
  control = list(trace = 0)
)
```

```{r costMin-check}
grade_this({if (round(.result$pars, 0) != c(2,2)) { 
    fail("The cost-minimizing labor and capital inputs are 2 each.") }
  pass()
})
```

### Individual Firm Supply Curve (Marginal Cost Curve)

The first step towards deriving the short-run supply curve is to derive the short-run total costs. We did this computationally in section 2.3. With only one variable input, doing this analytically is not so difficult: total fixed cost is the price of the fixed input times the planned level of the fixed input(s), or $rK = 50 \cdot 2$; total variable cost is $w \cdot L(Q) = 25L(Q)$, where $L(Q)$ is found by substituting $K=2$ into the production function and inverting it to solve for $L$ as a function of $Q$, or $L(Q) = ({Q}/{AK^{2/3}})^3 = Q^3/500,000$. Plugging everything into the cost function, we get:

$Q = 50*L^{1/3}2^{2/3}$ $L = Q^3/[50^3*2^{2}]$

$$ TC = 25 \cdot L(Q) + 50 \cdot 2 = \frac{Q^3}{20,000} + 100$$

Applying the power rule for derivatives and noting that when the firm in a competitive market is a *price taker* (i.e. the price does not depend on the *individual* firm's output choice), we get: 

$$P(Q_i) = MC = \frac{dTC}{dQ_i} = \frac{3Q_i^2}{20,000}$$

If you don't feel confident doing the derivation on your own, Ryacas can help as a tool:

```{r mc, echo = TRUE, message = FALSE}
library(Ryacas)
TC <- ysym('Q^3 / 20000 + 100')
MC <- deriv(TC, 'Q') 
MC
```

### Many Identical Firms

Since this *particular* production function always entails diminishing returns to labor, average variable cost reaches its minimum at $Q = 0$ so the firm's (inverse) supply curve is the entire marginal cost curve for $Q \ge 0$. 

One way to *explicitly* solve for the market supply is to get representative firm i's quantity supplied as a function of price ($Q_s^i = f(P)$) by inverting the marginal cost function, which in our case yields: 

$$Q_i(P) = \Bigg(\frac{20,000P}{3}\Bigg)^{1/2} = 100 \cdot \sqrt{\frac{2P}{3}}$$

If the market has 9,000 identical firms, then finding the market supply curve is pretty straightforward: multiply the individual firms' quantities supplied (at each given price) times 9,000. Hence, in our market, 

$$Q_s = 9,000 \cdot 100 \cdot \sqrt{\frac{2P}{3}} = 900,000 \cdot \sqrt{\frac{2P}{3}}$$

The inverse supply function (for plotting P on the y-axis) is: 

$$P(Q_s) = \frac{3Q^2}{2 \cdot 900,000^2} = \frac{Q^2}{540,000,000,000}$$

An *even quicker* way to get this result is to notice that when the firms are identical, each firm takes $\frac{1}{n^{th}}$ of the total market supply which in our case means $Q_i = \frac{Q_s}{9,000}$. 

$$P(Q_s) = \frac{3\Big(\frac{Q_s}{9,000}\Big)^2}{20,000} = \frac{\frac{3Q_s^2}{81,000,000}}{20,000} = \frac{\frac{Q_s^2}{27,000,000}}{20,000} = \frac{Q^2}{540,000,000,000}$$

Plot the market supply in the usual (Marshallian) way, with price on the vertical axis and quantity on the horizontal axis. 

```{r marketSupplyPlot, exercise = TRUE}

```

```{r marketSupplyPlot-solution}
library(ggplot2)
ggplot() + 
  geom_function(fun = function(x) x ^ 2 / 5.4e11, aes(color = "S")) + 
  lims(x = c(0,2500000), y = c(0, 2.5)) +
  labs(x = "Quantity", y = "Price", color = "") + 
  scale_color_manual(breaks = c("S"), values = c("red"))
```

```{r marketSupplyPlot-check}
grade_this({
  if (ith_geom_is(.result, "function", i = 1)) {
    pass(random_praise())
    }
  fail(random_encouragement())
  })
```

## Equilibrium 

### Graphical Approach

Plot the market (inverse) supply and demand curves on the same plot area. To remind you, we solved the market demand function as: 

$$P(Q_d) = \frac{1.5 \cdot 10^6}{Q}$$

We just derived the market supply function as: 

$$P(Q_s) = \frac{Q^2}{5.4 \cdot 10^{11}}$$

```{r equilibrium, exercise = TRUE}

```

```{r equilibrium-setup}
library(ggplot2)
```

```{r equilibrium-solution}
ggplot() + 
  geom_function(fun = function(x) 1500000/x, aes(color = "D")) + 
  geom_function(fun = function(x) x ^ 2 / 5.4e11, aes(color = "S")) + 
  lims(x = c(0,2500000), y = c(0, 2.5)) +
  labs(x = "Quantity", y = "Price", color = "") + 
  scale_color_manual(breaks = c("D", "S"), values = c("blue", "red"))
```

It looks like the equilibrium crosses somewhere between a price of 1.5 and 1.75 and a quantity of on the high side of the range between 750,000 and 1,000,000. In fact, we know that with 900,000 identical producers the quantity will be greater than 900,000 since at a price of 1.5 each firm would optimally produce 100 units. Let's solve for the equilibrium computationally. 

### Computational Approach

The equations we're trying to solve here are a nonlinear mess! Luckily, as long as we choose starting values for the variables that are "close" to the equilibrium, R will have no trouble solving them. 

```{r eqSolve, exercise = TRUE}

```

```{r eqSolve-setup}

```

```{r eqSolve-solution}
srEq <- nleqslv(
  c(1,900000), 
  function(x) c(
    x[1]*x[2] - 1500000,    # Demand
    5.4e11 * x[1] - x[2]^2  # Supply
    )
  )
srEq$pars
```

```{r eqSolve-check}
grade_this({if (round(.result, 1) != c(1.6,932169.8)) { 
    fail("The cost-minimizing labor and capital inputs are about 1.6 and 932170.") }
  pass()
})
```
