---
output: 
  learnr::tutorial: 
    progressive: true
runtime: shiny_prerendered
title: "1.1 Budget Constraint"
author: Jim Bang
description: "The consumer's budget constraint"
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
gradethis_setup()
tutorial_options(exercise.reveal_solution = FALSE, exercise.eval = FALSE, exercise.checker = gradethis_exercise_checker)
knitr::opts_chunk$set(echo = FALSE, fig.align = "center", fig.height = 6.5, fig.width = 8.5)
```


### Food Stamps

Consider a consumer who receives "food stamps" that subsidizes up to five donuts (but does not subsidize beer). Graph the budget line for this consumer using the original parameters ($m = 20$, $p_d = 1$, and $p_b = 2$). (Hint: You need to plot two separate lines because there are two segments to this budget line. The first is as if income has increased by the dollar value of 5 donuts; the second part is horizontal noting that you cannot spend the subsidy on beer and so the maximum amount of beer the consumer can buy does not change. Use the `segments(x0, x1, y0, y1)` function to draw this part.)

```{r foodStamps, exercise = TRUE}

```

```{r foodStamps-solution}
m = 20
p_d = 1
p_b = 2
s_d = 5
curve(
  s_d + m / p_d - p_b * x / p_d,
  from = 0,
  to = m / p_b,
  xlab = "Beer",
  ylab = "Donuts",
  xlim = c(0, m / p_b),
  ylim = c(0, s_d + m / p_d)
)
segments(
  x0 = m / p_b,
  x1 = m / p_b,
  y0 = 0,
  y1 = s_d
)
```

```{r foodStamps-check}
grade_code() 
```

