---
output: 
  learnr::tutorial: 
    progressive: true
runtime: shiny_prerendered
title: "Introduction to Mathing with R"
author: 
  name: "Created by [Jim Bang](http://www.github.com/bangecon)"
  email: BangJamesT@sau.edu
  affiliation: "[St. Ambrose University](http://www.sau.edu)"
date: "`r format(Sys.Date(), '%d %b %Y')`"
---

<style type="text/css">
h1{font-size: 24pt}
h2{font-size: 20pt}
h3{font-size: 18pt}
h4,h5,h6{font-size: 16pt}
body{font-size: 16pt}
#doc-metadata {
  margin-bottom: 10px;
  font-size: 0.66em;
}
</style>

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(ggcheck)
library(Ryacas)
library(Deriv)
gradethis_setup()
tutorial_options(exercise.reveal_solution = TRUE, exercise.eval = FALSE, exercise.checker = gradethis_exercise_checker)
knitr::opts_chunk$set(echo = FALSE)
```

## Getting Started with R

### Why R? 

R helps you talk to a computer and help solve big problems. Employers use R (or other similar languages like Python or Julia) for statistics and data analysis. R is not usually associated with calculus or mathematics, but can do anything that expensive mathematical-analysis programs like Matlab or Mathematica can do - and possibly more! Also, R is free! 

We will use R in this class for two reasons: 

1. As a practical alternative to pencil-and-paper calculus; 
2. To develop basic proficiency with a computing language.

### Installing R

```{r installR, echo=FALSE}
question("Where do you download R?",
    answer("[www.rstudio.com/download](www.rstudio.com/download)"),
    answer("[cloud.r-project.org](http://cloud.r-project.org)", correct = TRUE, message = "You can also download R from [cran.r-project.org](http://cran.r-project.org)"),
    answer("[www.r-project.org](www.r-project.org)", message = "Good try, but not exactly. www.r-project.org doesn't provide a download link, but it does provide a link to one the websites above."),
    answer("[www.r.com](www.r.com)"),
    allow_retry = TRUE
  )
```

### Updating R

```{r updaterR, echo=FALSE}
question("How often should you update R?",
    answer("Everytime you use it", message = "This will be too often unless you use R very rarely!"),
    answer("About once a year", correct = TRUE, "A new version of R is released about once a year. Update sooner if you encounter a bug that you cannot explain."),
    answer("Never", message = "A new version of R is released about once a year. I'll assume that you are using the newest version of R, which will be the fastest version with the fewest unexpected behaviors." ),
    allow_retry = TRUE
  )
```


## RStudio

### What is RStudio?

```{r rStudio, echo=FALSE}
  question("What is the RStudio IDE?",
    answer("An application that makes it easier to use R.", correct = TRUE, message = "RStudio organizes your workspace and makes it easier to write, use, debug, and save R code. I highly recommend it and will rely on the RStudio IDE in several tutorials."),
    answer("An application that let's you use R without writing any code", message = "And thank goodness! Code provides a reproducible record of your work, which is essential for data science."),
    answer("A spreadsheet program like Microsoft Excel."),
    answer("Another name for R", message = "R and RStudio are two separate things. R is a language, like English. Think of RStudio as a program that helps you use the language, kind of like how a word processing program helps you write in English."),
    allow_retry = TRUE
  )
```

### Where do I get RStudio? 

```{r installRStudio, echo=FALSE}
question("Where do you download RStudio?",
    answer("[www.rstudio.com/download](www.rstudio.com/download)", correct = TRUE, message = "For these tutorials, download and install the RStudio Desktop Open Source License."),
    answer("[cloud.r-project.org](http://cloud.r-project.org)", message = "This is where you download R, not RStudio."),
    answer("[www.r-project.org](www.r-project.org)"),
    answer("[cran.rstudio.org](http://cran.rstudio.org)"),
    allow_retry = TRUE
  )
```

### Do I still need R?

```{r installR_RStudio, echo=FALSE}
  question("Do you need to install R if you already have RStudio?",
    answer("Yes.", correct = TRUE),
    answer("No.", message = "R does not come with RStudio; you need to install R separately.")
  )
```

## Using R as a Calculator

### CalculatR

Write the R code required to perform the following operations:

- $21 + 21$
- $5(4-1)^2-3$
- $\sqrt{1764}$
- $ln(e^{42})$

```{r calculatR, exercise=TRUE}

```

```{r calculatR-solution, warning=FALSE, message = FALSE}
21+21
5*(4-1)^2-3
sqrt(1764)
log(exp(42))
```

```{r calculatR-hint}
1 + 1
```

```{r calculatR-check}
grade_result(
  pass_if(~identical(.result, 42))) 
```

### Using Objects

R is an object-oriented language. The "assignment" operator, "<-" (keyboard shortcut: "Alt" + "-") computes the operation to the right and gives it name in your *environment* (see top-right pane in RStudio). Name the calculations from the previous exercise as indicated below.

- $21 + 21$: answer
- $5(4-1)^2-3$: life
- $\sqrt{1764}$: universe
- $ln(e^{42})$: everything

Display the objects in your environment using the `ls()` function. 

```{r objects, exercise=TRUE}

```

```{r objects-solution, warning=FALSE, message = FALSE}
answer <- 21+21
life <- 5*(4-1)^2-3
universe <- sqrt(1764)
everything <- log(exp(42))
ls()
```

```{r objects-check}
grade_this({if (!inherits(.result, c("character"))) {
    fail("Your class of your answer should be a character list")}
  if (.result[1] != "answer") { 
    fail("The first position in your list whould be named 'answer.'") }
  pass()
})
```

Note that in some cases you can use the "equal to" operator, "=", but only if the thing you're operating and the name of the object you're assigning are unambiguous. The direction of the "<-" operator points *from* the operation you mean for R to calculate and *to* the name you want to give to it. In fact, you may reverse the direction to assign from left to right, as in `21 + 21 -> answer`. 

## Working with Math Functions

### Defining and Evaluating Functions

If you can define an object you can write (and evaluate) your own function! A function can be a simple mathematical function, or it can be a long script that outputs several different values. Define the function, `quadratic`, as a function of *variable* `x` and *parameters* `a`, `b`, and `c` as 
$$f(x) = ax^2 + bx + c.$$ 
Then, evaluate $f(x) = 8x^2 - 6x - 12$ at $x = 3$

```{r functions, exercise=TRUE}

```

```{r functions-solution, warning=FALSE, message = FALSE}
quadratic <- function(x, a, b, c) {
  a*x^2 + b*x + c
}
quadratic(3, 8, -6, -12)
```

```{r functions-check}
grade_result(
  pass_if(~identical(.result, 42))) 
```

For the purposes of getting computational solutions, there is not really a difference between a parameter and a variable. Economically, we consider parameters (a, b, c, and d) to be *exogenous* - determined by nature (like preference parameters) or markets (like prices or income) - whereas we consider variables (x) to be *endogenous* - chosen by households (like quantities consumed) or firms (like amounts of a resource hired). 

### Graphing a Function

<<<<<<< HEAD
The `ggplot2` package produces professional-looking graphs that look great with the defaults, and that you can learn to customize in infinitely many ways with experience and practice. The function, `ggplot()` initiates a plot, and you can add layer after layer of plots, labels, data, and formatting options using "geom" functions. For mathematical functions, the `geom_function()` function can add any function that you specify with `function(x)` followed by an `expression`. You can also store the plot properties as an object and add features as you think of them. 

Define a plot, `p` with the math function $f(x) = 8x^2 - 6x - 12$ \textcolor{blue}{*blue*}. 
=======
The `ggplot2` package produces professional-looking graphs that look great with the defaults, and that you can learn to customize in infinitely many ways with experience and practice. The function, `ggplot()` initiates a plot, and you can add layer after layer of plots, labels, data, and formatting options using "geom" functions. For mathematical functions, the `geom_function()` function can add any function that you specify with `function(x)` followed by an `expression` that includes mathematical operations applied to the variable, $x$.  

You can also store the plot properties as an object and add features as you think of them. Define a plot, `g` with the math function $f(x) = 8x^2 - 6x - 12$ \textcolor{blue}{*blue*}.  
>>>>>>> 89e5e3d9ff9bf79c337ecded449db08ebc3ff1fb

```{r plot, exercise=TRUE, fig.height=6.5, fig.width=4.5, fig.align='center'}

```

<<<<<<< HEAD
```{r plot-solution}
p <- ggplot() +
  geom_function(fun = function(x) 8*x^2 - 6*x - 12, color = "blue") 
p
=======
```{r plot-solution, warning=FALSE, message = FALSE}
g <- ggplot() +
  geom_function(fun = function(x) 8*x^2 - 6*x - 12, color = "blue") 
g
>>>>>>> 89e5e3d9ff9bf79c337ecded449db08ebc3ff1fb
```

```{r plot-check}
grade_this({
<<<<<<< HEAD
  if (ith_geom_is(.result, "function", i = 1)) {
    pass(random_praise())
  }
  fail(random_encouragement())
=======
    pass_if(ith_geom_is(.result, "function", 1)) 
    fail(random_encouragement())
>>>>>>> 89e5e3d9ff9bf79c337ecded449db08ebc3ff1fb
})
```

Add the line $y = 20x$ to the same plot in \textcolor{darkorange}{*dark orange*} by redefining `p` as `p + 'new function layer'`. 

```{r addPlot, exercise=TRUE, fig.height=6.5, fig.width=4.5, fig.align='center'}

```

<<<<<<< HEAD
```{r addPlot-setup}
=======
```{r addPlot-setup, warning=FALSE, message = FALSE}
>>>>>>> 89e5e3d9ff9bf79c337ecded449db08ebc3ff1fb
p <- ggplot() +
  geom_function(fun = function(x) 8*x^2 - 6*x - 12, color = "blue") 
```

<<<<<<< HEAD
```{r addPlot-solution}
=======
```{r addPlot-solution, warning=FALSE, message = FALSE}
>>>>>>> 89e5e3d9ff9bf79c337ecded449db08ebc3ff1fb
p <- p +
    geom_function(fun = function(x) 20*x, color = 'darkorange')
```

```{r addPlot-check}
grade_this({
  if (ith_geom_is(.result, "function", i = 2)) {
<<<<<<< HEAD
    pass(random_praise(), " Where is the second function though? Maybe we should fixe the scales of the axes.")
=======
    pass(random_praise())
>>>>>>> 89e5e3d9ff9bf79c337ecded449db08ebc3ff1fb
  }
  fail(random_encouragement())
})
```

Do the following to improve the appearance of the plot:  
1. Edit the scales of the axes to go from $x = -2$ to $x = 4$ and from $y = -25$ to $y = 75$ using `lims()`.  
2. Format the axes to read as $x$ $f(x)$ using labels

```{r lims, exercise=TRUE, fig.height=6.5, fig.width=4.5, fig.align='center'}

```

<<<<<<< HEAD
```{r lims-setup}
=======
```{r lims-setup, warning=FALSE, message = FALSE}
library(ggplot2)
>>>>>>> 89e5e3d9ff9bf79c337ecded449db08ebc3ff1fb
p <- ggplot() +
  geom_function(fun = function(x) 8*x^2 - 6*x - 12, color = "blue") +
  geom_function(fun = function(x) 20*x, color = 'darkorange')
```

<<<<<<< HEAD
```{r lims-solution}
=======
```{r lims-solution, warning=FALSE, message = FALSE}
>>>>>>> 89e5e3d9ff9bf79c337ecded449db08ebc3ff1fb
p <- p +
  lims(x = c(-2, 4), y = c(-25, 75)) + 
  labs(x = expression(italic(x)), y = expression(italic(f) ( italic(x) )))
```

```{r lims-check}
grade_this({
  if (gradethis_equal(.result, .solution)) {
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

There are also loads of other ways to make beautiful graphs using the package `ggplot2` using custom themes for the plot area, titles, and other options. This class focuses on solutions and basic plots of those solutions, for which `base R` plots more than suffice.

## Optimization
<<<<<<< HEAD

### Ocular Estimation
=======
>>>>>>> 89e5e3d9ff9bf79c337ecded449db08ebc3ff1fb

### Ocular Estimation

If we want to try to "eyeball" the minimum of our quadratic function, we can zoom in on the range $0.25 \le x \le 0.5$. Add new `lims()` to our plot, `p`, to graph only this portion of the domain of our function. Also set the y-axis `lims(0)` to a reasonable range to allow us to see how y changes along this domain. 

```{r zoom, exercise = TRUE, fig.height=6.5, fig.width=4.5, fig.align='center'}

```

<<<<<<< HEAD
```{r zoom-solution}
ggplot() +
  geom_function(fun = function(x) 8*x^2 - 6*x - 12) +
  lims(x = c(0.25, 0.5))
=======
```{r zoom-hint}
# A good range for the y-axis might be -13.25 to -13. Experiment with ranges that you think will help others see the minimum. 
```

```{r zoom-setup, warning=FALSE, message = FALSE}
library(ggplot2)
p <- ggplot() +
  geom_function(fun = function(x) 8*x^2 - 6*x - 12, color = "blue") +
  geom_function(fun = function(x) 20*x, color = 'darkorange') +
  lims(x = c(-2, 4), y = c(-25, 75)) + 
  labs(x = expression(italic(x)), y = expression(italic(f) ( italic(x) )))
```
```{r zoom-solution, warning=FALSE, message = FALSE}
p + lims(x = c(0.25, 0.5), y = c(-13.15, -13))
>>>>>>> 89e5e3d9ff9bf79c337ecded449db08ebc3ff1fb
```

```{r zoom-check}
grade_this({
  if (gradethis_equal(.result, .solution)) {
    pass(random_praise())
  }
  fail(random_encouragement())
<<<<<<< HEAD
})
```

It looks like the minimum is somewhere around $0.37$ or $0.38$. Next, let's use *calculus* get a more precise answer. 

### Analytic Solution

R can actually apply standard analytic solution algorithms to find the functional form for a derivative of a function using the `deriv` function available in the `Ryacas` package. Use `deriv` to find the general derivative of $$f(x) = ax^2 + bx + c$$ and the example with the specific parameter values, $f(x) = 8x^2 - 6x - 12$, with respect to `x`. 

```{r Deriv, exercise = TRUE}

```

```{r Deriv-solution}
deriv(ysym('a*x^2 + b*x + c'), 'x')
deriv(ysym('8*x^2 - 6*x - 12'), 'x')
```

```{r Deriv-check}
grade_this({if (!inherits(.result, c("yac_symbol"))) {
    fail("Your class of your answer should be a `yac_symbol` list")}
  if (.result$yacas_cmd != "16*x-6") { 
    fail("The derivative should be equal to '16*x-6'") }
  pass()
=======
>>>>>>> 89e5e3d9ff9bf79c337ecded449db08ebc3ff1fb
})
```

It looks like the minimum is somewhere around $0.37$ or $0.38$. Next, let's use *calculus* get a more precise answer. 

### Computational Solution

Some optimization problems are more complicated. For these, it might help to find the optimum using computational methods. These functions basically take an initial value of `x`, calculate the change in the function to see if it is increasing or decreasing, select a new `x` a little bit in the appropriate direction (decreasing direction for mimima; increasing direction for maxima), rinse, and repeat until the change in `x` is very, *very* small. 

Use the `solnp()` function from the `Rsolnp` package to find the minimum of $f(x) = 8x^2 - 6x - 12$.

```{r solnp, exercise = TRUE}

```

```{r solnp-hint1}
# You must define the function argument in the same way you defined it for `geom_function()` using `function(x) {expression}`.
```

```{r solnp-hint2}
# `solnp` can be finicky about its starting value, and sometimes gets confused when you choose zero as a starting value. Try almost any other value. 
```

<<<<<<< HEAD
```{r solnp-solution}
=======
```{r solnp-solution, warning=FALSE, message = FALSE}
>>>>>>> 89e5e3d9ff9bf79c337ecded449db08ebc3ff1fb
library(Rsolnp)
solnp(pars = 1, fun = function(x) 8*x^2 - 6*x - 12)
```

```{r solnp-check}
<<<<<<< HEAD
grade_this({if (!inherits(.result, c("list"))) {
=======
grade_this({
  if (!inherits(.result, c("list"))) {
>>>>>>> 89e5e3d9ff9bf79c337ecded449db08ebc3ff1fb
    fail("Your class of your answer should be a list")}
  if (round(.result$pars, 3) != 0.375) { 
    fail("The solution should be equal to 0.375.") }
  pass()
})
```

### Constrained Optimization

Most of the examples we explore in economics involve *constrained* optimization, such as when a consumer seeks to maximize utility *subject to a budget constraint* or when a firm seeks to minimize costs *subject to a production target*. Several packages in R perform constrained optimization. We will use the `solnp` function from the `Rsolnp` package for this type of problem. We will practice this type of problem later. Here is how you could use `solnp` to solve a cost minimization problem.  
$$min_{L, K} {wL + rK} \\
\text{s.t.} \\
LK \ge 100 \\
L \ge 0 \\
K \ge 0 \\
\text{where} \\
w = r = 10$$

<<<<<<< HEAD
```{r solnpC, exercise = TRUE}
solnp(pars = 0,                            # Set initial guess for the solution
=======
```{r constrainedSolnp, echo = TRUE}
library(Rsolnp)
solnp(pars = c(0,0),                       # Set initial guess for the solution
>>>>>>> 89e5e3d9ff9bf79c337ecded449db08ebc3ff1fb
      fun = function(x) 10*x[1] + 10*x[2], # Define the objective function 
      ineqfun = function(x) x[1]*x[2],     # Define the constraint function
      ineqLB = 100,                        # Set the lower bound of the constraint 
      ineqUB =  Inf,                       # Set the upper bound of the constraint
      LB = c(0,0))                         # Set the lower bound for the variables
<<<<<<< HEAD
```

```{r solnpC-solution}
solnp(pars = 0,                            # Set initial guess for the solution
      fun = function(x) 10*x[1] + 10*x[2], # Define the objective function 
      ineqfun = function(x) x[1]*x[2],     # Define the constraint function
      ineqLB = 100,                        # Set the lower bound of the constraint 
      ineqUB =  Inf,                       # Set the upper bound of the constraint
      LB = c(0,0))                         # Set the lower bound for the variables
```

```{r solnpC-check}
grade_this({if (!inherits(.result, c("list"))) {
    fail("Your class of your answer should be a list")}
  if (round(.result$pars, 3) != 0.375) { 
    fail("The solution should be about 0.375") }
  pass("Notice how computational solutions sometimes give very slightly less precise solutions.")
})
=======
>>>>>>> 89e5e3d9ff9bf79c337ecded449db08ebc3ff1fb
```

## Solving a System of Equations: Supply and Demand

Given the following (inverse) supply and demand functions: 

$$P = 10 - Q_d$$

$$P = Q_s$$

To solve a system of equations (linear or nonlinear) the `nleqslv` function from the `nleqslv` package is a useful tool that requires the following arguments: 

* x: A numeric vector with an initial guess of the solution (similar to `pars`)
* fn: A function of x that equals zero at the solution and returns a vector of function values with the same length as the vector x (i.e. to solve for two variables you need two functions). 

Solve for the equilibrium price and quantity in the example using `nleqslv`. 

```{r equilibrium, exercise = TRUE}

```

<<<<<<< HEAD
```{r equilibrium-hint1}
# A `function(x)` can be a vector that includes multiple equations. 
```

```{r equilibrium-hint2}
# `nleqslv` looks to set the values of `function(x)` to zero, so you might need to rearrange them.
```

```{r equilibrium-solution}
=======
```{r equilibrium-hint}
# A `function(x)` can be a vector that includes multiple equations. 
# `nleqslv` looks to set the values of `function(x)` to zero, so you might need to rearrange the equations.
```

```{r equilibrium-solution, warning=FALSE, message = FALSE}
>>>>>>> 89e5e3d9ff9bf79c337ecded449db08ebc3ff1fb
library(nleqslv)
nleqslv(c(1,1), function(x) c(x[1] - x[2], 10 - x[1] - x[2]))
```

```{r equilibrium-check}
grade_this({
  if (
    round(.result$x[1], 0) != round(.solution$x[1], 0) & 
    round(.result$x[2], 0) != round(.solution$x[2], 0)) {
    fail("The solution should be about (5, 5).") }
  pass()
})
```

(Note: This is a pretty easy system to solve; the objective is to code the solution "correctly.")

## Appendix 

### Basic Math Function Rules

$$(x^a)^b = x^{ab}$$
$$x^ax^b = x^{a+b}$$
$$\frac{x^a}{x^b} = x^{a-b}$$
$$log_b(xy) = log_b(x) + log_b(y)$$
$$log_b\bigg(\frac{x}{y}\bigg) = log_b(x) - log_b(y)$$
$$log_b(x^a) = alog_b(x)$$ 

### Basic Calculus Rules

$$\frac{d}{dx}\big(x^a\big) = ax^{a-1}$$
$$\frac{d}{dx}\bigg(ln(x)\bigg) = \frac{1}{x}$$  

### Analytic Optimization

R can actually apply standard analytic solution algorithms to find the functional form for a derivative of a function using the `deriv` function available in the `Ryacas` package. Use `deriv` to find the general derivative of $$f(x) = ax^2 + bx + c$$ and the example with the specific parameter values, $f(x) = 8x^2 - 6x - 12$, with respect to `x`. 

```{r Deriv, echo = TRUE}
deriv(ysym('a*x^2 + b*x + c'), 'x')
deriv(ysym('8*x^2 - 6*x - 12'), 'x')
```

`Ryacas` can also format the result with tex syntax which you can call in a document to produce a formatted result as shown below: 

$$`r tex(deriv(ysym('a*x^2 + b*x + c'), 'x'))`$$

<details><summary>Click here to see the code that generated this formatted equation.<p></p></summary>
\$\$`` ` r tex(deriv(ysym('a*x^2 + b*x + c'), 'x'))` ``\$\$`<br><br>
</details>

To find the minimum of the function, notice that to the left of the minimum, the function is *decreasing* (it has a negative derivative); whereas, to the right of the minimum, the function is *increasing* (has a positive derivative). Thus, the exact value of the minimum must occur where the derivative equals *zero*. The algebra of finding the exact solution in general form, $x = -b/2a$, or the specific example, $x = 3/8$ is pretty straightforward. 

### A Note on "Piping"

In our optimization example, we nested a `ysym` function (to convert a text equation into a `yac_symbol`). Suppose we wanted to solve the derivatives we just derived. We *could* nest the result of the derivative inside `solve` by running the code below: 

```{r nested, echo = TRUE}
solve(deriv(ysym('a*x^2 + b*x + c'), 'x'), 'x')
```

Alternatively, we could use a coding method called "piping," where you pass one object along as the `x` for another function to evaluate using the `|>` operator after each command before starting a new line (although starting a new line is not even required, either). This allows you to specify the commands you want to execute *in the order you want to execute them* (rather than needing to have the last execution on the outside, or first). Run the following code to compare the result of the two styles of coding.

```{r piped, echo = TRUE}
'a*x^2 + b*x + c' |> # Start with an object, function, or expression
  ysym() |>          # Type the first function you want to do on that object
  deriv('x') |>      # Now the second function  (you may add options)...
  solve('x')         # And so on...
```

Sometimes this can yield faster results, and some users find it more intuitive and easier to read. 
