---
title: "2.1 Production and Production Functions"
author: Jim Bang
output: 
  learnr::tutorial: 
    progressive: true
runtime: shiny_prerendered
---

<style type="text/css">
h1{font-size: 24pt}
h2{font-size: 20pt}
h3{font-size: 18pt}
h4,h5,h6{font-size: 16pt}
body{font-size: 16pt}
</style>

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(Ryacas)
library(Deriv)
gradethis_setup(exercise.reveal_solution = TRUE)
knitr::opts_chunk$set(echo = FALSE, fig.height = 6, fig.width = 6)
yac('MaxEvalDepth(100000)')
```

## Firm Decisions

### Three (Related) Decisions:

1. Cost-Minimizing Combination of Inputs: For a *given* level of (expected) output (firm demand) choose the *combination* of inputs that minimizes the cost of producing that amount to derive the cost function.
2. Profit-Maximizing Level of Output: Choose the *level* of output that maximizes profits by adjusting quantity and/or price of the output. 
3. Profit-Maximizing Level of Inputs: Choose the *levels* of all inputs that maximizes profits by adjusting the quantity and/or price for inputs. 

### Components

1. Production Technology
2. Resource Constraints
3. Input/Output Choices

### Complications

1. Time Horizon
a. Short Run
b. Long Run
c. Sunk versus Fixed Costs
2. Market Structure
a. Price Taker
b. Price Maker
c. In-Between
3. Transactions Costs
a. Negotiation
b. Monitoring
c. Enforcement

## Theory of the Firm

Firms suppress market transactions. 

Why do they exist? 

### Classical/Neoclassical Approach "Black Box"

Transform inputs into outputs

```{r, fig.height = 1, fig.width = 5}
par(mar = c(1, 1, 1, 1))
plot.new()
plot.window(xlim = c(0,1), ylim = c(0,0.1))
text(0.05, 0.05, "Hire\nInputs")
polygon(x = c(0.4, 0.6, 0.6, 0.4), y = c(0, 0, 0.1, 0.1), col = 'black')
arrows(x0 = 0.1, y0 = 0.05, x1 = 0.4, y1 = 0.05)
text(0.5, 0.05, "???", col = 'white')
arrows(x0 = 0.6, y0 = 0.05, x1 = 0.9, y1 = 0.05)
text(0.95, 0.05, "Sell\nOutputs")
```

<details style="line-height:105%"><summary>Click here to see the code for the above figure.</summary>

`par(mar = c(1, 1, 1, 1))`<br>
`plot.new()`<br>
`plot.window(xlim = c(0,1), ylim = c(0,0.1))`<br>
`text(0.05, 0.05, "Hire\nInputs")`<br>
`polygon(x = c(0.4, 0.6, 0.6, 0.4), y = c(0, 0, 0.1, 0.1), col = 'black')`<br>
`arrows(x0 = 0.1, y0 = 0.05, x1 = 0.4, y1 = 0.05)`<br>
`text(0.5, 0.05, "???", col = 'white')`<br>
`arrows(x0 = 0.6, y0 = 0.05, x1 = 0.9, y1 = 0.05)`<br>
`text(0.95, 0.05, "Sell\nOutputs")`<br>

</details>

### Bureaucratic Hierarchy

Manage jointly-owned resources

### "Nexus of Contracts"

Incentivize productivity and specialization within and between organizations

## The Production Function

$$y = f(L,K),\text{ where}$$
$$y\text{ is the amount of output the firm produces}$$ 
$$L\text{ is the amount of labor the firm hires}$$ 
$$K\text{ is the amount of capital the firm hires}$$ 
$$f()\text{ describes the technology for transforming inputs (L and K) into output (y)}$$ 

### Basic Properties 

1. Monotonic
2. Diminishing Returns

(Similar properties to utility functions; transitivity is trivial in industries where physical output is cardinally measurable.)

### Cobb-Douglas

Since production functions exhibit many of the same properties as utility functions, the Cobb-Douglas function is a common functional form for examples. In the context of production with labor and capital inputs, the Cobb-Douglass (production) function is: 

$$ y = AL^{\alpha}K^{(1-\alpha)},\text{ where}$$

$$\alpha\text{ is labor's productivity share}$$
$$A\text{ represents "total factor productivity," or the overall level of technology}$$

### Other Examples

1. Constant Elasticity of Substitution
$$U(x) = (L^r + K^r)^{\frac{1}{r}},\text{ where}$$ 

$$s=\frac{1}{1-r} \text{ is the elasticity of substitution.}$$ 

2. Polynomial (Short-Run Only?)
$$y = aL^3 + bL^2 + cL,\text{ where}$$
$$a<0$$ 
$$b>0$$ 
$$c>0$$ 
### Graphical Representation

Use the `persp(x, y, z) function in r to graph the Cobb-Douglas production function for $A = 100$, $0 \le L \le 50$, $0 \le K \le 50$, and $\alpha = 1/3$.

```{r CobbDouglas, exercise=TRUE, exercise.reveal_solution = TRUE}

```

```{r CobbDouglas-solution}
x = seq(0, 10, 0.1)
y = seq(0, 10, 0.1)
z = outer(100*x^(1/3), y^(2/3))
persp(x,
      y,
      z,
      xlab = "Labor",
      ylab = "Capital",
      zlab = "y = f(L,K)")
```

```{r CobbDouglas-check}
grade_this({if (!inherits(.result, c("matrix"))) {
    fail("Your class of your answer should be a matrix")}
  if (.result[,1] != c(0.2, 0, 0, -1)) { 
    fail("The first column should equal (0.2, 0, 0, -1.0).") }
  pass()
})
```

## Short-Run Production Function

In the short run at least one input is fixed (usually capital and/or land). In our two-input production function this makes output a function of only labor.  

Graphically, this is like slicing a vertical cross-section of the production function at a specific level of the fixed input (capital). 

### Cobb-Douglas Production Function

Plot production as a function of labor for $0 \le L \le 50$ and $K = 25$.

```{r cdShortRun, exercise=TRUE, exercise.reveal_solution = TRUE}

```

```{r cdShortRun-solution}
curve(100*25^(2/3)*x^(1/3),
  type = 'l',
  xlim = c(0, 50),
  xlab = "Labor",
  ylab = "Capital"
)
```

```{r cdShortRun-check}
grade_this({if (!inherits(.result, c("list"))) {
    fail("Your class of your answer should be a list")}
  if (round(.result$y[2], 4) != 678.6044) { 
    fail("The second y value should equal 678.6044.") }
  pass()
})
```

### Polynomial Production Function

Define and graph an expression for the following short-run production function: 

$$y = -L^3 + 60L^2 + 25L$$

```{r polynomial, exercise = TRUE, exercise.reveal_solution = TRUE}

```

```{r polynomial-solution}
polynomial <- function(x) {
  -x^3 + 60*x^2 + 25*x
}
curve(polynomial, 
      type = 'l', 
      xlim = c(0, 50), 
      xlab = "Labor",
      ylab = "Output"
)
```

```{r polynomial-check}
grade_this({if (!inherits(.result, c("list"))) {
    fail("Your class of your answer should be a list")}
  if (round(.result$y[2], 3) != 27.375) { 
    fail("The second y value should equal 27.375") }
  pass()
})
```

The *marginal product of labor* is the *change* in (i.e. slope, derivative of) the total production function with respect to $L$. 

The *average product of labor* is the amount (i.e. level) of output *per-worker*. 

```{r MPLvAPL, echo=FALSE}
question("When the marginal product curve reaches its maximum, the average product curve",
    answer("also reaches its maximum"),
    answer("lies above the marginal product curve "),
    answer("continues to rise", correct = TRUE),
    answer("is negative"),
    allow_retry = TRUE
  )
```

## Marginal and Average Product Product of Labor

### Marginal Product of Labor

The marginal product of labor is the rate of change in output for a given change in labor hired. 

$$MP_L = \frac{df(L,K)}{dL}$$

Graphically, the marginal product is the slope of the production function, holding the level of capital fixed. 

### Average Product

$$AP_L = \frac{f(L,K)}{L}$$

### Graphical Representation

Add `curve()` functions to the following code to graph the marginal and average products of labor from $0 \le L \le 50$ in a single plot area. Use $(-1250, 1250)$ as the limits for the vertical (output) axis and label your axes "Labor" and "Output." 

```{r MPL, exercise = TRUE, exercise.reveal_solution = TRUE}
yac.polynomial <- 
  deparse(polynomial)[[3]] |>   # Split polynomial as character vector, position 3
  ysym()                        # Convert to a yac_sym
dl.polynomial <- yac.polynomial |>
  deriv('x') |>                 # Take the derivative with respect to x
  as_r()                        # Convert to an r expression
yl.polynomial <- 
  yac.polynomial / ysym('x')    # Divide by x
yl.polynomial <- yl.polynomial |>
  simplify() |>
  as_r()                        # Convert to an r expression
polynomial.mpl <- function(x) {
  eval(dl.polynomial[[1]])
}                             # Define the mpl as a function. 
polynomial.apl <- function(x) {
  eval(yl.polynomial[[1]])
}                             # Define the apl as a function. 
# Begin Graphing Code Below. 

```

```{r MPL-solution}
yac.polynomial <- 
  deparse(polynomial)[[3]] |>   # Split polynomial as character vector, position 3
  ysym()                        # Convert to a yac_sym
dl.polynomial <- yac.polynomial |>
  deriv('x') |>                 # Take the derivative with respect to x
  as_r()                        # Convert to an r expression
yl.polynomial <- 
  yac.polynomial / ysym('x')    # Divide by x
yl.polynomial <- yl.polynomial |>
  simplify() |>
  as_r()                        # Convert to an r expression
polynomial.mpl <- function(x) {
  eval(dl.polynomial[[1]])
}                             # Define the mpl as a function. 
polynomial.apl <- function(x) {
  eval(yl.polynomial[[1]])
}                             # Define the apl as a function. 
# Begin Graphing Code Below. 
curve(
  polynomial.mpl,
  type = 'l',
  xlim = c(0, 50),
  ylim = c(-1250, 1250),
  xlab = "Labor",
  ylab = "Output"
)
curve(polynomial.apl, add = TRUE)
```

```{r MPL-check}
grade_this({if (!inherits(.result, c("list"))) {
    fail("Your class of your answer should be a list")}
  if (round(.result$y[2], 3) != 25.6) { 
    fail("The second y value should equal 25.6") }
  pass()
})
```

## Production in the Long Run

Flexibility of input choices

### Isoquants and their Properties

Based on the preference axioms introduced earlier, we can derive three properties for indifference curves.

1.  Isoquants farther from the origin represent *greater output* (monotonicity);
2.  Isoquants *do not cross* (transitivity);
3.  Isoquants slope *downward* (monotonicity + diminishing returns);
4.  Isoquants are *bowed inward* (convexity - a mix of inputs is usually more productive).

Graphically, an isoquant is a horizontal cross-section of the production function at a fixed level of output. 

For the Cobb-Douglass example with $\alpha = 1/3$ and $A = 100$, plot the isoquants for $y_1 = 1000$ and $y_2 = 2500$ for $0 \le L \le 50$ and $0 \le K \le 50$.

```{r cdIsoquants, exercise=TRUE, exercise.reveal_solution = TRUE}

```

```{r cdIsoquants-solution}
# y.cobbDouglas <- ysym('100*L^0.25 * K^(1-0.25)')
# y.isoquant1 <- solve(y.cobbDouglas, 1000, 'K')
curve((1000 / (100*x^(1/3)))^(3/2),
  type = 'l',
  ylim = c(0, 50),
  xlim = c(0, 50),
  xlab = "Labor",
  ylab = "Capital"
)
curve((2500 / (100*x^(1/3)))^(3/2), add = TRUE)
```

```{r cdIsoquants-check}
grade_this({if (!inherits(.result, c("list"))) {
    fail("Your class of your answer should be a list")}
  if (round(.result$y[2], 3) != 176.777) { 
    fail("The second y value should equal 176.777") }
  pass()
})
```

### Marginal Rate of Technical Substitution

The *marginal rate of technical substitution* between two inputs (e.g. labor and capital, $MRTS_{LK}$) is the ratio of the marginal productivities, $\frac{MP_L}{MP_K}$. The MRTS tells us the firm's ability (through the available production technology) to exchange capital for labor on the margin in such a way that the firm produces the same total amount of output. Hence, the MRS also represents the slope of the isoquant curve through any given values of $L$ and $K$. 

Use the `Simplify` function from the `Deriv` package to get $MRTS_{LK} = \frac{MP_L}{MP_K}$ for the previous exercise. Remember that marginal utility results stored as `yac_symbol` objects, so you need to convert them to strings before applying `Simplify`, and then finally converting it back to a `ysym`. Try using piping to keep it readable! <br>(Note: `Ryacas` has a `simplify` function, but it just doesn't work as well for the functions we're trying to simplify.)

```{r marginalRateOfSubstitution, exercise = TRUE}

```

```{r marginalRateOfSubstitution-solution}
y <- ysym('100 * L^(1/3) * K^(2/3)')
MRTS <- (deriv(y, 'L') / deriv(y, 'K')) |>
  yac_str() |> 
  Simplify() |>
  ysym()
```

```{r marginalRateOfSubstitution-check}
grade_this({if (!inherits(.result, c("yac_symbol"))) {
    fail("Your class of your answer should be a yac_symbol")}
  if (.result != 'y: K/(2*L)') { 
    fail("The MRTS should equal K/(2*L)") }
  pass()
})
```

### General Form

```{r}
q <- ysym('A * L^(alpha) * K^(1-alpha)')
MRTSq <- (deriv(q, 'L') / deriv(q, 'K')) |>
  yac_str() |> 
  Simplify() |>
  ysym()
```

$$`r tex(MRTSq)`$$

## Returns to Scale

Let's see what happens when we scale up *all* of the inputs in the production function. 

Evaluate the Cobb Douglas production function from our example for $(L,K) = {(10,10), (20,20), (30,30)}$. (Hint: create a new function of `L` and `K` called `f` that converts the `yac_symbol` `y` into an `r` expression \[`yac_expr'\] and evaluates it \[`eval`\].)

```{r RTS, exercise = TRUE, exercise.reveal_solution = TRUE}
f <- function(L, K) {eval(yac_expr(y))}

```

```{r RTS-solution}
f <- function(L, K) {eval(yac_expr(y))}
f(10, 10)
f(20, 20)
f(30, 30)
```

```{r RTS-check}

```

```{r double, echo=FALSE}
question("What happens when we double the inputs in the Cobb-Douglas function?",
    answer("Output more than doubles"),
    answer("Output doubles", correct = TRUE),
    answer("Output increases by less than double"),
    answer("Output decreases"),
    allow_retry = TRUE
  )
```

### Constant Returns to Scale

A production technology has *constant returns to scale* whenever scaling up all inputs by the same proportion (e.g. double) increases output by the same proportion (e.g. double). 

Examples of constant returns to scale production functions: 

1. Cobb-Douglas $y = AL^{\alpha}K^{1-\alpha}$
2. Constant Elasticity of Substitution $y = (aL^r + bK^r)^{1/r}$
3. Linear (Perfect Substitutes) $y = aL + bK$
4. Perfect Complements (Linear Expansion Path) $y = min\{aL, bK\}$

Isoquants for a constant-returns technology are evenly spaced as the firm chooses to expand production.

Production functions with constant returns to scale correspond to constant long-run average costs. 

### Decreasing Returns to Scale

A production technology has *decreasing returns to scale* whenever scaling up all inputs by the same proportion (e.g. double) increases output by less than that proportion (e.g. less than double). 

```{r isoquant, echo=FALSE}
question("What happens to the spacing of the isoquants as output expands by equal amounts for a decreasing-returns production technology?",
    answer("They get closer together."),
    answer("They remain equally spaced."),
    answer("They get farther apart.", correct = TRUE),
    answer("The spacing cannot be determined with the given information."),
    allow_retry = TRUE
  )
```

Production functions with *decreasing* returns to scale correspond to *increasing* long-run average costs (diseconomies of scale). 

### Increasing Returns to Scale

A production technology has *increasing returns to scale* whenever scaling up all inputs by the same proportion (e.g. double) increases output by more than that proportion (e.g. more than double). 

Production functions with *increasing* returns to scale correspond to *decreasing* long-run average costs (economies of scale). 
