---
title: "Utility Maximization in R"
author: "James Bang"
date: "6/3/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Analytic Solution

### Equilibrium

The following code finds the first-order conditions for a utility maximization problem involving a two-good Cobb-Douglass utility function: 

$$\max_xU(x_1, x_2) = x_1^{\alpha}x_2^{1-\alpha}$$
$$s.t.: p_1x_1+p_2x_2 \le I$$

The "Lagrangean" function for this problem is: 

$$\mathcal{L} = x_1^{\alpha}x_2^{1-\alpha} + \lambda (I - p_1x_1 - p_2x_2)$$

The `Deriv` function in the `Deriv` package "solves" basic algorithmic derivatives (i.e. memorizes the rules so you don't have to): 

```{r fml}
library(Deriv)
L <- 'x_1^(alpha) * x_2^(1-alpha) + lambda*(I - p_1*x_1 - p_2*x_2)'
(fml <- c(Deriv(L, 'x_1'), Deriv(L, 'x_2'), Deriv(L, 'lambda')))
```

Using various "wildcard" characters and other "regular expression" (regex) syntax, I can apply some substitutions in this formula to make the output values more readable as equations for typesetting in Rmarkdown: 

```{r gsubs}
fmlrmd <- gsub('(\\d*\\.\\d*)', '{\\1}', fml)
fmlrmd <- gsub('(\\(.)', '{\\1', fmlrmd)
fmlrmd <- gsub('(.))', '\\1)}', fmlrmd)
fmlrmd <- gsub('(\\))($)', '\\1)}\\2', fmlrmd)
fmlrmd <- gsub('\\*', '', fmlrmd)
fmlrmd <- gsub('alpha', '\\\\alpha', fmlrmd)
(fmlrmd <- gsub('lambda', '\\\\lambda', fmlrmd))
```

I can reference the output of the (edited) derivatives inline within the equation editor using \$`` ` r command` ``\$. Notice the use of backquotes inside the inline equation escape character (\$). To produce the following full-line equation I used \$\$`` ` r fmlrmd[1]` ``\$\$. Here are the first-order conditions:
$$`r fmlrmd[1]` = 0 \text{ (w.r.t. good 1)}$$
$$`r fmlrmd[2]` = 0 \text{ (w.r.t. good 2)}$$

$$`r fmlrmd[3]` = 0 \text{ (w.r.t. }\lambda)$$

Next, we can define the marginal rate of substitution as $MU_1/MU_2$. The `Simplify` function in `Deriv` helps clean things up. 

```{r mrs}
mu1 <- gsub('( - lambda)(.*)', '', fml[1])
mu2 <- gsub('( - lambda)(.*)', '', fml[2])
mrs <- paste0('(', mu1, ')', '/', '(', mu2, ')')
(mrs <- Simplify(mrs))
```

Accordingly, we can substitute various brackets and escape characters to allow translation to Markdown equation syntax:

```{r mrsrmd}
mrsrmd <- gsub('(\\d*\\.\\d*)', '{\\1}', mrs)
mrsrmd <- gsub('(\\(.)', '{\\1', mrsrmd)
mrsrmd <- gsub('(.))', '\\1)}', mrsrmd)
mrsrmd <- gsub('(\\))($)', '\\1}\\2', mrsrmd)
mrsrmd <- gsub('\\*', '', mrsrmd)
mrsrmd <- gsub('alpha', '\\\\alpha', mrsrmd)
(mrsrmd <- gsub('lambda', '\\\\lambda', mrsrmd))
```

$$`r mrsrmd`$$

Note that in the (linearly-homogeneous) Cobb-Douglass example, setting $MRS_{1,2} = p_1/p_2$, we get the intuition that the preference parameter $\alpha$ represents the budget share of good 1, or: 

$$\frac{\alpha}{1-\alpha} = \frac{p_1x_1}{p_2x_2}$$
This, combined with the budget constraint gives a linear system of two equations and two unknowns by: 

$$x_1 = \frac{\alpha p_2x_2}{(1-\alpha)p_1} \text{ (from MRS)}$$
$$p_1x_1 + p_2x_2 = I \text{ (Budget Constrant)}$$

Demand for good 1 is given by: 

$$x_1 = \frac{\alpha I}{p_1}$$
Demand for good 2 is given by: 

$$x_2 = \frac{(1 - \alpha) I}{p_2}$$

### Comparative Statics

Next, we can analyze what happens when various parameters (prices, income, budget-share preferences) change.

```{r cs}
d1 <- 'alpha * I/p_1'
d.d1 <- c(Deriv(d1, 'alpha'), Deriv(d1, 'I'), Deriv(d1, 'p_1'))

```


## Numerical example

Let's try solving a numerical example with specific parameters. Let $\alpha = 0.25$, $p_1 = 1.5$, $p_2 = 1$, and $I = 100$. 

$$\max_xU(x_1, x_2) = x_1^{0.25}x_2^{0.75}$$
$$s.t.: 1.5x_1+x_2 \le 100$$

This yields the following Lagrangean: 

$$\mathcal{L} = x_1^{0.25}x_2^{0.75} + \lambda (100 - 1.5x_1 - x_2)$$

### Equilibrium

Let's orient ourselves to the solution using the analytical derivative.
```{r fml2, echo = FALSE}
L <- 'x_1^(0.25) * x_2^(0.75) + lambda*(100 - 1.5*x_1 - x_2)'
fml <- c(Deriv(L, 'x_1'), Deriv(L, 'x_2'), Deriv(L, 'lambda'))
fmlrmd <- gsub('(\\d*\\.\\d*)', '{\\1}', fml)
fmlrmd <- gsub('(\\(.)', '{\\1', fmlrmd)
fmlrmd <- gsub('(.))', '\\1)}', fmlrmd)
fmlrmd <- gsub('\\*', '', fmlrmd)
fmlrmd <- gsub('alpha', '\\\\alpha', fmlrmd)
fmlrmd <- gsub('lambda', '\\\\lambda', fmlrmd)
mu1 <- gsub('( - lambda)(.*)', '', fml[1])
mu2 <- gsub('( - lambda)(.*)', '', fml[2])
mrs <- paste0('(', mu1, ')', '/', '(', mu2, ')')
mrs <- Simplify(mrs)
mrsrmd <- gsub('(\\d*\\.\\d*)', '{\\1}', mrs)
mrsrmd <- gsub('(\\(.)', '{\\1', mrsrmd)
mrsrmd <- gsub('(.))', '\\1)}', mrsrmd)
mrsrmd <- gsub('\\*', '', mrsrmd)
mrsrmd <- gsub('alpha', '\\\\alpha', mrsrmd)
mrsrmd <- gsub('lambda', '\\\\lambda', mrsrmd)
```

The first-order conditions are:

$$`r fmlrmd[1]` = 0$$
$$`r fmlrmd[2]` = 0$$
$$`r fmlrmd[3]` = 0$$

The marginal rate of substitution is: 
$$`r mrsrmd`$$
The solution follows from $MRS_{1,2} = p_1/p_2$ 

$$\frac{0.25}{0.75} = \frac{1.5x_1}{x_2},$$
and the budget constraint:
$$100 = 1.5x_1 + x_2.$$

### Numerical solution

```{r}
u <- function(x) {
  x[1] ^ (alpha) * x[2] ^ (1 - alpha)
}
alpha = 0.25
p_1 = 1.5
p_2 = 1
I = 100
equilibrium <-
  constrOptim(
    c(0.01, 0.01),
    u,
    NULL,
    ui = c(-p_1,-p_2),
    ci = c(-I),
    control = list(fnscale = -1)
  )
equilibrium$lambda <-
  alpha * equilibrium$par[1] ^ (alpha - 1) * equilibrium$par[2] ^ (1 - alpha) / p_1
equilibrium[c(1, 2, 8)]
```

Note that the intuition from the analytical solution tells us that the "true" answer is $x_1 = 50/3 \approx 16.667$ and $x_2 = 75$ to ensure that this consumer spends 25% of their planned budget for the two goods on good 1, or one-third as much as they spend on good 2. Computational methods only give approximate answers, so it's good to double check! 

We can also plot the equilibrium.

```{r plot, fig.height=6.5, fig.width=4.5, fig.align='center'}
curve(
  equilibrium$value ^ (1 / (1 - alpha)) / (x ^ (alpha / (1 - alpha))),
  0,
  I / p_1,
  xlab = expression(x[1]), 
  ylab = expression(x[2]),
  xlim = c(0, I / p_1),
  ylim = c(0, I / p_2),
  col = 'dark green'
)
abline(
  I / p_2,
  -p_1 / p_2,
  col = 'red'
)
points(equilibrium$par[1],
       equilibrium$par[2],
       col = 'blue',
       pch = 8)

```

### Comparative Statics
