---
title: "2.2 Cost Minimization"
output: 
  learnr::tutorial: 
    progressive: true
runtime: shiny_prerendered
author: 
  name: "Created by [Jim Bang](http://www.github.com/bangecon)"
  email: BangJamesT@sau.edu
  affiliation: "[St. Ambrose University](http://www.sau.edu)"
date: "`r format(Sys.Date(), '%d %b %Y')`"
---

<style type="text/css">
#doc-metadata {
  margin-bottom: 10px;
  font-size: 0.8em;
}
</style>

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(Deriv)
library(Ryacas)
library(Rsolnp)
gradethis_setup(exercise.reveal_solution = FALSE, exercise.eval = TRUE)
knitr::opts_chunk$set(echo = FALSE, fig.height = 6, fig.width = 6)
```

## Offshoring Tire Production

Suppose that Flamerock Tires must decide where to produce one million tires: the US, where wages are 30 and the cost of capital is 5; or China, where wages are 5 and the cost of capital is 25. Production in each location follows the same technology (production function) given by: 

$$Q = f(L, K) = L^{0.25} * K^{0.75}$$

Computationally solve the cost-minimizing input levels in each location to produce the goal of 1 million tires; 

```{r costMin, exercise = TRUE}

```

```{r costMin-solution}
library(MicroWithR)
a = 1 / 4
q = 1000000
w_US = 30
r_US = 5
costMin_US <-
  solnp2(
    pars = c(1, 1),
    fun = function(x)
      w_US * x[1] + r_US * x[2],
    ineqfun = function(x)
      x[1] ^ a * x[2] ^ (1 - a),
    ineqLB = q,
    ineqUB = Inf,
    LB = c(0, 0)
  )
w_Ch = 5
r_Ch = 25
costMin_Ch <-
  solnp2(
    pars = c(1, 1),
    fun = function(x)
      w_Ch * x[1] + r_Ch * x[2],
    ineqfun = function(x)
      x[1] ^ a * x[2] ^ (1 - a),
    ineqLB = q,
    ineqUB = Inf,
    LB = c(0, 0)
  )
```

```{r costMin-check}
grade_this({if (!inherits(.result, c("list"))) {
    fail("Your class of your answer should be a list")}
  if (round(.result$pars, 1) != c(1466834.8, 880115.4)) { 
    fail("The values of L and K for production in China should equal (1466834.8, 880115.4).") }
  pass()
})
```

Which location minimizes costs of producing the target quantity? 
