---
title: "3.3 Monopoly"
output: 
  learnr::tutorial: 
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
author: 
  name: "Created by [Jim Bang](http://www.github.com/bangecon)"
  email: BangJamesT@sau.edu
  affiliation: "[St. Ambrose University](http://www.sau.edu)"
date: "`r format(Sys.Date(), '%d %b %Y')`"
---

<style type="text/css">
h1{font-size: 24pt}
h2{font-size: 20pt}
h3{font-size: 18pt}
h4,h5,h6{font-size: 16pt}
body{font-size: 16pt}
#doc-metadata {
  margin-bottom: 10px;
  font-size: 0.66em;
}
</style>

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
gradethis_setup(exercise.reveal_solution = TRUE, exercise.eval = TRUE)
knitr::opts_chunk$set(echo = FALSE, fig.height = 6, fig.width = 6)
```

## The Firm's Three Problems

1. Input cost minimization
2. Output profit maximization
3. Input profit maximization

### Monopolistic Output Markets

$$\matrix{ \matrix{\text{1. Single Seller} \\ \text{2. No Close Substitutes} } \Biggr\} \\ \text{3. Barred Entry} \\ \text{  }} \matrix{\text{ } \\ \implies \\ \text{ } \\ \implies \\ \text{ } }  \matrix{\Biggl\{ \matrix{\text{Firms are} \\ \text{price makers}} \\ \text{Profits/losses may} \\ \text{be permanent}}$$

### The Producer's Input-Hiring Problem

1. Objective: 
2. Endogenous (choice) variable(s): 
3. Exogenous variable(s): 

## Monopolist Demand, Total Revenue, and Marginal Revenue

The monopolist responds to the demand curve for the entire market, *not* a separate, perfectly (or more) elastic demand curve from the market-level equilibrium. 

THE MONOPOLIST ***IS*** THE MARKET!!!

### Monopolist Demand

Recall that the demand curve for the example we have followed was: 

$$Q_d(P) = D(P) = 6,000,000 - 2,000,000P$$  

for which the *inverse* demand curve:

$$P(Q) = D^{-1}(Q) = 3 - \frac{Q}{2,000,000}$$  

### Monopolist Total Revenue

This makes the *total* revenue curve: 

$$TR(Q) = P(Q)\cdot Q = 3Q - \frac{Q^2}{2,000,000},$$ 

Plot the total revenue function. 

```{r tr, exercise = TRUE}

```

```{r tr-hint}
# Check the plausible axis limits. 
# For linear demand curves MR = 0 midway between the x and y intercepts. 
```

```{r tr-solution}
library(ggplot2)
ggplot() + 
  geom_function(fun = function(x) x*(3 - x/2e6)) + 
  lims(x = c(0, 6000000), y = c(0, 4500000)) + 
  labs(x = "Quantity", y = "Total Revenue")
```

```{r tr-check}
grade_this({
  if (ith_geom_is(.result, "function", i = 1)) {
    pass(random_praise())
    }
  fail(random_encouragement())
  })
```

### Monopolist Marginal Revenue

For linear demand, the marginal revenue function has the same intercept on the price axis and is twice as steep as the inverse demand function. The *marginal* revenue function for our example is:

$$MR(Q) = 3 - \frac{Q}{1,000,000}$$

<details><summary>
Click here to see the derivation of the marginal revenue function from the (inverse) demand function.</summary>
Since $P = 3 - \frac{Q}{2,000}$, $TR = PQ = \Big(3 - \frac{Q}{2,000,000}\Big)Q = 3Q - \frac{Q^2}{2,000,000}$. Using the power rule for derivatives, the derivative of total revenue with respect to quantity is $\frac{dTR}{dQ} = 3 - \frac{2Q^1}{2,000,000} = 3 - \frac{Q}{1,000,000}$. `Ryacas` can do this for us: 
```{r mr, echo = TRUE, message = FALSE}
library(Ryacas)
tr.ysym <- ysym('3*x - x^2/2000000')
mr.ysym <- deriv(tr.ysym, 'x') 
mr.ysym
```

$$MR = \frac{dTR}{dQ} = `r tex(mr.ysym)`$$

For any linear demand function, the marginal revenue function will also be linear, will have the same vertical (price) intercept, and its slope will be twice as steep. If the demand function is not linear, we might need to figure out the marginal revenue function analytically (either by hand or using the `deriv()` function). 
</details>

Plot demand and marginal revenue. 

```{r demandGraph, exercise = TRUE}

```

```{r demandGraph-solution}
library(ggplot2)
ggplot() + 
  geom_function(fun = function(x) 3 - x/2000000, aes(color = "D")) + 
  geom_function(fun = function(x) 3 - x/1000000, aes(color = "MR")) + 
  lims(x = c(0, 6e+06), y = c(0, 3)) + 
  labs(x = "Quantity", y = "$", color = "") +
  scale_color_manual(breaks = c("D", "MR"), values = c("blue", "deepskyblue"))
```


```{r demandGraph-check}
grade_this({
  if (ith_geom_is(.result, "function", i = 1) & 
      ith_geom_is(.result, "function", i = 2)) {
    pass(random_praise())
    }
  fail(random_encouragement())
  })
```

## Costs

Suppose like we did in the last section that the production function for that firm took a Cobb-Douglas form with labor share equal to $1/3$ and total factor productivity parameter $50$. It also faced input prices of $w = 25$ and $r = 50$. In the competitive market, our firm planned its initial capital input (and planned labor input) based on an expected quantity of $100$.  

Here, the firm is the *entire* market, so let's suppose that this firm will plan its capital budget such that its long-run marginal cost equals marginal revenue. 

### Review: Cost Minimization and Short-Run Marginal Cost

$$\min_{L,K} \bigg\{C(L, K) = 10L + 20K \bigg\}$$

$$\text{s.t.: } f(L,K) = 50L^{1/3}K^{2/3} \ge 2500$$

The "Lagrangian" function for our example is: 

$$\mathcal{L} = wL + rK + \lambda (2500 - 50L^{1/3}K^{2/3})$$

### Monopolist's Capital Budgeting 

If one firm were to satisfy the entire market's demand, it would minimize costs for a quantity where the demand curve intersects the long-run marginal cost curve, which in our case (given constant returns to scale) is the same as the long-run average cost curve.  

Recall that in our example $LRAC = 0.3$, $LRAC = LRMC$ (because of CRTS), and the inverse demand is $P(Q) = 3 - Q/2,000,000$. 

Solve for (1) the expected quantity produced; and (2) the optimal capital input for this level of output. 

```{r capitalBudget, exercise = TRUE}

```

```{r capitalBudget-solution}
library(nleqslv)
demand <- function(x) 3 - x/2000000 - 0.3
q <- nleqslv(1, demand, control = list(trace = 0))$x
library(MicroWithR)
A = 100
a = 1/3
p1 = 10
p2 = 20
qx <- function(x) {
  A * x[1] ^ (a) * x[2] ^ (1 - a)
}
tc <- function(x) {
  p1*x[1] + p2*x[2]
}
costmin <- solnp2(
  pars = c(1,1),
  fun = tc,
  ineqfun = qx,
  ineqLB = q,
  ineqUB = Inf,
  LB = c(0, 0),
  UB = c(Inf, Inf), 
  control = list(trace = 0)
)
K <- costmin$pars[2]
K
```

```{r capitalBudget-check}
grade_this({if (round(.result, 0) != 54000) { 
    fail("The capital input is about 54,000") }
  pass()
})
```

This happens at $Q = 5,400,000$, which corresponds to labor and capital inputs of $L = 54,000$ and $K - 54,000$. 

### Short-Run Total Cost 

For $K^* = 54,000$, our firm's short run costs as a function of quantity of output are: 

<details><summary> Click here to see the derivation of the cost function from the production function and total costs. </summary>
$$f(K,L) = 100L^{1/3}K^{2/3} $$
$$L(Q) = \Bigg(\frac{Q}{54,000^{2/3} \cdot 100} \Bigg)^3 = \frac{Q^3}{2.916 \cdot 10^{15}}$$
$$C(Q) = 10L(Q) + 20(54,000) = \frac{Q^3}{2.916 \cdot 10^{14}} + 1,080,000$$
</details>

$$ C(Q) = \frac{Q^3}{2.916 \cdot 10^{14}} + 1,080,000$$

### Marginal Cost 

For this cost function, marginal cost will be: 

$$\frac{3Q^2}{2.916 \cdot 10^{14}}$$ 

### Average Cost

Average costs will be: 

$$\frac{Q^2}{2.916 \cdot 10^{14}} + \frac{1,080,000}{Q}$$

### Average Variable Cost

Average variable costs will be: 

$$\frac{Q^2}{2.916 \cdot 10^{14}}$$
Add marginal cost, average cost, and average variable cost for this monopolist to the demand/marginal revenue graph (`g`).

```{r costs, exercise = TRUE}

```

```{r costs-setup}
library(ggplot2)
g <- ggplot() + 
  geom_function(fun = function(x) 3 - x/2000000, aes(color = "D")) + 
  geom_function(fun = function(x) 3 - x/1000000, aes(color = "MR")) + 
  lims(x = c(0, 6e+06), y = c(0, 3)) + 
  labs(x = "Quantity", y = "$", color = "") +
  scale_color_manual(breaks = c("D", "MR"), values = c("blue", "deepskyblue"))
```


```{r costs-solution}
g <- g + 
  geom_function(fun = function(x) 3*x^2 / (2.916 * 10^(14)), aes(color = "MC")) + 
  geom_function(fun = function(x) x^2 / (2.916 * 10^(14)) + 1080000 / x, 
                aes(color = "ATC")) +
  geom_function(fun = function(x) x^2 / (2.916 * 10^(14)), aes(color = "AVC")) +
  scale_color_manual(
    breaks = c("D", "MR", "MC", "ATC", "AVC"), 
    values = c("blue", "deepskyblue", 'red', 'firebrick', 'hotpink2'))
```

## Solving for Equilibrium

Like any firm, the monopolist maximizes profits and sets its output according to $MR(Q^*) = MC(Q^*)$ and sets its price according to the demand curve at that quantity, $D^{-1}(Q^*)$.

### Computational Approach

Solve for the profit-maximizing quantity and find the monopolist's price at that quantity.

```{r sr, exercise = TRUE}

```

```{r sr-setup}
library(nleqslv)
demand <- function(x) 3 - x/2000000 - 0.3
q <- nleqslv(1, demand, control = list(trace = 0))$x
library(MicroWithR)
A = 100
a = 1/3
w = 10
r = 20
qx <- function(x) {
  A * x[1] ^ (a) * x[2] ^ (1 - a)
}
tc <- function(x) {
  w*x[1] + r*x[2]
}
costmin <- solnp2(
  pars = c(1,1),
  fun = tc,
  ineqfun = qx,
  ineqLB = q,
  ineqUB = Inf,
  LB = c(0, 0),
  UB = c(Inf, Inf), 
  control = list(trace = 0)
)
K <- costmin$pars[2]
```

```{r sr-solution}
d.Inv <- function(x) {
  3 - qx(x)/2000000
}
profit <- function(x) {
    qx(x) * d.Inv(x) - tc(x)
}
eq <- solnp2(
  par = c(1, 1), 
  fun = profit, 
  eqfun = function(x) x[2],
  eqB = K,
  LB = c(0,0), 
  control = list(trace = 0), 
  max = TRUE
  )
eq$pars
qx(eq$pars)
d.Inv(eq$pars)
profit(eq$pars)
```

```{r sr-check, echo = FALSE}
grade_this({if (round(.result, 0) != 3331447) { 
    fail("The maximum level of profit should be about 3,331,447") }
  pass()
})
```

## Efficiency

* **Productive Efficiency:** We can immediately see that the monopolist does not reach the minimum of ATC: minimum ATC occurs where MC = ATC. Since MC < ATC at the monopoly equilibrium, the monopolist utilizes *excess capacity*.  
We can further see this from the fact that the *optimal* capital-labor ratio for our example is $1:1$, but in the monopoly equilibrium this ratio is about $6.4:1$. We can also see that the monopolist produces at an average cost of about 0.4 while the competitive market produces at an average cost of 0.3.
* **Allocative Efficiency:** Allocative efficiency we can see from the fact that $MC \approx 0.09 < P \approx 1.54$, which means that since $MU = P$ from the consumer problem (since we implicitly assume consumers are price-takers), $MC \ne MU$. Even if the market otherwise satisfies the usual efficiency conditions, the fact that a monopolist exists spoils allocative efficiency. 

Let's calculate the consumer and producer surplus under each market structure to see just how much total surplus the monopolist destroys by restricting output. 

### Graphical Method (Competitive Equilibrium)

Graphically, the total surplus in the competitive equilibrium is shown below as the sum of the consumer surplus in light blue (sum of marginal utilities enjoyed above the price paid) and the producer surplus in pink (sum of revenues received above the marginal costs incurred). 

```{r echo = FALSE}
library(ggplot2)
A = 100
a = 1/3
w = 10
r = 20
qx <- function(x) {
  A * x[1] ^ (a) * x[2] ^ (1 - a)
}
d.Inv <- function(x) {
  3 - qx(x)/2000000
}
g <- ggplot() + 
  geom_function(fun = function(x) 3 - x/2000000, aes(color = "D")) + 
  geom_function(fun = function(x) 3 - x/1000000, aes(color = "MR")) + 
  lims(x = c(0, 6e+06), y = c(0, 3)) + 
  labs(x = "Quantity", y = "$", color = "") +
  scale_color_manual(breaks = c("D", "MR"), values = c("blue", "deepskyblue")) + 
  geom_function(fun = function(x) 3*x^2 / (2.916 * 10^(14)), aes(color = "MC")) + 
  geom_function(fun = function(x) x^2 / (2.916 * 10^(14)) + 1080000 / x, 
                aes(color = "ATC")) +
  geom_function(fun = function(x) x^2 / (2.916 * 10^(14)), aes(color = "AVC")) +
  scale_color_manual(
    breaks = c("D", "MR", "MC", "ATC", "AVC"), 
    values = c("blue", "deepskyblue", 'red', 'firebrick', 'hotpink2'))
```

```{r, echo = TRUE}
schedule = data.frame(x = seq(0, qx(c(54000, 54000)), 100), p = NA, d = NA, mc = NA)
schedule$p = d.Inv(c(54000, 54000))
schedule$d = sapply(schedule$x, FUN = function(x) 3 - x/2000000) 
schedule$mc = sapply(schedule$x, FUN = function(x) 3*x^2 / (2.916 * 10^(14)))
g + 
  geom_ribbon(aes(x = x, ymin = p, ymax = d, fill = "CS"), data = schedule, alpha = 0.1)  + 
  geom_ribbon(aes(x = x, ymin = mc, ymax = p, fill = "PS"), data = schedule, alpha = 0.1) + 
  labs(x = "Quantity", y = "$", color = "", fill = "") + 
  scale_fill_manual(breaks = c("CS", "PS"), values = c("blue", "red"))
```

### Computational Method (Competitive Equilibrium)

We calculate the surplus by finding the area of these shaded regions. We could do the total surplus all at once, but doing it separately helps see how the monopolist *transfers* surplus from the consumer to itself as well as destroying some of the total surplus. 

For linear functions areas below and between curves is easy: we only need to calculate areas of triangles and/or rectangles; for curves, we calculate the areas below by taking the *integral* (gasp!) of the diferrence between the curves. 

Calculate the consumer surplus using the fact that we know that the demand function is linear with intercept of 3, the equilibrium quantity is 5,400,000, and the equilibrium price is 0.3.

```{r cs, exercise = TRUE}

```

```{r cs-hint}

```

```{r cs-setup}
library(nleqslv)
demand <- function(x) 3 - x/2000000 - 0.3
q <- nleqslv(1, demand, control = list(trace = 0))$x
library(MicroWithR)
A = 100
a = 1/3
w = 10
r = 20
qx <- function(x) {
  A * x[1] ^ (a) * x[2] ^ (1 - a)
}
tc <- function(x) {
  w*x[1] + r*x[2]
}
costmin <- solnp2(
  pars = c(1,1),
  fun = tc,
  ineqfun = qx,
  ineqLB = q,
  ineqUB = Inf,
  LB = c(0, 0),
  UB = c(Inf, Inf), 
  control = list(trace = 0)
)
K <- costmin$pars[2]
d.Inv <- function(x) {
  3 - qx(x)/2000000
}
profit <- function(x) {
    qx(x) * d.Inv(x) - tc(x)
}
eq <- solnp2(
  par = c(1, 1), 
  fun = profit, 
  eqfun = function(x) x[2],
  eqB = 54000,
  LB = c(0,0), 
  control = list(trace = 0), 
  max = TRUE
  )
```

```{r cs-solution}
cs <- 0.5*(3 - d.Inv(c(54000,54000)))*qx(c(54000,54000))
cs
```

```{r cs-check}
grade_this({if (round(.result, 0) != 7290000) { 
    fail("The consumer surplus is about 7,290,000") }
  pass()
})
```

Next, calculate the producer surplus using the `integrate()` function over the function $0.3 - MC(Q)$. (Unfortunately there's no simple formula shortcut since the marginal cost curve is a curve.) 

```{r ps, exercise = TRUE}

```


```{r ps-setup}
library(nleqslv)
demand <- function(x) 3 - x/2000000 - 0.3
q <- nleqslv(1, demand, control = list(trace = 0))$x
library(MicroWithR)
A = 100
a = 1/3
w = 10
r = 20
qx <- function(x) {
  A * x[1] ^ (a) * x[2] ^ (1 - a)
}
tc <- function(x) {
  w*x[1] + r*x[2]
}
costmin <- solnp2(
  pars = c(1,1),
  fun = tc,
  ineqfun = qx,
  ineqLB = q,
  ineqUB = Inf,
  LB = c(0, 0),
  UB = c(Inf, Inf), 
  control = list(trace = 0)
)
K <- costmin$pars[2]
d.Inv <- function(x) {
  3 - qx(x)/2000000
}
profit <- function(x) {
    qx(x) * d.Inv(x) - tc(x)
}
eq <- solnp2(
  par = c(1, 1), 
  fun = profit, 
  eqfun = function(x) x[2],
  eqB = 54000,
  LB = c(0,0), 
  control = list(trace = 0), 
  max = TRUE
  )
```

```{r ps-hint}
```


```{r ps-solution}
ps <-
  integrate(
    function(x)
      0.3 - (3 * x ^ 2) / 2916e11,
    lower = 0,
    upper = 5.4e+06
  )[1] |>
  unlist()
ps
```

```{r ps-check}
grade_this({if (round(.result, 0) != 1080000) { 
    fail("The producer surplus is about 1,080,000") }
  pass()
})
```

### Graphical Method (Monopoly Equilibrium)

Graphically, we can visualize the two sets of surpluses in a single plot space by shading the different surpluses on top of one another using slightly different colors.

```{r echo = FALSE}
library(ggplot2)
schedule = data.frame(x = seq(0, qx(c(54000, 54000)), 100), p = NA, d = NA, mc = NA)
schedule$p = d.Inv(costmin$pars)
schedule$d = sapply(schedule$x, FUN = function(x) 3 - x/2000000) 
schedule$mc = sapply(schedule$x, FUN = function(x) 3*x^2 / (2.916 * 10^(14)))
# schedulem = data.frame(x = seq(0, qx(eq$pars), 100), p = NA, d = NA, mc = NA)
# schedulem$pm = d.Inv(eq$pars)
# schedulem$d = sapply(schedulem$x, FUN = function(x) 3 - x/2000000) 
# schedulem$mc = sapply(schedulem$x, FUN = function(x) 3*x^2 / (2.916 * 10^(14)))
g <- ggplot() + 
  geom_function(fun = function(x) 3 - x/2000000, aes(color = "D")) + 
  geom_function(fun = function(x) 3 - x/1000000, aes(color = "MR")) + 
  geom_function(fun = function(x) 3*x^2 / (2.916 * 10^(14)), aes(color = "MC")) + 
  geom_function(fun = function(x) x^2 / (2.916 * 10^(14)) + 1080000 / x, 
                aes(color = "ATC")) +
  geom_function(fun = function(x) x^2 / (2.916 * 10^(14)), aes(color = "AVC")) +
  geom_ribbon(aes(x = x, ymin = p, ymax = d, fill = "CS"), data = schedule, alpha = 0.1) + 
  geom_ribbon(aes(x = x, ymin = mc, ymax = p, fill = "PS"), data = schedule, alpha = 0.1) + 
  lims(x = c(0, 6e+06), y = c(0, 3)) + 
  labs(x = "Quantity", y = "$", color = "", fill = "") +
  scale_color_manual(
    breaks = c("D", "MR", "MC", "ATC", "AVC"), 
    values = c("blue", "deepskyblue", 'red', 'firebrick', 'hotpink2')) + 
  scale_fill_manual(
    breaks = c("CS", "PS"), 
    values = c("blue", "red"))
```

```{r, echo = TRUE, fig.width=9}
schedule$pm = d.Inv(eq$pars)
mc <- function(x) {
  3*x^2/(2.916e14)
}
g + 
  geom_ribbon(
    aes(x = x, ymin = pm, ymax = d, fill = "CS'"),
    data = subset(schedule, x <= qx(eq$pars)), alpha = 0.1) + 
  geom_ribbon(
    aes(x = x, ymin = mc, ymax = pm, fill = "PS'"), 
    data = subset(schedule, x <= qx(eq$pars)), alpha = 0.1) +
  geom_text(aes(
    x = qx(eq$pars)/2, 
    y = (d.Inv(eq$pars) + 2*d.Inv(costmin$pars))/3, label = "A"), 
    size = 5) + 
  geom_text(aes(
    x = (2*qx(eq$pars) + qx(costmin$pars))/3, 
    y = (d.Inv(eq$pars) + 2*d.Inv(costmin$pars))/3, label = "B"), 
    size = 5) + 
  geom_text(aes(
    x = (2*qx(eq$pars) + qx(costmin$pars))/3, 
    y = (mc(qx(eq$pars)) + 2*d.Inv(costmin$pars))/3, label = "C"), 
    size = 5) + 
  scale_fill_manual(
    breaks = c("CS", "PS", "CS'", "PS'"), 
    values = c("blue", "red", "deepskyblue", "firebrick"))
```

We can see from this figure that some of the surpluses generated by a competitive market get left out by the monopolist since the monopolist chooses to produce a quantity below the point where marginal costs intersect with demand and $P = MU = MC$. The name for these unrealized gains are called "deadweight losses." Next, we will calculate these for our example. 

### Computational Method (Monopoly Equilibrium)

Calculate the consumer surplus under this monopolist.  

```{r cs_, exercise = TRUE}

```

```{r cs_m-setup}
library(nleqslv)
demand <- function(x) 3 - x/2000000 - 0.3
q <- nleqslv(1, demand, control = list(trace = 0))$x
library(MicroWithR)
A = 100
a = 1/3
w = 10
r = 20
qx <- function(x) {
  A * x[1] ^ (a) * x[2] ^ (1 - a)
}
tc <- function(x) {
  w*x[1] + r*x[2]
}
costmin <- solnp2(
  pars = c(1,1),
  fun = tc,
  ineqfun = qx,
  ineqLB = q,
  ineqUB = Inf,
  LB = c(0, 0),
  UB = c(Inf, Inf), 
  control = list(trace = 0)
)
K <- costmin$pars[2]
d.Inv <- function(x) {
  3 - qx(x)/2000000
}
profit <- function(x) {
    qx(x) * d.Inv(x) - tc(x)
}
eq <- solnp2(
  par = c(1, 1), 
  fun = profit, 
  eqfun = function(x) x[2],
  eqB = 54000,
  LB = c(0,0), 
  control = list(trace = 0), 
  max = TRUE
  )
```

```{r cs_m-hint}
```


```{r cs_m-solution}
cs.m <- 0.5 * (3 - d.Inv(eq$pars)) * qx(eq$pars)
cs.m
```

```{r cs_m-check}
grade_this({if (round(.result, 0) != 2120982) { 
    fail("The producer surplus is about 2,120,982") }
  pass()
})
```

Calculate the producer surplus under this monopolist.  

```{r ps_m, exercise = TRUE}

```

```{r ps_m-setup}
library(nleqslv)
demand <- function(x) 3 - x/2000000 - 0.3
q <- nleqslv(1, demand, control = list(trace = 0))$x
library(MicroWithR)
A = 100
a = 1/3
w = 10
r = 20
qx <- function(x) {
  A * x[1] ^ (a) * x[2] ^ (1 - a)
}
tc <- function(x) {
  w*x[1] + r*x[2]
}
costmin <- solnp2(
  pars = c(1,1),
  fun = tc,
  ineqfun = qx,
  ineqLB = q,
  ineqUB = Inf,
  LB = c(0, 0),
  UB = c(Inf, Inf), 
  control = list(trace = 0)
)
K <- costmin$pars[2]
d.Inv <- function(x) {
  3 - qx(x)/2000000
}
profit <- function(x) {
    qx(x) * d.Inv(x) - tc(x)
}
eq <- solnp2(
  par = c(1, 1), 
  fun = profit, 
  eqfun = function(x) x[2],
  eqB = 54000,
  LB = c(0,0), 
  control = list(trace = 0), 
  max = TRUE
  )
```

```{r ps_m-hint}
```

```{r ps_m-solution}
ps.m <-
  integrate(
    function(x)
      d.Inv(eq$pars) - (3 * x ^ 2) / 2916e11,
    lower = 0,
    upper = qx(eq$pars)
  )[1] |>
  unlist()
ps.m
```

```{r ps_m-check}
grade_this({if (round(.result, 0) != 4411447 ) { 
    fail("The producer surplus is about 4,411,447") }
  pass()
})
```

The following table presents a comparison of the surplusses under a competitive market and a monopolist, respectively. 

```{r}
library(nleqslv)
demand <- function(x) 3 - x/2000000 - 0.3
q <- nleqslv(1, demand, control = list(trace = 0))$x
library(MicroWithR)
A = 100
a = 1/3
w = 10
r = 20
qx <- function(x) {
  A * x[1] ^ (a) * x[2] ^ (1 - a)
}
tc <- function(x) {
  w*x[1] + r*x[2]
}
costmin <- solnp2(
  pars = c(1,1),
  fun = tc,
  ineqfun = qx,
  ineqLB = q,
  ineqUB = Inf,
  LB = c(0, 0),
  UB = c(Inf, Inf), 
  control = list(trace = 0)
)
K <- costmin$pars[2]
d.Inv <- function(x) {
  3 - qx(x)/2000000
}
profit <- function(x) {
    qx(x) * d.Inv(x) - tc(x)
}
eq <- solnp2(
  par = c(1, 1), 
  fun = profit, 
  eqfun = function(x) x[2],
  eqB = 54000,
  LB = c(0,0), 
  control = list(trace = 0), 
  max = TRUE
  )
cs <- 0.5*(3 - d.Inv(c(54000,54000)))*qx(c(54000,54000))
ps <-
  integrate(
    function(x)
      0.3 - (3 * x ^ 2) / 2916e11,
    lower = 0,
    upper = 5.4e+06
  )[1] |>
  unlist()
cs.m <- 0.5 * (3 - d.Inv(eq$pars)) * qx(eq$pars)
ps.m <-
  integrate(
    function(x)
      d.Inv(eq$pars) - (3 * x ^ 2) / 2916e11,
    lower = 0,
    upper = qx(eq$pars)
  )[1] |>
  unlist()

surplusTable <- data.frame(
  Competition = c(cs, ps, cs + ps), 
  Monopolist = c(cs.m, ps.m, cs.m + ps.m),
  Change = c(cs.m - cs, ps.m - ps, (cs.m + ps.m) - (cs + ps)), 
  row.names = c("Consumer Surplus", "Producer Surplus", "Total Surplus")
)
surplusTable
```

Summary:  

1. Consumers lose the blue portion of the deadweight loss (the upper triangle, labeled B) *and* transfer some of their original surplus (the double-shaded rectangle, labeled A) to the producer. Together, these losses equal $-(A+B)$.  
2. Producers lose the red portion of the deadweight loss (the lower curved triangle, labeled C) *but* gain back even more by capturing some of the consumers' original surplus (the double-shaded portion, labeled A) through higher prices on the lower quantity produced. Together the net gain for the producer equals $A - C$.   
3. Adding up we get the total net change in surplus, which is the sum of the changes to consumer and producer surplusses, $-(A+B) + (A-C) = -(B+C)$.
