---
output: 
  learnr::tutorial: 
    progressive: true
runtime: shiny_prerendered
title: "1.1 Budget Constraint"
author: 
  name: "Created by [Jim Bang](http://www.github.com/bangecon)"
  email: BangJamesT@sau.edu
  affiliation: "[St. Ambrose University](http://www.sau.edu)"
date: "`r format(Sys.Date(), '%d %b %Y')`"
---

<style type="text/css">
#doc-metadata {
  margin-bottom: 10px;
  font-size: 0.8m;
}
</style>

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(ggcheck)
gradethis_setup()
tutorial_options(exercise.reveal_solution = FALSE, exercise.eval = FALSE, exercise.checker = gradethis_exercise_checker)
knitr::opts_chunk$set(echo = FALSE, fig.align = "center", fig.height = 6.5, fig.width = 8.5)
```


### Food Stamps

Consider a consumer who receives "food stamps" that subsidizes up to five donuts (but does not subsidize beer). Graph the budget line for this consumer using the original parameters ($m = 20$, $p_d = 1$, and $p_b = 2$). (Hint: You need to plot two separate lines because there are two segments to this budget line. The first is as if income has increased by the dollar value of 5 donuts; the second part is horizontal noting that you cannot spend the subsidy on beer and so the maximum amount of beer the consumer can buy does not change. Use the `geom_segment(aes(x, xend, y, yend))` function to draw this part.)

```{r foodStamps, exercise = TRUE}

```

```{r foodStamps-solution, warning=FALSE, message = FALSE}
library(ggplot2)
m = 20
p_d = 1
p_b = 2
s_d = 5
ggplot() + 
  geom_function(fun = function(x) s_d + m / p_d - p_b * x / p_d, 
    xlim = c(0, m/p_b)) +
  geom_segment(aes(x = m/p_b, xend = m/p_b, y = 0, yend = s_d)) +
  lims(x = c(0, m/p_b), y = c(0, s_d + m/p_d)) + 
  labs(x = "Beer", y = "Donuts")
```

```{r foodStamps-check}
grade_this({
  if ((ith_geom_is(.result, "function", 1) & ith_geom_is(.result, "segment", 2)) | 
      ith_geom_is(.result, "segment", 1) & ith_geom_is(.result, "function", 2)) {
    pass(random_praise())
  }
  fail(random_encouragement())
})
```
