---
title: "1.5 Comparative Statics I: Income Changes"
editor_options: 
  markdown: 
    wrap: 72
output: 
  learnr::tutorial: 
    progressive: true
runtime: shiny_prerendered
author: 
  name: "Created by [Jim Bang](http://www.github.com/bangecon)"
  email: BangJamesT@sau.edu
  affiliation: "[St. Ambrose University](http://www.sau.edu)"
date: "`r format(Sys.Date(), '%d %b %Y')`"
---

<style type="text/css">
#doc-metadata {
  margin-bottom: 10px;
  font-size: 0.8em;
}
</style>

```{r setup, include = FALSE}
library(learnr)
library(gradethis)
library(ggcheck)
gradethis_setup()
tutorial_options(exercise.eval = TRUE, exercise.reveal_solution = FALSE)
knitr::opts_chunk$set(echo = FALSE, fig.align = "center", fig.height = 6, fig.width = 6)
knitr::include_graphics
```

## In-Kind versus Cash Transfers  

Economists claim that cash transfers are more "efficient" than in-kind transfers because there is no dead-weigth utility loss from the fact that the recipient might have spent the money differently.^[I recommend testing this hypothesis with someone in a gift-giving situation like with your significant other on Valentines Day. Don't forget to report your results to the class!]

### Initial Solution

Consider a consumer with income of 100 ($m = 100$), which they can allocate by consuming necessities (e.g. food, housing) or invseting in capital (e.g. home production, education) at prices price ($p_n = 1$ and $p_k = 2$). Also suppose that the consumer has preferences dictated by a constant elasticity of substitution utility function given by: 

$$u(q_n, q_k) = (0.1q_n^{1/2} + 0.9q_k^{1/2})^2$$

This makes the consumer's maximization problem: 

$$max \big\{(0.1q_n^{1/2} + 0.9q_k^{1/2})^2\big\}$$
subject to: 

$$1q_n + 2q_k \le 100$$

Find the initial solution to the consumer's utility-maximization problem. Print the utility maximizing consumption choice (`pars`) for the solution.  

```{r initialEquilibrium, exercise = TRUE}

```

```{r initialEquilibrium-solution}
library(MicroWithR)
library(Rsolnp)
m = 100
pn = 1
pk = 2
r = 0.5
a = 0.1
control = list(trace = 0)
ux <- function(x)
  (a * x[1] ^ (r) + (1 - a) * x[2] ^ (r)) ^ (1 / r)
bc <- function(x)
  pn * x[1] + pk * x[2]
eq <- solnp2(
  c(1, 1),
  ux,
  ineqfun = bc,
  ineqLB = 0,
  ineqUB = m,
  LB = c(0, 0),
  UB = c(Inf, Inf),
  max = TRUE
)
```

```{r initialEquilibrium-check}
grade_this({if (!inherits(.result, c("list"))) {
    fail("Your class of your answer should be a list")}
  if (round(.result$pars, 2) != c(11.11, 44.44)) { 
    fail("The optimal amount of goods one and two after the change should be 11.11 and 44.44") }
  pass()
})
```

### In-Kind Transfers

Suppose that the government offers this household an in-kind transfer of necessities, such as food stamps or housing assistance equal to \$15 but that the consumer can only redeem for necessities ($q_n$). Find the new quilibrium. Print the utility maximizing consumption choice (`pars`) for the solution.  

```{r inkindEquilibrium, exercise = TRUE}

```

```{r inkindEquilibrium-hint}
# The solution is nearly the same as the previous example, except you add 20 to the budget AND cap expenditures on capital investments at m/pk by setting an upper bound (UB) option. 
```

```{r inkindEquilibrium-setup}
library(MicroWithR)
library(Rsolnp)
m = 100
pn = 1
pk = 2
r = 0.5
a = 0.1
control = list(trace = 0)
ux <- function(x)
  (a * x[1] ^ (r) + (1 - a) * x[2] ^ (r)) ^ (1 / r)
bc <- function(x)
  pn * x[1] + pk * x[2]
eq <- solnp2(
  c(1, 1),
  ux,
  ineqfun = bc,
  ineqLB = 0,
  ineqUB = m,
  LB = c(0, 0),
  UB = c(Inf, Inf),
  max = TRUE
)
```

```{r inkindEquilibrium-solution}
sn = 10
inkindEq <- solnp2(
  c(1, 1),
  ux,
  ineqfun = bc,
  ineqLB = 0,
  ineqUB = m + sn,
  LB = c(0, 0),
  UB = c(Inf, m/pk),
  max = TRUE
)
inkindEq$pars
```

```{r inkindEquilibrium-check}
grade_this({if (!inherits(.result, c("list"))) {
    fail("Your class of your answer should be a list")}
  if (.result$pars != c(20, 50)) { 
    fail("The optimal amount of goods one and two after the change should be 15 and 50") }
  pass()
})
```

### Cash Transfers

Suppose that the government offers this household an in-kind transfer of necessities, such as food stamps or housing assistance equal to \$15 but that the consumer can only redeem for necessities ($q_n$). Find the new quilibrium. Print the utility maximizing consumption choice (`pars`) for the solution.  

```{r cashEquilibrium, exercise = TRUE}

```

```{r cashEquilibrium-hint}
# The solution is nearly the same as the previous example, except you add 20 to the budget WITHOUT capping expenditures on capital investments. 
```

```{r cashEquilibrium-setup}
library(MicroWithR)
library(Rsolnp)
m = 100
pn = 1
pk = 2
r = 0.5
a = 0.1
control = list(trace = 0)
ux <- function(x)
  (a * x[1] ^ (r) + (1 - a) * x[2] ^ (r)) ^ (1 / r)
bc <- function(x)
  pn * x[1] + pk * x[2]
eq <- solnp2(
  c(1, 1),
  ux,
  ineqfun = bc,
  ineqLB = 0,
  ineqUB = m,
  LB = c(0, 0),
  UB = c(Inf, Inf),
  max = TRUE
)
sn = 10
inkindEq <- solnp2(
  c(1, 1),
  ux,
  ineqfun = bc,
  ineqLB = 0,
  ineqUB = m + sn,
  LB = c(0, 0),
  UB = c(Inf, m/pk),
  max = TRUE
)
```

```{r cashEquilibrium-solution}
cashEq <- solnp2(
  c(1, 1),
  ux,
  ineqfun = bc,
  ineqLB = 0,
  ineqUB = m + sn,
  LB = c(0, 0),
  UB = c(Inf, Inf),
  max = TRUE
)
cashEq$pars
```

```{r cashEquilibrium-check}
grade_this({if (!inherits(.result, c("list"))) {
    fail("Your class of your answer should be a list")}
  if (round(.result$pars, 2) != c(13.33, 53.33)) { 
    fail("The optimal amount of goods one and two after the change should be 12.78 and 51.11") }
  pass()
})
```

### Graph

Graph all three equilibria (indifference curves and budget lines) on a single `ggplot()`.   

```{r inKindGraph, exercise = TRUE}
```

```{r inKindGraph-setup}
```

```{r inKindGraph-solution}
ggplot() +
  geom_function(
    fun = function(x)
      m / pk - pn * x / pk
  ) +
  geom_segment(aes(
    x = 0,
    xend = sn,
    y = m / pk,
    yend = m / pk
    ),
    linetype = 'dashed',
    color = 'blue') +
  geom_segment(aes(
    x = sn,
    xend = (m + sn) / pn,
    y = m / pk,
    yend = 0
    ),
    linetype = 'dashed',
    color = 'blue') +
  geom_function(
    fun = function(x)
      (m + sn) / pk - pn * x / pk,
    linetype = 'dashed',
    color = 'green'
  ) + 
  geom_function(
    fun = function(x) 
      ((ux(eq$pars)^r - a*x^r)/(1-a))^(1/r)
  ) + 
  geom_function(
    fun = function(x) 
      ((ux(inkindEq$pars)^r - a*x^r)/(1-a))^(1/r), 
    linetype = 'dashed', color = 'blue'
  ) + 
  geom_function(
    fun = function(x) 
      ((ux(cashEq$pars)^r - a*x^r)/(1-a))^(1/r), 
    linetype = 'dashed', color = 'green'
  ) + 
  lims(x = c(0, 1.2 * m / pn), y = c(0, 1.2 * m / pk)) + 
  labs(x = "Consumption of Necessities", y = "Investment in Human/Physical Capital")
```

```{r inKindGraph-check}
grade_code() 
```
