---
title: "2.3 Cost Curves"
output: 
  learnr::tutorial: 
    progressive: true
runtime: shiny_prerendered
author: 
  name: "Created by [Jim Bang](http://www.github.com/bangecon)"
  email: BangJamesT@sau.edu
  affiliation: "[St. Ambrose University](http://www.sau.edu)"
date: "`r format(Sys.Date(), '%d %b %Y')`"
---

<style type="text/css">
h1{font-size: 24pt}
h2{font-size: 20pt}
h3{font-size: 18pt}
h4,h5,h6{font-size: 16pt}
body{font-size: 16pt}
#doc-metadata {
  margin-bottom: 10px;
  font-size: 0.66em;
}
</style>

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(ggcheck)
gradethis_setup(exercise.reveal_solution = TRUE)
knitr::opts_chunk$set(echo = FALSE, fig.height = 6, fig.width = 6)
```

## Review: The Firm's Three Problems:

1. <span style="color: red;">Cost-Minimizing Combination of Inputs: For a *given* level of (expected) output (firm demand) choose the *combination* of inputs that minimizes the cost of producing that amount to derive the cost function.</span>
2. Profit-Maximizing Level of Output: Choose the *level* of output that maximizes profits by adjusting quantity and/or price of the output. 
3. Long-Run Readjustment: Adjust the level of capital to minimize costs at the given price and quantity in the market. This might happen at the firm level (adjusting firm capacity) or at the industry level (entry/exit of firms).

### Long-Run Cost Curves

A rational firm starts out by taking a long-run view of their market. To do this, it will determine the different levels and combinations of inputs that minimize its long run costs. 

Then, it must budget its fixed inputs (in our example, capital) based on (1) the quantity it expects to sell; and (2) the least-costly way to produce that quantity with its production methods.  

### Initial Solution

For expected sales of $y = 2500$ units of output and inpput prices $w = 10$ and $w = 20$, the firm in our previous example minimized costs at 25 units of labor and 25 units of capital at a total cost of 75. 

```{r initialSolution, echo=TRUE}
library(MicroWithR)     # Don't forget to load libraries!
A = 100
a = 1 / 3
w = 10
r = 20
q = 2500
costMin <- solnp(
  pars = c(1, 1),
  fun = function(x)
    w * x[1] + r * x[2],
  ineqfun = function(x)
    A * x[1] ^ a * x[2] ^ (1 - a),
  ineqLB = q,
  ineqUB = Inf,
  LB = c(0, 0),
  control = list(trace = 0)
)
```

To find costs over several quantities (i.e. a cost *schedule*), all we need to do is solve the cost-minimization problem for many different levels of output. Computationally, we can do this using a *for-loop*. 

Run the following code to iterate the cost-minimization problem over any values of output to find the optimal levels of labor, capital, and cost for levels of output from 0 to 5,000 in increments of 100. 

<details style="line-height:105%"><summary>Click here to view the steps for writing the loop.</summary>
1. Initialize a new `data.frame` object named `costSchedule` with variables `Quantity`, `Labor`, `Capital`, `Total_Cost`, and `Average_Cost`. <br> 
Set the values of `Quantity` equal to the sequence of outputs from 0 to 5,000 by increments of 100, and the other variables equal to `NA`.
2. Begin a `for(...) {...}` loop starting at `i = 1` through the number of rows (`nrow()`) in costSchedule that does the following for each `i`:
a. Compute the solution to the cost-minimization problem for the quantity of output in row `i` of `costSchedule$Quantity` and (temporarily) stores the result as `costMinTemp`. 
b. Assigns the $i^th$ row of the `Labor` variable in `costSchedule` equal to the first value in the `solution` variable in the `costMinTemp` object, and assigns the $i^th$ row of `Capital` equal to the second value of `solution`.
c. Assigns the last value (`tail(x, 1)`) of the `values` in `costMinTemp` to the `Total_Cost` variable, and calculates the average cost as `Total_Cost` divided by `Quantity`.
</details>

```{r costSchedule, echo = TRUE}
library(MicroWithR)     # Don't forget to load libraries!
A = 100
a = 1 / 3
w = 10
r = 20
q = 2500
crtsSchedule <-
  data.frame(
    Quantity = seq(from = 0, to = 5000, by = 100),          # Initialize 51 Q's 
    Labor = NA,
    Capital = NA,
    Total_Cost = NA,
    Average_Cost = NA
  )
tc <- function(x) {
  w * x[1] + r * x[2]                                       # Define objective 
}
crts <- function(x) {
  A * x[1] ^ a * x[2] ^ (1 - a)                   # Define constraint
}
for (i in 1:nrow(crtsSchedule)) {                           # Initialize for-loop
  costMinTemp <- solnp(
    pars = c(1, 1),                                         # Set initial pars
    fun = tc,
    ineqfun = crts,
    ineqLB = crtsSchedule$Quantity[i],                      # Set ineqLB to Q[i]
    ineqUB = Inf,
    LB = c(0, 0),                                           # Nonnegative constraint
    UB = c(Inf, Inf), 
    control = list(trace = 0)
  )
  crtsSchedule$Labor[i] <- costMinTemp$pars[1]              # Optimal labor input
  crtsSchedule$Capital[i] <- costMinTemp$pars[2]            # Optimal capital input
  crtsSchedule$Total_Cost[i] <- tail(costMinTemp$values, 1) # Calculate total cost
}
crtsSchedule$Average_Cost <-
  crtsSchedule$Total_Cost / crtsSchedule$Quantity           # Calculate ATC
head(crtsSchedule)                                          # Print the 1st 6 rows
```

## Graphical Analysis

### Constant Returns to Scale

Plot total costs as functions of quantity using the data frame we generated the cost-minimization solutions in the example (`crtsSchedule`). Name the plot `tcGraph`.

```{r tcGraph, exercise = TRUE, exercise.lines = 8}

```

```{r tcGraph-setup}
library(Rsolnp)     # Don't forget to load libraries!
A = 100
a = 1 / 3
w = 10
r = 20
crtsSchedule <-
  data.frame(
    Quantity = seq(from = 0, to = 5000, by = 100),          # Initialize 51 Q's 
    Labor = NA,
    Capital = NA,
    Total_Cost = NA,
    Average_Cost = NA
  )
tc <- function(x) {
  w * x[1] + r * x[2]                                       # Define objective 
}
crts <- function(x) {
  A * x[1] ^ a * x[2] ^ (1 - a)                             # Define constraint
}
for (i in 1:nrow(crtsSchedule)) {                           # Initialize for-loop
  costMinTemp <- solnp(
    pars = c(1, 1),                                         # Set initial pars
    fun = tc,
    ineqfun = crts,
    ineqLB = crtsSchedule$Quantity[i],                      # Set ineqLB to Q[i]
    ineqUB = Inf,
    LB = c(0, 0),                                           # Nonnegative constraint
    UB = c(Inf, Inf), 
    control = list(trace = 0)
  )
  crtsSchedule$Labor[i] <- costMinTemp$pars[1]              # Optimal labor input
  crtsSchedule$Capital[i] <- costMinTemp$pars[2]            # Optimal capital input
  crtsSchedule$Total_Cost[i] <- tail(costMinTemp$values, 1) # Calculate total cost
}
crtsSchedule$Average_Cost <-
  crtsSchedule$Total_Cost / crtsSchedule$Quantity           # Calculate ATC
```

```{r tcGraph-solution}
library(ggplot2)
tcGraph <- ggplot() + 
  geom_line(aes(x = Quantity, y = Total_Cost), data = crtsSchedule) +
  labs(x = "Quantity", y = "Total Cost")
tcGraph
```

```{r tcGraph-check}
grade_code()
```

Plot average costs as functions of quantity. Store the plot as `atcGraph`. Set your y-axis limits to [0,1]. 

```{r atcGraph, exercise = TRUE}

```

```{r atcGraph-setup}
library(ggplot2)
library(Rsolnp)     # Don't forget to load libraries!
A = 100 
a = 1 / 3
w = 10
r = 20
crtsSchedule <-
  data.frame(
    Quantity = seq(from = 0, to = 5000, by = 100),          # Initialize 51 Q's 
    Labor = NA,
    Capital = NA,
    Total_Cost = NA,
    Average_Cost = NA
  )
tc <- function(x) {
  w * x[1] + r * x[2]                                       # Define objective 
}
crts <- function(x) {
  A * x[1] ^ a * x[2] ^ (1 - a)                   # Define constraint
}
for (i in 1:nrow(crtsSchedule)) {                           # Initialize for-loop
  costMinTemp <- solnp(
    pars = c(1, 1),                                         # Set initial pars
    fun = tc,
    ineqfun = crts,
    ineqLB = crtsSchedule$Quantity[i],                      # Set ineqLB to Q[i]
    ineqUB = Inf,
    LB = c(0, 0),                                           # Nonnegative constraint
    UB = c(Inf, Inf), 
    control = list(trace = 0)
  )
  crtsSchedule$Labor[i] <- costMinTemp$pars[1]              # Optimal labor input
  crtsSchedule$Capital[i] <- costMinTemp$pars[2]            # Optimal capital input
  crtsSchedule$Total_Cost[i] <- tail(costMinTemp$values, 1) # Calculate total cost
}
crtsSchedule$Average_Cost <-
  crtsSchedule$Total_Cost / crtsSchedule$Quantity           # Calculate ATC
```

```{r atcGraph-solution}
acGraph <- ggplot() + 
  geom_line(aes(x = Quantity, y = Average_Cost), data = crtsSchedule[-1,]) +
  labs(x = "Quantity", y = "Average Cost") + lims(y = c(0,1))
```

```{r atcGraph-check}
grade_this({
  if (ith_geom_is(.result, "line", i = 1)) {
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

### Side-by-Side Plots

You can display the plots on the same plot area using `par(mfrow = c(nrow, ncol))`. I demonstrate this below: 

```{r costsSideBySideSetup}
library(ggplot2)
library(Rsolnp)     # Don't forget to load libraries!
A = 100
a = 1 / 3
w = 10
r = 20
crtsSchedule <-
  data.frame(
    Quantity = seq(from = 0, to = 5000, by = 100),          # Initialize 51 Q's 
    Labor = NA,
    Capital = NA,
    Total_Cost = NA,
    Average_Cost = NA
  )
tc <- function(x) {
  w * x[1] + r * x[2]                                       # Define objective 
}
crts <- function(x) {
  A * x[1] ^ a * x[2] ^ (1 - a)                   # Define constraint
}
for (i in 1:nrow(crtsSchedule)) {                           # Initialize for-loop
  costMinTemp <- solnp(
    pars = c(1, 1),                                         # Set initial pars
    fun = tc,
    ineqfun = crts,
    ineqLB = crtsSchedule$Quantity[i],                      # Set ineqLB to Q[i]
    ineqUB = Inf,
    LB = c(0, 0),                                           # Nonnegative constraint
    UB = c(Inf, Inf), 
    control = list(trace = 0)
  )
  crtsSchedule$Labor[i] <- costMinTemp$pars[1]              # Optimal labor input
  crtsSchedule$Capital[i] <- costMinTemp$pars[2]            # Optimal capital input
  crtsSchedule$Total_Cost[i] <- tail(costMinTemp$values, 1) # Calculate total cost
}
crtsSchedule$Average_Cost <-
  crtsSchedule$Total_Cost / crtsSchedule$Quantity           # Calculate ATC
tcGraph <- ggplot() + 
  geom_line(aes(x = Quantity, y = Total_Cost), data = crtsSchedule) +
  labs(x = "Quantity", y = "Total Cost")
acGraph <- ggplot() + 
  geom_line(aes(x = Quantity, y = Average_Cost), data = crtsSchedule[-1,]) +
  labs(x = "Quantity", y = "Total Cost") + lims(y = c(0,1))
```

```{r costsSideBySide, echo = TRUE, fig.width = 8}
library(ggpubr)
ggarrange(tcGraph, acGraph)
```

### Graphical Example with Diminishing Returns

A quick way to get an increasing cost function is to redefine the production function so that the sum of the Cobb-Douglas exponents is less than one. This creates *decreasing* (instead of constant) returns to scale, and therefore *increasing* average costs. For example, 
$$Q = f(L, K) = 100 L^{1/3} K^{1/3}$$

The following code illustrates increasing costs. 

```{r drtsGraph, echo = TRUE}
drtsSchedule <-
  data.frame(
    Quantity = seq(from = 0, to = 5000, by = 100),          # Initialize 51 Q's 
    Labor = NA,
    Capital = NA,
    Total_Cost = NA,
    Average_Cost = NA
  )
drts <- function(x) {
  A * x[1] ^ a * x[2] ^ a                                   # Define constraint
}
# Since alpha = 1/3, the exponents sum to less than one. 
for (i in 1:nrow(drtsSchedule)) {                           # Initialize for-loop
  costMinTemp <- solnp(
    pars = c(1, 1),                                         # Set initial pars
    fun = tc,
    ineqfun = drts,
    ineqLB = drtsSchedule$Quantity[i],                      # Set ineqLB to Q[i]
    ineqUB = Inf,
    LB = c(0, 0),                                           # Nonnegative constraint
    UB = c(Inf, Inf), 
    control = list(trace = 0)
  )
  drtsSchedule$Labor[i] <- costMinTemp$pars[1]              # Optimal labor input
  drtsSchedule$Capital[i] <- costMinTemp$pars[2]            # Optimal capital input
  drtsSchedule$Total_Cost[i] <- tail(costMinTemp$values, 1) # Calculate total cost
}
drtsSchedule$Average_Cost <-
  drtsSchedule$Total_Cost / drtsSchedule$Quantity           # Calculate ATC
tcGraphDRTS <- ggplot() + 
  geom_line(aes(x = Quantity, y = Total_Cost), data = drtsSchedule) +
  labs(x = "Quantity", y = "Total Cost")
acGraphDRTS <- ggplot() + 
  geom_line(aes(x = Quantity, y = Average_Cost), data = drtsSchedule [-1,]) +
  labs(x = "Quantity", y = "Average Cost") + lims(y = c(0,2))
ggarrange(tcGraphDRTS, acGraphDRTS)
```

### Graphical Example with Increasing Returns

Similarly, we can force a decreasing cost function by redefining the exponents in our Cobb-Douglas function so that their sum is greater than one. Such a technology exhibits *increasing* returns to scale and therefore *decreasing* average costs. For example, 
$$Q = f(L, K) = 100 L^{2/3} K^{2/3}$$

The following code simulates decreasing costs.  

```{r irtsGraph, echo = TRUE}
irtsSchedule <-
  data.frame(
    Quantity = seq(from = 0, to = 5000, by = 100),          # Initialize 51 Q's 
    Labor = NA,
    Capital = NA,
    Total_Cost = NA,
    Average_Cost = NA
  )
irts <- function(x) {
  A * x[1] ^ (2*a) * x[2] ^ (2*a)                           # Define constraint
}
# Since alpha = 1/3, the exponents sum to more than one. 
for (i in 1:nrow(irtsSchedule)) {                           # Initialize for-loop
  costMinTemp <- solnp(
    pars = c(1, 1),                                         # Set initial pars
    fun = tc,
    ineqfun = irts,
    ineqLB = irtsSchedule$Quantity[i],                      # Set ineqLB to Q[i]
    ineqUB = Inf,
    LB = c(0, 0),                                           # Nonnegative constraint
    UB = c(Inf, Inf), 
    control = list(trace = 0)
  )
  irtsSchedule$Labor[i] <- costMinTemp$pars[1]              # Optimal labor input
  irtsSchedule$Capital[i] <- costMinTemp$pars[2]            # Optimal capital input
  irtsSchedule$Total_Cost[i] <- tail(costMinTemp$values, 1) # Calculate total cost
}
irtsSchedule$Average_Cost <-
  irtsSchedule$Total_Cost / irtsSchedule$Quantity           # Calculate ATC
tcGraphIRTS <- ggplot() + 
  geom_line(aes(x = Quantity, y = Total_Cost), data = irtsSchedule) +
  labs(x = "Quantity", y = "Total Cost")
acGraphIRTS <- ggplot() + 
  geom_line(aes(x = Quantity, y = Average_Cost), data = irtsSchedule [-1,]) +
  labs(x = "Quantity", y = "Average Cost") + lims(y = c(0,0.5))
ggarrange(tcGraphIRTS, acGraphIRTS)
```


At the end of the day the nature of costs as the firm scales up its inputs is an empirical question. Businesses seeking to optimize input and output decisions hire professional economists to estimate both sales and costs. 

## Short-Run Cost Curves 

In our example, the firm initially anticipated sales of 2500 units. Based on this expectation it budgeted for 25 "units" of capital. Next, the firm just choose inputs (and outputs) based on its short-run costs. 

### Total Cost and Average Cost

With a fixed input (in the short run), a constant returns *to scale* technology becomes a diminishing returns *to the variable input(s)* technology. This introduces increasing costs, as demonstrated by the increasing slope of total cost. 

The following code simulates and graphs the short run total and average cost curves for our example at the long-run planned capital input of 25. 

```{r srTotalCost, echo = TRUE}
srCrtsSchedule <-
  data.frame(
    Quantity = seq(from = 0, to = 5000, by = 100),
    Labor = NA,
    Capital = 25,
    Total_Cost = NA,
    Average_Cost = NA
  )
K = 25
srCrts <- function(x) {
  A * x ^ (a) * 25 ^ (1 - a)
}
srtc <- function(x) {
  w * x + r * K
}
for (i in 1:nrow(srCrtsSchedule)) {
  costMinTemp <- solnp2(
    pars = 1,
    fun = srtc,
    ineqfun = srCrts,
    ineqLB = srCrtsSchedule$Quantity[i],
    ineqUB = Inf,
    LB = 0,
    UB = Inf,
    control = list(trace = 0)
  )
  srCrtsSchedule$Labor[i] <- costMinTemp$pars[1]
  srCrtsSchedule$Total_Cost[i] <- tail(costMinTemp$values, 1)
}
srCrtsSchedule$Average_Cost <-
  srCrtsSchedule$Total_Cost / srCrtsSchedule$Quantity
head(srCrtsSchedule)
```

```{r srTotalCostGraph, fig.align='center', fig.height=6}
srtcGraph <- ggplot() + 
  geom_line(aes(x = Quantity, y = Total_Cost, color = "TC"), data = srCrtsSchedule) +
  lims(x = c(0, 5000), y = c(0, 2500)) + labs(x = "Quantity", y = "Total Cost") + 
  labs(x = "Quantity", y = "Cost", color = NULL) 
sracGraph <- ggplot() + 
  geom_line(aes(x = Quantity, y = Average_Cost, color = "ATC"), data = srCrtsSchedule) +
  lims(x = c(0, 5000), y = c(0, 1)) + 
  labs(x = "Quantity", y = "Cost", color = NULL) 
ggarrange(srtcGraph, sracGraph)
```

### Variable Costs

Variable costs are costs associated with variable inputs (in our example labor). They are costs (and input decisions) that the firm can directly control in the short run. 

The following code simulates and graphs total and average *variable* costs for our example. 

```{r variableCost, echo = TRUE}
srCrtsSchedule$Variable_Cost <- w * srCrtsSchedule$Labor
srCrtsSchedule$Average_Variable_Cost <-
    srCrtsSchedule$Variable_Cost / srCrtsSchedule$Quantity
```

```{r variableCost-solution}
srtcGraph <- srtcGraph +
  geom_line(
    aes(x = Quantity, y = Variable_Cost, color = "TVC"), data = srCrtsSchedule) +
  scale_colour_manual(breaks = c("TC", "TVC"), values = c("black","purple"))
sracGraph <- sracGraph +
  geom_line(
    aes(x = Quantity, y = Average_Variable_Cost, color = "AVC"), 
    data = srCrtsSchedule[-1, ]) +
  scale_colour_manual(breaks = c("ATC", "AVC"), values = c("black","purple"))
ggarrange(srtcGraph, sracGraph)
```

### Marginal Costs

Marginal cost measures the rate at which total (or total vairable) cost changes as the firm increases its output by very small amounts. In our numerical simulation, we really only approximate the instantaneous value of the change because our code requires discrete changes (and because we're not deriving the full cost function symbolically). 

To partially compensate for this, we will average the change between the current cost and the *previous* cost with the change between the current cost and the *next* cost. 

By definition, the *derivative* of the cost function  takes a similar approach, but takes the limit as $\Delta Q \to 0$.

Run the following code to add marginal cost to the cost schedule and to the graphs. 

```{r marginalCost, echo=TRUE}
srCrtsSchedule$Marginal_Cost <- 
  c(NA, diff(srCrtsSchedule$Total_Cost, lag = 2)/200, NA)
```

```{r marginalCostGraph}
sracGraph <- sracGraph +
  geom_line(
    aes(x = Quantity, y = Marginal_Cost, color = "MC"), data = srCrtsSchedule) +
  scale_colour_manual(breaks = c("ATC", "AVC", "MC"), values = c("black","purple", "red"))
ggarrange(srtcGraph, sracGraph)
```
